---
title: "Explore stressor projections"
author: "Casey O'Hara"
---

```{r setup}
library(tidyverse)
library(terra)

```

## Summary

Read in stressor layers, examine distributions through plots and statistical summaries.

## Data

Data and results are presented in: Halpern, B. S., Frazier, M., O’Hara, C. C., Vargas-Fonseca, O. A., & Lombard, A. T. (2025). Cumulative impacts to global marine ecosystems projected to more than double by midcentury. Science, 0(0), eadv2906. https://doi.org/10.1126/science.adv2906

Data were retrieved from https://knb.ecoinformatics.org/view/doi%3A10.5063%2FF18K77KZ.

## Habitats

For each habitat layer, read in and plot.

```{r}
hab_fs <- list.files(here::here('data/habitat_maps'), pattern = '.tif', full.names = TRUE)
r <- rast(hab_fs)

plot(r)
```

## Stressors

Identify stressor layers by name, time period, and scenario.  Then examine key layers.

```{r}
str_df <- data.frame(f = list.files(here::here('data/pressures_raw'), 
                                    pattern = '.tif', full.names = TRUE)) %>%
  mutate(str = str_remove_all(basename(f), '_ssp.+'),
         scenario = str_extract(basename(f), 'ssp[0-9]+'),
         period   = str_extract(basename(f), 'current|medium-term'))

```

### Current

Plot current, ignoring air heat index for now.  Use spp585 scenario.

```{r}
str_current_df <- str_df %>%
  filter(period == 'current') %>%
  filter(str_detect(scenario, 'ssp5')) %>%
  filter(!str_detect(str, 'air-heat'))

r_current <- rast(str_current_df$f) %>% setNames(str_current_df$str)
plot(r_current)

```

```{r}
str_future_df <- str_df %>%
  filter(period == 'medium-term') %>%
  filter(str_detect(scenario, 'ssp5')) %>%
  filter(!str_detect(str, 'air-heat'))

r_future <- rast(str_future_df$f) %>% setNames(str_future_df$str)
plot(r_future)

```

### Compare differences

```{r}
plot(r_future - r_current)
```

