---
title: "Explore traits from framework and AquaX"
author: "Casey O'Hara"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
```

## Summary

Read in traits data from Butt framework and compare to/consolidate with AquaX.

## Data

Data and results are presented in: Butt, N., Halpern, B. S., O’Hara, C. C., Allcock, A. L., Polidoro, B., Sherman, S., Byrne, M., Birkeland, C., Dwyer, R. G., Frazier, M., Woodworth, B. K., Arango, C. P., Kingsford, M. J., Udyawer, V., Hutchings, P., Scanes, E., McClaren, E. J., Maxwell, S. M., Diaz-Pulido, G., … Klein, C. J. (2022). A trait-based framework for assessing the vulnerability of marine species to human impacts. Ecosphere, 13(2), e3919. https://doi.org/10.1002/ecs2.3919


## Traits from Vulnerability Framework

Traits data were installed manually into `data/spp_vuln_framework_traits` and added to .gitignore

```{r}
junction_f  <- here::here('data/spp_vuln_framework_traits/spp_traits_junction.csv')
taxa_f      <- here::here('data/spp_vuln_framework_traits/spp_traits_taxa_lookup.csv')
trait_val_f <- here::here('data/spp_vuln_framework_traits/spp_traits_trait_val_lookup.csv')
gapfill_f   <- here::here('data/spp_vuln_framework_traits/spp_traits_gapfill_levels.csv')

junction_df  <- data.table::fread(junction_f)
taxa_df      <- data.table::fread(taxa_f)
trait_val_df <- data.table::fread(trait_val_f)
gapfill_df   <- data.table::fread(gapfill_f)

spp_traits_df <-  junction_df %>%
  dplyr::left_join(taxa_df, by = 'gf_spp_id') %>%
  dplyr::left_join(trait_val_df, by = 'gf_trait_id') %>%
  dplyr::left_join(gapfill_df, by = 'gf_match_id') %>%
  ### drop id columns:
  dplyr::select(-ends_with('_id'))
```

## Traits from AquaX Metadata

AquaX data are in `/home/shares/data-aquax/META_02122025.Rdata` as object `META`

```{r}
load('/home/shares/data-aquax/META_02122025.Rdata')

aquax_meta <- META %>%
  janitor::clean_names() %>%
  filter(sdm != "No SDM") %>%
  select(aphia_id, sciname = scientific_name, 
         hab, depth_range_shallow, depth_range_deep, iucn_map,
         pelagic:bathypelagic) %>%
  distinct()
```

