---
title: Download stressors
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

Download stressor and habitat data from:

* Halpern, B. S., Frazier, M., O’Hara, C. C., Vargas-Fonseca, O. A., & Lombard, A. T. (2025). Cumulative impacts to global marine ecosystems projected to more than double by midcentury. Science, 0(0), eadv2906. https://doi.org/10.1126/science.adv2906

Data are hosted at https://knb.ecoinformatics.org/view/doi:10.5063/F18K77KZ:

* habitat_maps - get
* pressure_raw - get
* pressures_rescaled - get
* impacts_by_pressure - skip
* cumulative_impact - skip

## Methods

```{r}
library(tidyverse)
library(here)
library(terra)
```

For each data set, check for existence in the `data` folder, and if necessary, download/unzip.

For select layers,  reproject from 10 km Mollweide to 0.05 degree WGS 84.

Make sure to manually add these to .gitignore!

### Download habitat maps

```{r}
hab_maps_dir <- here::here('_data_raw/halpern_2025/habitat_maps')

if(dir.exists(hab_maps_dir)) {
  ### check if any tifs; if not, delete and re-download
  hab_maps_existing <- list.files(hab_maps_dir, pattern = '.tif$')
  if(length(hab_maps_existing) == 0) {
    unlink(hab_maps_dir, recursive = TRUE)
  }
}

if(!dir.exists(hab_maps_dir)) {
  hab_maps_url <- 'https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3A8598ed8c-e906-47fb-8ba5-4edfed35df0f'
  destfile <- here::here('hab_maps.zip')
  download.file(hab_maps_url, destfile, mode = 'wb')
  
  exdir <- here::here('_data_raw/halpern_2025')
  unzip(destfile, exdir = exdir)
  file.remove(destfile)
}
```

### Download raw stressor maps

These are not rescaled.

```{r}
prs_dir <- here::here('_data_raw/halpern_2025/pressures_raw')

if(dir.exists(prs_dir)) {
  ### check if any tifs; if not, delete and re-download
  prs_maps_existing <- list.files(prs_dir, pattern = '.tif$')
  if(length(prs_maps_existing) == 0) {
    unlink(prs_dir, recursive = TRUE)
  }
}

if(!dir.exists(prs_dir)) {
  prs_url <- 'https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Aebbdaaa0-bfc8-42c5-8973-76379dd65ee7'
  destfile <- here::here('prs_raw.zip')
  download.file(prs_url, destfile, mode = 'wb')
  
  exdir <- here::here('_data_raw/halpern_2025')
  unzip(destfile, exdir = exdir)
  file.remove(destfile)
}
```

### Download rescaled stressor maps

```{r}
prs_rescaled_dir <- here::here('_data_raw/halpern_2025/pressures_rescaled')
if(dir.exists(prs_rescaled_dir)) {
  ### check if any tifs; if not, delete and re-download
  prs_maps_existing <- list.files(prs_rescaled_dir, pattern = '.tif$')
  if(length(prs_maps_existing) == 0) {
    unlink(prs_rescaled_dir, recursive = TRUE)
  }
}

if(!dir.exists(prs_rescaled_dir)) {
  prs_url <- 'https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Acd2ce9a6-eaf4-4bb8-b3ca-09d253e036d9'
  destfile <- here::here('prs_raw.zip')
  download.file(prs_url, destfile, mode = 'wb')
  
  exdir <- here::here('_data_raw/halpern_2025')
  unzip(destfile, exdir = exdir)
  file.remove(destfile)
}
```

```{r}
cruft <- list.files(here::here('_data_raw/halpern_2025'), recursive = TRUE, include.dirs = TRUE,
                      full.names = TRUE, pattern = '^\\.DS|^__MAC|^~')

unlink(cruft, recursive = TRUE)
```

### Rescale and Reproject Stressors

Some stressors we will use the data directly from Halpern et al. 2025:

* light pollution
* ocean acidification
* coastal disturbance
* sea level rise
* nutrient pollution
* marine heat waves = ocean warming/SST
    * this was based on the extreme event weeks, analogous to marine heat waves

For these layers, let's use the rescaled layers directly for now.

```{r}
lyrs <- list.files(prs_rescaled_dir, full.names = TRUE, pattern = '.tif')
strs <- c('light.pollution', 'oa', 'coastal.disturbance', 'slr', 'nutrient.input', 'sst.extremes')

lyrs <- lyrs[grepl(paste(strs, collapse = '|'), basename(lyrs))]

reproject_layer <- function(f) {
  ### f <- lyrs[1]
  f_out <- sprintf(here('_output/pressures_rescaled/%s'), basename(f) %>% str_replace_all('-', '_'))
  if(!file.exists(f_out)) {
      r <- rast(f)
      new_r_base <- rast(ext = ext(-180, 180, -90, 90), res = 0.05, crs = 'epsg:4326')
      new_r <- project(r, new_r_base)
      writeRaster(new_r, f_out)
  }
  return(rast(f_out))
}

x <- parallel::mclapply(lyrs, reproject_layer, mc.cores = 6)
```

