---
title: Download species traits data
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

Download updated species traits data from:

* Butt et al. (2022). A trait-based framework for assessing the vulnerability of marine species to human impacts. Ecosphere, 13(2), e3919. https://doi.org/10.1002/ecs2.3919


Data are hosted at https://knb.ecoinformatics.org/view/doi:10.5063/F1QC0211 as:

* [spp_vuln_framework_traits.zip](https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Ab04c2ef1-7e75-4cd7-977f-56710fc57a9c)
* [spp_vuln_framework_scores.zip](https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3A614cd599-78eb-4e6b-905e-e68a2b2a0489)


## Methods

Check for existence in the `data` folder, and if necessary, download/unzip.

Make sure to manually add these to .gitignore!

### Download traits data

```{r}
traits_dir <- here::here('data/spp_vuln_framework_traits')
if(!dir.exists(traits_dir)) {
  traits_url <- 'https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Ab04c2ef1-7e75-4cd7-977f-56710fc57a9c'
  destfile <- here::here('traits.zip')
  download.file(traits_url, destfile, mode = 'wb')
  
  exdir <- here::here('data')
  unzip(destfile, exdir = exdir)
  file.remove(destfile)
}
```

### Download vulnerability scores

```{r}
vuln_dir <- here::here('data/spp_vuln_framework_scores')
if(!dir.exists(vuln_dir)) {
  vuln_url <- 'https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3A614cd599-78eb-4e6b-905e-e68a2b2a0489'
  destfile <- here::here('vuln.zip')
  download.file(vuln_url, destfile, mode = 'wb')
  
  exdir <- here::here('data')
  unzip(destfile, exdir = exdir)
  file.remove(destfile)
}
```

```{r}
cruft <- list.files(here::here('data'), recursive = TRUE, include.dirs = TRUE,
                      full.names = TRUE, pattern = '^\\.DS|^__MAC')

unlink(cruft, recursive = TRUE)
```

