---
title: Identify IUCN species and align to WoRMS
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

Access IUCN listed marine species and align with WoRMS taxonomy.


## Methods

### Gather IUCN assessed marine species

Access IUCN API to get a list of all assessed marine spp.

```{r load packages}
#| message: false
#| warning: false
library(tidyverse)

library(iucnredlist) ### remotes::install_github("IUCN-UK/iucnredlist")

### Initialize the API with API key from .Renviron
api <- init_api(Sys.getenv('IUCN_API_KEY'))

source(here::here('common_fxns.R'))
```

Identify comprehensively assessed groups, then pull assessment info for them.

```{r}
comp_gp_assess_f <- here::here('data/iucn_redlist/comp_group_assessments.csv')

if(!file.exists(comp_gp_assess_f)) {
  comp_gps <- list_codes(api, 'comprehensive_groups')
  
  comp_gps_marine <- comp_gps %>%
    filter(comp) %>%
    mutate(keep = case_when(
      higher_name == 'selected_bony_fishes' ~ TRUE,
      higher_name == 'birds'                ~ TRUE,
      higher_name == 'mammals'              ~ TRUE,
      higher_name == 'reef_forming_corals'  ~ TRUE,
      ### exclude chameleons from reptiles:
      name == 'chameleons'                  ~ FALSE,
      higher_name == 'selected_reptiles'    ~ TRUE,
      higher_name == 'selected_gastropods'  ~ TRUE,
      higher_name == 'sharks_and_rays'      ~ TRUE,
      name == 'hagfishes'                   ~ TRUE,
      name == 'mangrove_plants'             ~ TRUE,
      name == 'seagrasses'                  ~ TRUE,
      name == 'lobsters'                    ~ TRUE,
      TRUE                                  ~ FALSE)) %>%
    filter(keep) %>% select(-keep)
  
  gp_names <- comp_gps_marine$name %>% sort()
  
  comp_gp_assessments <- purrr::map(
    .x = gp_names, 
    .f = ~ assessments_by_group(
      api = api,
      group = 'comprehensive_groups',
      code = .x,
      scope = 1
    )
  )
  
  comp_gp_assess_df <- comp_gp_assessments %>%
    setNames(gp_names) %>%
    bind_rows(.id = 'group') %>%
    select(group, sis_taxon_id, assessment_id, year_published) %>%
    distinct() %>%
    mutate(year_published = as.integer(year_published))
  
  write_csv(comp_gp_assess_df, comp_gp_assess_f)
}

comp_gp_assess_df <- read_csv(comp_gp_assess_f, show_col_types = FALSE)
table(comp_gp_assess_df$group)
```

Using `sis_taxon_id` query scientific names.  While we're here, parse red list category and year assessed.  Chunk first checks to see if the final file is complete; if not, loops over each assessment, pinging API only if temp file is not present, then binds all temp files to create the final file.

```{r}
iucn_spp_assessment_f <- here::here('data/iucn_redlist/iucn_species_assessments.csv')

if(!file.exists(iucn_spp_assessment_f)) {
  assess_id_vec <- comp_gp_assess_df$assessment_id
  
  assess_tmp_fstem <- here::here('tmp/redlist_assess_%s.csv')
  
  assess_tmp_f_vec <- sprintf(assess_tmp_fstem, assess_id_vec)
  assess_to_do <- assess_id_vec[!file.exists(assess_tmp_f_vec)]
  
  message('Processing ', length(assess_to_do), ' remaining assessments...')
  
  ### Loop over assessment IDs
  for(i in seq_along(assess_to_do)) {
    ### i <- 1
    
    assess_id <- as.integer(assess_to_do[i])
    assess_tmp_f <- sprintf(assess_tmp_fstem, assess_id)
    
    if(file.exists(assess_tmp_f)) {
      # message(i, '. file exists: ', basename(assess_tmp_f), '... skipping!')
      next()
    }
    
    group <- comp_gp_assess_df$group[which(comp_gp_assess_df$assessment_id == assess_id)]
    # message(i, '. Finding assessment data for assessment ', assess_id, '...')
    
    Sys.sleep(1) ### pause to be kind to API
    x <- assessment_data(api, assess_id)
    
    ### extract data
    sis_taxon_id <- x$sis_taxon_id
    sciname      <- x$taxon$scientific_name
    rl_version   <- x$red_list_category$version
    rl_category  <- x$red_list_category$code
    year         <- x$year_published %>% as.integer()
    scope        <- x$scopes[[1]]$description$en
    
    df <- data.frame(group,
                     sis_taxon_id,
                     assessment_id = assess_id,
                     sciname,
                     redlist_version = rl_version,
                     redlist_category = rl_category,
                     assessed_year = year,
                     scope)
    
    write_csv(df, assess_tmp_f)
  }
  
  ### bind all temp files and write out final file
  assess_tmp_f_vec <- sprintf(assess_tmp_f, assess_id_vec)
  if(any(!file.exists(assess_tmp_f_vec))) stop('Missing one or more tmp files!')
  system.time({
  assess_all_list <- purrr::map(.x = assess_tmp_f_vec, 
                                .f = read_csv, 
                                show_col_types = FALSE)
  })
  system.time({
  assess_all_list <- parallel::mclapply(
    X = assess_tmp_f_vec, 
    FUN = read_csv, 
    show_col_types = FALSE,
    mc.cores = 40)
  })
  system.time({
  assess_all_df <- assess_all_list %>%
    setNames(str_extract(basename(assess_tmp_f_vec), '[0-9]+')) %>%
    bind_rows(.id = 'assessment_id')
  })
  
  write_csv(assess_all_df, iucn_spp_assessment_f)
}

assess_all_df <- read_csv(iucn_spp_assessment_f)

# x <- list.files(here::here('tmp'), pattern = 'redlist_assess_.*.csv', full.names = TRUE)
# unlink(x)
```

### Compare IUCN scinames to WoRMS names

First, check against the species already listed in the `spp_vuln_framework` or `funct_entity` datasets to identify Aphia IDs.  For any species not listed there, ping WoRMS API (via `taxize`) to resolve missed matches.

Finally, compare to AquaX metadata to see if any IUCN-mapped species are not included (remembering that these are only for comprehensively-assessed groups!)  
