---
title: Download MPA dataset
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

Download stressor and habitat data from:

* UNEP-WCMC and IUCN (2026), Protected Planet: The World Database on Protected Areas (WDPA) and World Database on Other Effective Area-based Conservation Measures (WD-OECM) [Online], July 2026, Cambridge, UK: UNEP-WCMC and IUCN. Available at: www.protectedplanet.net.
* https://www.protectedplanet.net/en/search-areas?filters%5Bis_type%5D%5B%5D=marine
* download link for WDPA and WDOECM (July 2026): https://d1gam3xoknrgr2.cloudfront.net/current/WDPA_WDOECM_Jul2026_Public_1b091e2b51c258eb3b24328044a4a3d6979e00fa2029d7538aef2befd46da362_shp.zip



## Methods

```{r}
library(tidyverse)
library(here)
library(sf)
library(terra)
```

Download WDPA database, filter to MPAs and marine OECMs.
Rasterize to AquaX resolution.

Make sure to manually add these to .gitignore!

### Download protected area maps

```{r}
mpa_data_dir <- here('_data_raw/unep_wcmc_2026/')

if(!dir.exists(mpa_data_dir)) {
  mpa_maps_url <- 'https://d1gam3xoknrgr2.cloudfront.net/current/WDPA_WDOECM_Jul2026_Public_1b091e2b51c258eb3b24328044a4a3d6979e00fa2029d7538aef2befd46da362_shp.zip'
  destfile <- here::here('wdpa.zip')
  download.file(mpa_maps_url, destfile, mode = 'wb')
  
  unzip(destfile, exdir = mpa_data_dir)
  file.remove(destfile)
}
```

### Load in WDPA maps

Load and inspect WDPA maps for completeness.  Unzipped folder contains new zips of several different shapefiles that combined represent the entirety of the polygons (and points?)

```{r}
gpkg_out <- file.path(mpa_data_dir, 'wdpa_wdoecm_july2026.gpkg')

if(!file.exists(gpkg_out)) {
  zips <- list.files(mpa_data_dir, pattern = '.zip', full.names = TRUE)
  shp_dir <- file.path(mpa_data_dir, 'WDPA_shp')
  for(z in zips) {
    ### z <- zips[1]
    unzip(z, exdir = file.path(shp_dir, gsub('.zip', '', basename(z))))
    unlink(zips)
  }

  shp_fs <- list.files(shp_dir, pattern = '.shp$', 
                      recursive = TRUE, full.names = TRUE)

  wdpa_sf <- parallel::mclapply(shp_fs, 
                                mc.cores = length(shp_fs), 
                                FUN = read_sf) %>%
    bind_rows() %>%
    janitor::clean_names()
  write_sf(wdpa_sf, gpkg_out)
  unlink(shp_dir, recursive = TRUE)
}
```

### Download EEZ data

Download EEZ polygons from Marine Regions:

* Flanders Marine Institute (2023). Maritime Boundaries Geodatabase: Maritime Boundaries and Exclusive Economic Zones (200NM), version 12. Available online at https://www.marineregions.org/. https://doi.org/10.14284/632

```{r}
eez_sf <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```