---
title: Download EEZ and georegion data
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

Create zonal rasters for EEZs based on region (continent-scale), sub-region (subcontinental), intermediate (further regionalized) for more detailed MPA analysis.

* Marine Regions attribute table: https://www.marineregions.org/eezattribute.php
* UN M49 codes: https://unstats.un.org/unsd/methodology/m49/overview/# (semicolon-separated values)

## Methods

```{r}
library(tidyverse)
library(here)
library(sf)
library(terra)
```

EEZ data saved on Aurora.  UN M49 georegions saved in repo.  Read in shapefile, bind georegions by M49 code, and rasterize out to 

```{r}
base_r <- rast(ext(c(-180, 180, -90, 90)), res = 0.05, crs = 'epsg:4326')

eez_sf <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg') |>
  janitor::clean_names() |>
  ### coerce numeric UN codes to zero-padded 3 digit character
  mutate(across(starts_with('un_'), .fns = \(x) ifelse(is.na(x), NA, sprintf('%03d', as.integer(x))))) |>
  ### Use territorial code if available; otherwise use sov code
  mutate(m49_code = ifelse(is.na(un_ter1), un_sov1, un_ter1))

### georegions in semicolon-sep values, so use read_csv2
unm49_df <- read_csv2(here('_data_raw/eezs/UNSD — Methodology.csv')) |>
  janitor::clean_names()

eez_georgns <- inner_join(eez_sf, unm49_df, by = c('m49_code')) %>%
  mutate(intermediate_region_code = ifelse(is.na(intermediate_region_code), 
                                           sub_region_code, 
                                           intermediate_region_code)) %>%
  mutate(across(contains('region_code'), as.integer))

dropped_ter <- anti_join(eez_sf, unm49_df, by = c('m49_code'))

eez_r  <- rasterize(eez_georgns |> mutate(m49_code = as.integer(m49_code)), 
                    base_r, field = 'm49_code')
rgn1_r <- rasterize(eez_georgns |> filter(!is.na(region_code)), 
                    base_r, field = 'region_code')
rgn2_r <- rasterize(eez_georgns |> filter(!is.na(sub_region_code)), 
                    base_r, field = 'sub_region_code')
rgn3_r <- rasterize(eez_georgns |> filter(!is.na(sub_region_code)), 
                    base_r, field = 'intermediate_region_code')

rgn_stack <- c(eez_r,rgn1_r, rgn2_r, rgn3_r) %>%
  setNames(c('un_m49_code', 'region_code', 'sub_region_code', 'intermediate_region_code'))

plot(rgn_stack)

writeRaster(rgn_stack, here('_data_raw/eezs/region_codes.tif'), overwrite = TRUE)
```

