---
title: Download EEZ and georegion data
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

Create zonal rasters for EEZs based on region (continent-scale), sub-region (subcontinental), intermediate (further regionalized) for more detailed MPA analysis.

* Marine Regions attribute table: https://www.marineregions.org/eezattribute.php
* UN M49 codes: https://unstats.un.org/unsd/methodology/m49/overview/# (semicolon-separated values)

## Methods

```{r}
library(tidyverse)
library(here)
library(sf)
library(terra)
```

EEZ data saved on Aurora.  UN M49 georegions saved in repo.  Read in shapefile, rasterize out using un_ter1 and un_sov1 codes (M49 codes).  These can be joined to the georegion dataset.

```{r}
base_r <- rast(ext(c(-180, 180, -90, 90)), res = 0.05, crs = 'epsg:4326')

eez_sf <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg') |>
  janitor::clean_names() |>
  ### Use territorial code if available; otherwise use sov code
  mutate(m49_code = ifelse(is.na(un_ter1), un_sov1, un_ter1),
         m49_code = as.integer(m49_code))

m49_r  <- rasterize(eez_sf, base_r, field = 'm49_code')

writeRaster(m49_r, here('_data/eezs/un_m49_codes.tif'), overwrite = TRUE)
```

