---
title: Download fisheries data from FAO and ISIMIP
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

Download observed fisheries statistics from FAO and projected marine biomass outputs from ISIMIP3b.

### FAO fisheries statistics (saved to `_data/fao_2026/`)
* Capture fisheries: https://www.fao.org/fishery/collection/capture/en
  * Direct download: https://www.fao.org/fishery/static/Data/Capture_2026.1.0.zip
* Aquaculture production: https://www.fao.org/fishery/collection/aquaculture/en
  * Direct download: https://www.fao.org/fishery/static/Data/Aquaculture_2026.1.0.zip

### ISIMIP3b EcoOcean projections (saved to `_data/isimip_2026/`)
* Model: EcoOcean, forced with GFDL-ESM4 under SSP5-8.5
* Data portal: https://data.isimip.org/search/tree/ISIMIP3b
* File list: `_data/isimip_2026/filelist.txt`


## Methods


Make sure to manually add these to .gitignore!

### Download FAO .zips

```{r}
capture_q_f <- here::here('_data/fao_2026/capture/Capture_Quantity.csv')
if(!dir.exists(dirname(capture_q_f))) dir.create(dirname(capture_q_f))

if(!file.exists(capture_q_f)) {
  capture_url <- 'https://www.fao.org/fishery/static/Data/Capture_2026.1.0.zip'
  destfile <- here::here('capture.zip')
  download.file(capture_url, destfile, mode = 'wb')
  
  exdir <- here::here('_data/fao_2026/capture')
  unzip(destfile, exdir = exdir)
  file.remove(destfile)
}
```

```{r}
aquaculture_q_f <- here::here('_data/fao_2026/aquaculture/Aquaculture_Quantity.csv')
if(!dir.exists(dirname(aquaculture_q_f))) dir.create(dirname(aquaculture_q_f))

if(!file.exists(aquaculture_q_f)) {
  aquaculture_url <- 'https://www.fao.org/fishery/static/Data/Aquaculture_2026.1.0.zip'
  destfile <- here::here('aquaculture.zip')
  download.file(aquaculture_url, destfile, mode = 'wb')
  
  exdir <- here::here('_data/fao_2026/aquaculture')
  unzip(destfile, exdir = exdir)
  file.remove(destfile)
}
```

### Download ISIMIP netCDFs


```{r}
filelist <- readr::read_lines(here::here('_data/isimip_2026/filelist.txt'))

data_dir <- here::here('_data/isimip_2026/rasts')
if(!dir.exists(data_dir)) dir.create(data_dir)

file_df <- data.frame(source = filelist, 
                      dest = file.path(data_dir, basename(filelist)))

for(i in 1:nrow(file_df)) {
  ### i <- 1
  src <- file_df$source[i]
  dst <- file_df$dest[i]
  if(file.exists(dst)) next()
  
  ### else:
  download.file(src, dst, mode = 'wb')
}
```