---
title: Download data for functional entity traits
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

Download traits used to calculate functional entities for:

* O’Hara et al. (2024). Cumulative human impacts on global marine fauna highlight risk to biological and functional diversity. PLOS ONE, 19(9), e0309788. https://doi.org/10.1371/journal.pone.0309788

Data are hosted at https://github.com/mapping-marine-spp-vuln/spp_vuln_mapping/tree/main/_data/traits_grouping:

* [trait_length_cm.csv](https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_length_cm.csv)
* [trait_mobility.csv](https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_mobility.csv)
* [trait_reproduction.csv](https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_reproduction.csv)
* [trait_trophic_level.csv](https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_trophic_level.csv)
* [trait_water_col_position.csv](https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_water_col_position.csv)

The lookup table for correcting species names per WoRMS Aphia ID is located in https://github.com/mapping-marine-spp-vuln/spp_vuln_framework:

* [update_sciname_aphia.csv](https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_framework/refs/heads/master/update/update_sciname_aphia.csv)

## Methods

```{r}
library(tidyverse)
```

Read in all files, combine, and save out.

Make sure to manually add these to .gitignore!

### Load traits data

```{r}
fe_len  <- read_csv('https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_length_cm.csv')
fe_mob  <- read_csv('https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_mobility.csv')
fe_repr  <- read_csv('https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_reproduction.csv')
fe_troph <- read_csv('https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_trophic_level.csv')
fe_wcol  <- read_csv('https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_mapping/refs/heads/main/_data/traits_grouping/trait_water_col_position.csv')

```

### Identify species names with updated WoRMS accepted names/Aphia ID

```{r}
all_spp_vec <- c(fe_len$species, fe_mob$species, fe_repr$species, fe_troph$species, fe_wcol$species) %>%
  unique() %>% sort()

spp_vuln_aphia <- read_csv('https://raw.githubusercontent.com/mapping-marine-spp-vuln/spp_vuln_framework/refs/heads/master/update/update_sciname_aphia.csv') %>%
  filter(!is.na(aphia_id))

spp_trait_match <- data.frame(sciname_fe = all_spp_vec) %>%
  inner_join(spp_vuln_aphia, by = c('sciname_fe' = 'sciname'))
```

### Deal with duplicate entries (i.e., multiple values from FishBase)

```{r}
resolve_numeric_dupes <- function(df) {
  df_dupes <- df %>%
    janitor::get_dupes(species, trait) %>%
    summarize(.by = c(species, trait),
              sd = sd(value), value = mean(value))
  df_out <- df %>%
    filter(!species %in% df_dupes$species) %>%
    bind_rows(df_dupes)
}

fe_len1 <- resolve_numeric_dupes(fe_len) %>% rename(length_cm = value, sd_len = sd)
fe_trp1 <- resolve_numeric_dupes(fe_troph) %>% rename(troph = value, sd_trp = sd)

mob_dupes <- janitor::get_dupes(fe_mob, species)
### duplicate mobility bins ok for now; collect into one obs

resolve_chr_dupes <- function(df) {
  df_dupes <- df %>%
    janitor::get_dupes(species, trait) %>%
    summarize(.by = c(species, trait),
              value = paste0(sort(unique(value)), collapse = ';'))
  df_out <- df %>%
    filter(!species %in% df_dupes$species) %>%
    bind_rows(df_dupes)
}
fe_mob1 <- resolve_chr_dupes(fe_mob) %>% rename(adult_mob = value, sd_mob = sd)

fe_wcol1 <- resolve_chr_dupes(fe_wcol) %>%
  rename(wcol_pos = value) %>% select(-source)
### no duplicates found

### drop numeric value column and allow for duplicate age_mat and fec bins
fe_age_mat1 <- fe_repr %>%
  filter(trait == 'age_mat') %>%
  select(-source, -val_num) %>%
  resolve_chr_dupes() %>%
  rename(age_mat = value)

fe_fec1 <- fe_repr %>%
  filter(trait == 'fec') %>%
  select(-source, -val_num) %>%
  resolve_chr_dupes() %>%
  rename(fec = value)

```

### update species name and add aphia_id

```{r}
fe_df <- fe_len1 %>%
  full_join(fe_trp1, by = 'species') %>%
  full_join(fe_mob1, by = 'species') %>%
  full_join(fe_wcol1, by = 'species') %>%
  full_join(fe_age_mat1, by = 'species') %>%
  full_join(fe_fec1, by = 'species') %>%
  select(-starts_with(c('trait', 'nspp', 'gf_level'))) %>%
  inner_join(spp_trait_match, by = c('species' = 'sciname_fe')) %>%
  select(-species, sciname = valid_sciname)

write_csv(fe_df, here::here('data/funct_entity_traits/fe_spp_traits.csv'))
```

