---
title: Download EEZ and georegion data
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

Create zonal rasters for EEZs and UN M49 georegions (continent, sub-region, intermediate) for use in MPA and regional analyses. Also includes the ESRI World Continents polygon layer for continent-level aggregation.

Data sources:

* EEZ polygons — Marine Regions, World EEZ v12 (2023-10-25), stored on shared server at `/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg`
  * Attribute table: https://www.marineregions.org/eezattribute.php
  * Citation: Flanders Marine Institute (2023). Maritime Boundaries Geodatabase: Maritime Boundaries and Exclusive Economic Zones (200NM), version 12. Available online at https://www.marineregions.org/. https://doi.org/10.14284/632
* UN M49 codes — https://unstats.un.org/unsd/methodology/m49/overview/ (semicolon-separated values)
* ESRI World Continents — https://hub.arcgis.com/datasets/esri::world-continents/explore

## Methods

```{r}
library(tidyverse)
library(here)
library(sf)
library(terra)
```

EEZ data saved on Aurora.  UN M49 georegions saved in repo.  Read in shapefile, rasterize out using un_ter1 and un_sov1 codes (M49 codes).  These can be joined to the georegion dataset.

```{r}
base_r <- rast(ext(c(-180, 180, -90, 90)), res = 0.05, crs = 'epsg:4326')

eez_sf <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg') |>
  janitor::clean_names() |>
  ### Use territorial code if available; otherwise use sov code
  mutate(m49_code = ifelse(is.na(un_ter1), un_sov1, un_ter1),
         m49_code = as.integer(m49_code))

m49_r  <- rasterize(eez_sf, base_r, field = 'm49_code')

writeRaster(m49_r, here('_data/eezs/un_m49_codes.tif'), overwrite = TRUE)
```

