---
title: Download stressors
author: "Casey O'Hara"
---

## Summary

Download stressor and habitat data from:

* Halpern, B. S., Frazier, M., O’Hara, C. C., Vargas-Fonseca, O. A., & Lombard, A. T. (2025). Cumulative impacts to global marine ecosystems projected to more than double by midcentury. Science, 0(0), eadv2906. https://doi.org/10.1126/science.adv2906

Data are hosted at https://knb.ecoinformatics.org/view/doi:10.5063/F18K77KZ:

* habitat_maps - get
* pressure_raw - get
* pressures_rescaled - skip
* impacts_by_pressure - skip
* cumulative_impact - skip

## Methods

For each data set, check for existence in the `data` folder, and if necessary, download/unzip.

Make sure to manually add these to .gitignore!

### Download habitat maps

```{r}
hab_maps_dir <- here::here('data/habitat_maps')
if(!dir.exists(hab_maps_dir)) {
  hab_maps_url <- 'https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3A8598ed8c-e906-47fb-8ba5-4edfed35df0f'
  destfile <- here::here('hab_maps.zip')
  download.file(hab_maps_url, destfile, mode = 'wb')
  
  exdir <- here::here('data')
  unzip(destfile, exdir = exdir)
  file.remove(destfile)
  
  cruft <- list.files(exdir, recursive = TRUE, include.dirs = TRUE,
                      full.names = TRUE, pattern = '.DS*|__MAC')
  unlink(cruft, recursive = TRUE)
}
```

### Download raw stressor maps

These are not rescaled.

```{r}
prs_dir <- here::here('data/pressures_raw')
if(!dir.exists(hab_maps_dir)) {
  prs_url <- 'https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Aebbdaaa0-bfc8-42c5-8973-76379dd65ee7'
  destfile <- here::here('prs_raw.zip')
  download.file(prs_url, destfile, mode = 'wb')
  
  exdir <- here::here('data')
  unzip(destfile, exdir = exdir)
  file.remove(destfile)
  
  cruft <- list.files(exdir, recursive = TRUE, include.dirs = TRUE,
                      full.names = TRUE, pattern = '.DS*|__MAC')
  unlink(cruft, recursive = TRUE)
}
```

