---
title: Map impact stats by stressor across all species
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

For each stressor, combine all species-wide stressor vulnerability maps with stressor distribution maps.

## Data

* vulnerability maps generated in 2_map_vulnerability_stressor_all_spp.qmd
* stressor maps generated in XXXX

## Methods

* Read in SDMs for multiple species as long data frame
* Group by pixel and summarize stats.
* Save out as raster or parquet.

```{r attach packages}
#| message: false
#| 
library(tidyverse)
library(arrow)
library(collapse)
library(here)
library(terra)

source(here('common_fxns.R'))
```

### Gather species richness maps

```{r gather spp_richness}
nspp_df <- read_parquet(here('_output/vuln_maps/spp_richness_all_spp.parquet'))
nspp_r  <- rast(nspp_df)

plot(nspp_r)

### pull scenarios from data
scenarios <- names(nspp_df)[!names(nspp_df) %in% c('x', 'y', 'tx', 'f')]
```

### Iterate over stressors and scenarios

For each stressor, iterate over scenarios to generate a single mapset of pooled mean, sd, skew, and kurtosis.

```{r find stressor vuln maps}
vuln_mapfs <- list.files(here('_output/vuln_maps'), pattern = '^all_spp_vuln_.+parquet$', full.names = TRUE)
str_vulns  <- basename(vuln_mapfs) %>% str_remove_all('all_spp_vuln_|.parquet')

str_mapfs  <- list.files(here('_data_raw/halpern_2025/pressures_rescaled'), pattern = '.tif', full.names = TRUE)
str_dists  <- basename(str_mapfs) %>% str_remove_all('_ssp[0-9].+') %>% unique()
```

### Marine Heat Waves
```{r}
mhw_vuln_df <- read_parquet(vuln_mapfs[str_detect(vuln_mapfs, 'marine_heat_waves')]) %>%
  fsubset(stat == 'mean') %>% select(-stat)
mhw_vuln <- rast(mhw_vuln_df, crs = 'epsg:4326')
plot(mhw_vuln)
```

```{r}
mhw_maps <- str_mapfs[str_detect(str_mapfs, 'sst-extremes') & str_detect(str_mapfs, 'ssp245_current|medium-term')]
mhw_dist <- rast(mhw_maps) %>% setNames(scenarios) %>% project(mhw_vuln)
plot(mhw_dist)
```

```{r}
mhw_impact <- mhw_vuln * mhw_dist

plot(mhw_impact)
```



### Nutrient Pollution
```{r}
nutr_vuln_df <- read_parquet(vuln_mapfs[str_detect(vuln_mapfs, 'nutrient_pollution')]) %>%
  fsubset(stat == 'mean') %>% select(-stat)
nutr_vuln <- rast(nutr_vuln_df, crs = 'epsg:4326')
plot(nutr_vuln)
```

```{r}
nutr_maps <- str_mapfs[str_detect(str_mapfs, 'nutrient-input') & str_detect(str_mapfs, 'ssp2_current|medium-term')]
nutr_dist <- rast(nutr_maps) %>% setNames(scenarios) %>% project(nutr_vuln)
plot(nutr_dist)
```

```{r}
nutr_impact <- nutr_vuln * nutr_dist

plot(nutr_impact)
```

### Ocean Acidification

```{r}
oa_vuln_df <- read_parquet(vuln_mapfs[str_detect(vuln_mapfs, '_oa\\.')]) %>%
  fsubset(stat == 'mean') %>% select(-stat)
oa_vuln <- rast(oa_vuln_df, crs = 'epsg:4326')
plot(oa_vuln)
```

```{r}
oa_maps <- str_mapfs[str_detect(basename(str_mapfs), '^oa_') & str_detect(str_mapfs, 'ssp245_current|medium-term')]
oa_dist <- rast(oa_maps) %>% setNames(scenarios) %>% project(oa_vuln)
plot(oa_dist)
```

```{r}
oa_impact <- oa_vuln * oa_dist

plot(oa_impact) 
```

