---
title: Map impact stats by stressor across all species
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

For each stressor, combine all species-wide stressor vulnerability maps with stressor distribution maps.

## Data

* vulnerability maps generated in 2_map_vulnerability_stressor_all_spp.qmd
* stressor maps generated in XXXX

## Methods

* Read in SDMs for multiple species as long data frame
* Group by pixel and summarize stats.
* Save out as raster or parquet.

```{r attach packages}
#| message: false
#| 
library(tidyverse)
library(arrow)
library(collapse)
library(here)
library(terra)

source(here('common_fxns.R'))
```

### Gather species richness maps

```{r gather spp_richness}
nspp_df <- read_parquet(here('_output/vuln_maps/spp_richness_all_spp.parquet'))
nspp_r  <- rast(nspp_df)

plot(nspp_r)

### pull scenarios from data
scenarios <- names(nspp_df)[!names(nspp_df) %in% c('x', 'y', 'tx', 'f')]
```

### Iterate over stressors and scenarios

For each stressor, iterate over scenarios to generate a single mapset of pooled mean, sd, skew, and kurtosis.

```{r find stressor vuln maps}
vuln_mapfs <- list.files(here('_output/vuln_maps'), pattern = '^all_spp_vuln_.+parquet$', full.names = TRUE)
str_vulns  <- basename(vuln_mapfs) %>% str_remove_all('all_spp_vuln_|.parquet')

str_mapfs  <- list.files(here('_data_raw/halpern_2025/pressures_rescaled'), pattern = '.tif', full.names = TRUE)
str_dists  <- basename(str_mapfs) %>% str_remove_all('_ssp[0-9].+') %>% unique()
```


```{r}
calc_impacts <- function(vuln_f, str_f, 
                         scenarios = c('current', 'rcp45_2050', 'rcp85_2050'), 
                         out_f, overwrite = FALSE) {
  
  if(file.exists(out_f) & !overwrite) {
    return(read_parquet(out_f))
  }
  
  vuln_df <- read_parquet(vuln_f)

  vuln_mean_df <- vuln_df %>% fsubset(stat == 'mean') %>% select(-stat)
  vuln_sd_df   <- vuln_df %>% fsubset(stat == 'sd')   %>% select(-stat)

  vuln_mean_r <- rast(vuln_mean_df, crs = 'epsg:4326')
  vuln_sd_r   <- rast(vuln_sd_df,   crs = 'epsg:4326')

  str_dist <- rast(str_f) %>% setNames(scenarios) %>% project(vuln_mean_r)

  impact_mean_r <- vuln_mean_r * str_dist
  impact_sd_r   <- vuln_sd_r   * str_dist

  impact_df <- as.data.frame(impact_mean_r, xy = TRUE) %>%
    fmutate(stat = 'mean') %>%
    rowbind(as.data.frame(impact_sd_r, xy = TRUE) %>%
              fmutate(stat = 'sd')) %>%
    fmutate(across(where(is.numeric), \(x) round(x, 3)))

  write_parquet(impact_df, out_f)
  return(impact_df)
}

plot_from_df <- function(df, plotstat = 'mean') {
  if('stat' %in% names(df)) df <- fsubset(df, stat == plotstat) %>% select(x, y, everything(), -stat)
  r <- rast(df)
  plot(r)
}
```

### Marine Heat Waves
```{r}
mhw_vuln_f <- vuln_mapfs[str_detect(vuln_mapfs, 'marine_heat_waves')]
mhw_maps <- str_mapfs[str_detect(str_mapfs, 'sst-extremes') & str_detect(str_mapfs, 'ssp245_current|medium-term')]

mhw_impact_df <- calc_impacts(vuln_f = mhw_vuln_f,
                              str_f  = mhw_maps,
                              out_f  = here('_output/impact_maps/impact_marine_heat_waves.parquet'))

plot_from_df(mhw_impact_df)
plot_from_df(mhw_impact_df, plotstat = 'sd')
```

### Nutrient Pollution
```{r}
nutr_vuln_f <- vuln_mapfs[str_detect(vuln_mapfs, 'nutrient_pollution')]
nutr_maps <- str_mapfs[str_detect(str_mapfs, 'nutrient-input') & str_detect(str_mapfs, 'ssp2_current|medium-term')]

nutr_impact_df <- calc_impacts(vuln_f = nutr_vuln_f,
                              str_f   = nutr_maps,
                              out_f   = here('_output/impact_maps/impact_nutrient_pollution.parquet'))
plot_from_df(nutr_impact_df)
plot_from_df(nutr_impact_df, plotstat = 'sd')
```

### Ocean Acidification

```{r}
oa_vuln_f <- vuln_mapfs[str_detect(vuln_mapfs, '_oa\\.')]

oa_maps <- str_mapfs[str_detect(basename(str_mapfs), '^oa_') & str_detect(str_mapfs, 'ssp245_current|medium-term')]

oa_impact_df <- calc_impacts(vuln_f = oa_vuln_f,
                             str_f  = oa_maps,
                             out_f  = here('_output/impact_maps/impact_ocean_acidification.parquet'))
plot_from_df(oa_impact_df)
plot_from_df(oa_impact_df, plotstat = 'sd')
```

