---
title: Map SST impact stats across all species
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

For SST rise, assign temperature bands to species, and generate temperature stressor maps for each band.  Then, somehow apply those maps across all species in a sensible way...

## Data

* vulnerability maps generated in 2_map_vulnerability_stressor_all_spp.qmd
* SST data from BioOracle.

## Methods

* From AquaX metadata, generate bins of temperature preferences (upper range) for species.
* For each bin, identify species, read in SDM, apply vulnerability score
* For each bin, generate a stressor value map from SST data
* For each bin/spp group, generate stats summaries of impact across all spp present.

How to account for differences:

* Benthic vs pelagic spp?  Depth zones?
* Max/min T vs mean T?
* Summarized by taxon?

```{r attach packages}
#| message: false
#| warning: false
#| 
library(tidyverse)
library(arrow)
library(collapse)
library(here)
library(terra)

# remotes::install_github('bio-oracle/biooracler')
library(biooracler)

source(here('common_fxns.R'))
```

### Identify spp depth

From AquaX metadata, identify whether pelagic or benthic temperature is more relevant.  In the metadata, "benthopelagic" is classified as "demersal" which makes sense.  Assume demersal are generally more on bottom than surface.

```{r}
aquax_meta <- get_aquax_meta()
aquax_wcol_raw <- aquax_meta %>%
  select(aphia_id, pelagic, benthic, demersal, hab)

benth_pel_check <- aquax_wcol_raw %>%
  mutate(wcol = case_when(pelagic & !demersal & !benthic ~ 'pelagic',
                          !pelagic & (demersal | benthic) ~ 'benthic',
                          pelagic & (demersal | benthic) ~ 'benthic/pelagic', ### no cases
                          TRUE ~ 'huh'))

# benth_pel_check %>% select(hab, wcol_flag) %>% table()
# #                  wcol_flag
# # hab               benth   pel
# #   bathydemersal    1419     0
# #   bathypelagic        0   986
# #   benthic         13575     0
# #   benthopelagic    1083     0
# #   demersal         5176     0
# #   pelagic             0  1156
# #   pelagic-neritic     0   625
# #   pelagic-oceanic     0   393
# #   reef-associated  5168     0
# #   sessile          1928     0

spp_wcol <- benth_pel_check %>%
  select(aphia_id, wcol) %>%
  distinct()
```

### Generate temperature bins

Thermal niches from AquaX Dropbox, processed and rounded to identify species with similar niches.  Thermal niche defined in same way as O'Hara et al. 2024 (taken from AquaMaps):

$$s_T = \begin{cases}
  0 &\text{where }\bar T \leq T_{max}^{p}\\
  \frac{\bar T - T_{max}^{p}}{T_{max}^a - T_{max}^p} &\text{where }T_{max}^{p} < \bar T < T_{max}^{a}\\
  1 &\text{where }\bar T \geq T_{max}^{p}
\end{cases}$$
               
Where:

* $T^a_{max}$ = 75th percentile + 1.5 × interquartile or absolute maximum mean annual temperature (whichever is greater)  
* $T^p_{max}$ = 90th percentile of observed variation in mean annual temperature

But: the AquaX data only has 75%ile and 95%ile, not 90%ile.  Let's use 75%ile instead (for now).
               
```{r}
thermal_niches_raw <- read_parquet(here('_data/thermal_niche/spp_thermal_niches.parquet')) %>%
  fsubset(variable == 'thetao_mean') %>%
  inner_join(spp_wcol, by = c('aphia_id', 'wcol')) %>%
  mutate(t_a_max = q75 + (q75-q25)*1.5,
         t_a_max = ifelse(t_a_max > max, max, t_a_max),
         t_p_max = q75) %>%
  select(t_a_max, t_p_max, aphia_id, wcol, src)

src_range_check <- thermal_niches_raw %>%
  group_by(aphia_id) %>%
  arrange(src) %>%
  summarize(wcol = first(wcol),
            t_a_range = first(t_a_max) - last(t_a_max),
            t_p_range = first(t_p_max) - last(t_p_max),
            n_src = n_distinct(src), .groups = 'drop')

hist(src_range_check$t_a_range); hist(src_range_check$t_p_range)
weird <- src_range_check %>% filter(abs(t_a_range) > 20 | abs(t_p_range) > 20) %>%
  left_join(aquax_meta)
```

### Round thermal envelope and form bins

```{r}
thermal_niches <- thermal_niches_raw %>%
  filter(src == 'occ') %>% 
  select(-src) %>%
  mutate(t_a_max = round(t_a_max * 2) / 2,
         t_p_max = round(t_p_max * 2) / 2) %>%
  mutate(t_a_max = round(t_a_max),
         t_p_max = round(t_p_max)) %>%
  mutate(range = t_a_max - t_p_max)

bins <- thermal_niches %>% 
  group_by(t_a_max, t_p_max) %>%
  summarize(nspp = n_distinct(aphia_id))
### half-degree bins: 1213 different envelopes
### one-degree bins: 338 different envelopes
```


***QUESTION***: why are we missing some species? 31246 have thermal preferences; but 31509 spp in AquaX.  Re-`rclone`?

```{r}
missing <- aquax_meta %>%
  filter(!aphia_id %in% thermal_niches$aphia_id)
```

### Load Thermal Layers

Need:

* Current SST and seafloor T
* Future SST and seafloor T - decade 2050 at SSP2-4.5 and SSP5-8.5.

```{r}
# biooracler::list_layers('Ocean Temperature')
# # thetao_baseline_2000_2019_depthmean
# # thetao_baseline_2000_2019_depthsurf
# x <- biooracler::list_layers(simplify = FALSE)

dataset_id <- c('thetao_ssp245_2020_2100_depthmean',
                'thetao_ssp245_2020_2100_depthsurf',
                'thetao_ssp585_2020_2100_depthmean',
                'thetao_ssp585_2020_2100_depthsurf',
                'thetao_baseline_2000_2019_depthmean',
                'thetao_baseline_2000_2019_depthsurf')

out_fs <- file.path(here('_data_raw/biooracle_2026/ocean_temp'), paste0(dataset_id, '.tif'))
out_fs <- str_replace(out_fs, '2020_2100', '2050')
out_fs <- str_replace(out_fs, '2000_2019', '2010')

### Check what's actually in it (time range, variable names, units)
# biooracler::info_layer(dataset_id['surface'])

for (i in seq_along(dataset_id)) {
  ### i <- 1
  id <- dataset_id[i]
  out_f <- out_fs[i]
  
  ### set time constraints to a single decade, depending on dataset
  if(str_detect(id, 'ssp')) constraints <- list(time = c('2050-01-01T00:00:00Z', '2050-01-01T00:00:00Z'))
  else  constraints <- list(time = c('2010-01-01T00:00:00Z', '2010-01-01T00:00:00Z'))

  variables <- c('thetao_mean')
  
  ### Download as a SpatRaster(default)
  if(!file.exists(out_f)) {
    t_layer <- biooracler::download_layers(id, variables, constraints)
    writeRaster(t_layer, out_f)
  }
}

consolidated_f <- here('_data/ocean_temp/thetao_consolidated.parquet')
if(!file.exists(consolidated_f)) {
  r <- rast(out_fs) %>%
    setNames(basename(out_fs) %>% str_remove('.tif'))
  df <- as.data.frame(r, xy = TRUE) %>%
    pivot_longer(-c(x, y), names_to = 'scenario', values_to = 't_mean') %>%
    fmutate(layer = ifelse(str_detect(scenario, 'surf'), 'surface', 'seafloor'),
            scenario = scenario %>% str_remove('_depth.+')) %>%
    fmutate(t_mean = round(t_mean, 3),
            ### repair rounding errors in x and y:
            x = round(x, 3), y = round(y, 3)) %>%
    ### simplify scenarios
    fmutate(scenario = ifelse(str_detect(scenario, 'baseline'), 'current', scenario),
          scenario = str_replace(scenario, 'thetao_ssp[0-9]', 'rcp'))

  write_parquet(df, consolidated_f)
}

ocean_t_df <- read_parquet(consolidated_f)
```

### Calculate Thermal Stressors

For each species: given an Aphia ID, $T^a_{max}$, $T^p_{max}$, and vulnerability score (in a single row data frame):

* Load SDM
* Load appropriate temperature maps as long data frame: 
    * `scenario` = `current`, `rcp45_2050`, `rcp85_2050`
    * `t_mean` = mean annual temp for that scenario
* Intersect temp map with SDM
* Calculate stressor intensity (define separate function and to each column)
* Multiply stressor intensity by vulnerability

```{r}
calc_therm_stress <- function(t, t_a_max, t_p_max, digits = 3) {
  t_stress <- (t - t_p_max) / (t_a_max / t_p_max)
  t_stress[t_stress < 0] <- 0
  t_stress[t_stress > 1] <- 1
  return(round(t_stress, digits))
}

sst_impact_spp <- function(spp_df, t_map, 
                           scen = c('current', 'rcp45_2050', 'rcp85_2050')) {
  spp_sdm <- get_sdm(spp_df$aphia_id, scenario = scen) %>%
    pivot_longer(cols = scen, names_to = 'scenario', values_to = 'presence') %>%
    fsubset(!is.na(presence)) %>%
    select(aphia_id, x, y, scenario)
  spp_t_map <- t_map %>%
    collapse::join(spp_sdm, how = 'inner', on = c('x', 'y', 'scenario')) %>%
    fmutate(therm_stress = calc_therm_stress(t = t_mean, 
                                             t_a_max = spp_df$t_a_max, 
                                             t_p_max = spp_df$t_p_max))
  
}

```

### Iterate over taxa

Consider: do these in taxonomic groups?  Break taxa into chunks of ~300 (more?), process each chunk and summarize; save as temp; then when all chunks done, read in and re-summarize.

```{r}
assign_chunks <- function(df, chunk_size = 500) {
  tx_spp_count <- df$aphia_id %>% n_distinct()
  
  n_chunks <- max(1, round(tx_spp_count / chunk_size))
  df_chunked <- df %>%
    mutate(chunk = rep(1:n_chunks, length.out = tx_spp_count))
  return(df_chunked)
}

sample_decomp2 <- function(df) {
  ### MAJOR REWRITE - Claude suggestions for full vectorization and use of
  ### collapse::fsum etc within group objects
  
  # df <- tx_tmp_dfs %>% fsubset(x < -170 & scenario == 'rcp45_2050')
  g <- GRP(df, ~ x + y + scenario)
  
  n     <- df$n
  mn    <- df$mean
  var_i <- df$sd^2
  skw   <- df$skew
  krt   <- df$kurt
  
  ### broadcast pooled n and pooled mean back to row level (for deviations)
  pool_n_bc    <- fsum(n, g = g, TRA = "replace")
  pool_mean_bc <- fsum(n * mn, g = g, TRA = "replace") / pool_n_bc
  deviation    <- mn - pool_mean_bc
  
  SS <- (n - 1) * var_i
  SS[n == 1] <- 0
  
  ### group-level pooled n / mean / var / sd
  pool_n    <- fsum(n, g = g)
  pool_mean <- fsum(n * mn, g = g) / pool_n
  pool_dev_sum <- fsum(n * deviation^2, g = g)
  pool_SS   <- fsum(SS, g = g) + pool_dev_sum
  pool_var  <- pool_SS / (pool_n - 1)
  pool_sd   <- sqrt(pool_var)
  
  # ### skew
  # SC        <- skw * SS^1.5 / sqrt(n)
  # SC[n <= 2] <- 0
  # pool_SC   <- fsum(SC, g = g) + 3 * fsum(SS * deviation, g = g) + fsum(n * deviation^3, g = g)
  # pool_skew <- sqrt(pool_n) * pool_SC / pool_SS^1.5
  # 
  # ### kurtosis
  # SQ        <- krt * SS^2 / n
  # SQ[n == 1] <-  0
  # pool_SQ   <- fsum(SQ, g = g) + 4 * fsum(SC * deviation, g = g) +
  #              6 * fsum(SS * deviation^2, g = g) + fsum(n * deviation^4, g = g)
  # pool_kurt <- pool_n * pool_SQ / pool_SS^2
  
  summary_df <- cbind(
    g$groups,
    n    = pool_n,
    mean = pool_mean,
    sd   = pool_sd  #,
    # skew = pool_skew,
    # kurt = pool_kurt
  )
  
  return(summary_df)
}
```

```{r}

t_map_list <- list(
  pelagic = ocean_t_df %>% fsubset(layer == 'surface'),
  benthic = ocean_t_df %>% fsubset(layer == 'seafloor')
)

```

```{r}

spp_envelope_df <- get_spp_vuln() %>%
  select(aphia_id, taxon) %>%
  distinct() %>%
  inner_join(thermal_niches, by = 'aphia_id')
tx_vec <- spp_envelope_df$taxon %>% unique() %>% sort()
  
tx_sst_filestem <- here('_output/impact_maps/impact_by_taxon/impact_sst_%s.parquet')

for(tx in tx_vec) {
  ### tx <- tx_vec[2]
  tx_sst_file <- sprintf(tx_sst_filestem, tx)
  if(file.exists(tx_sst_file)) next()
  
  # break into chunks of size ~500?
  tx_df <- spp_envelope_df %>%
    filter(taxon == tx) %>%
    assign_chunks()
  
  chunk_vec <- tx_df$chunk %>% unique() %>% sort()
  
  tmp_tx_sst_fstem <- here('tmp/impact_sst/impact_sst_%s_%s.parquet')
  if(!dir.exists(dirname(tmp_tx_sst_fstem))) dir.create(dirname(tmp_tx_sst_fstem))
  for(i in chunk_vec) {
    ### i <- chunk_vec[1]
    tmp_tx_sst_file <- sprintf(tmp_tx_sst_fstem, tx, i)
    if(file.exists(tmp_tx_sst_file)) next()

    
    tx_chunk_df <- tx_df %>%
      filter(chunk == i)
    spp_vec <- tx_chunk_df$aphia_id
    
    set_collapse(nthreads = 1)
    spp_therm_impact_list <- parallel::mclapply(
      mc.cores = 64,
      X = spp_vec,
      FUN = \(x) {
        ### x <- spp_vec[1]
        spp_df <- tx_chunk_df %>% fsubset(aphia_id == x)
        t_map  <- t_map_list[[spp_df$wcol]]
        out_df <- sst_impact_spp(spp_df, t_map)
        return(out_df)
      })
    set_collapse(nthreads = 4)
    
    ### process list into df, then summarize across scenarios
    tx_chunk_therm_impact_df <- spp_therm_impact_list %>%
      collapse::rowbind() %>%
      select(x, y, scenario, therm_stress)
    tx_chunk_summary_df <- tx_chunk_therm_impact_df %>%
      fgroup_by(x, y, scenario) %>%
      qsu(higher = FALSE) %>%
      as.data.frame() %>%
      tidyr::extract(Variable,
                     into = c("x", "y", "scenario"),
                     regex = "^(-?\\d+\\.\\d+)\\.(-?\\d+\\.\\d+)\\.(.+)$",
                     remove = TRUE) %>%
      fmutate(x = as.numeric(x), y = as.numeric(y))
    
    ### write out temp file
    write_parquet(tx_chunk_summary_df, tmp_tx_sst_file)
  } ### END LOOP OVER CHUNKS
  
  
  
  ###############################
  ###   Assemble temp files   ###
  ###############################
  ### read in all summary chunks, bind together, and resummarize
  tx_tmp_fs <- list.files(dirname(tmp_tx_sst_fstem),
                          pattern = paste0('impact_sst_', tx),
                          full.names = TRUE)

  ### small taxon, only one chunk, just pivot longer and save out
  if(length(chunk_vec) == 1 & length(tx_tmp_fs) == 1) {
    tmp_df <- read_parquet(tx_tmp_fs) %>%
      janitor::clean_names() %>%
      select(x, y, scenario, n, mean, sd) %>%
      pivot_longer(n:sd, names_to = 'stat', values_to = 'value') %>%
      pivot_wider(names_from = 'scenario', values_from = 'value')
                       
    write_parquet(tmp_df, tx_sst_file)
    stopifnot(file.exists(tx_sst_file))
    unlink(tx_tmp_fs)
    next()
  }
  
  if(length(chunk_vec) != length(tx_tmp_fs)) stop('Mismatch between chunk length and temp file length!')
  
  # system.time({
  # ### larger taxon, multiple chunks
  # tx_tmp_dfs1 <- parallel::mclapply(tx_tmp_fs, read_parquet, 
  #                                  mc.cores = min(length(tx_tmp_fs), 24)) %>%
  #   rowbind() %>%
  #   janitor::clean_names() %>%
  #   fsubset(n > 0) 
  # }) ### 27 s
  
  # system.time({
  tx_tmp_dfs <- arrow::open_dataset(tx_tmp_fs) %>%
    rename_with(janitor::make_clean_names) %>%   
    filter(n > 0) %>%                            ### pushed down to Arrow, not R
    collect()                                    ### materializes only the filtered rows
  # }) ### 0.6 s! for identical results
  

  set_collapse(nthreads = 4)
  
  system.time({
  tx_summary_df1 <- tx_tmp_dfs %>%
    fsubset(scenario == 'rcp45_2050' & x < -170) %>%
    fgroup_by(x, y, scenario) %>%
    fsummarize(
      ### sample_decomp() is from common_fxns.R, rewritten
      ### to fix NA errors in utilities::sample.decomp()
      sample_decomp(n = n,
                    sample.mean = mean, sample.sd   = sd,
                    sample.skew = skew, sample.kurt = kurt,
                    include.sd = TRUE)
    ) %>%
    select(-sample.var) %>%
    setNames(str_remove(names(.), 'sample.'))
  })  ### 62.572s

  system.time({
    tx_summary_df2 <- tx_tmp_dfs %>%
      fsubset(scenario == 'rcp45_2050' & x < -170) %>%
      sample_decomp2()
  })  ### 7.87 s
  f <- \(x) ifelse(is.na(x), NA, round(x, 4))
  xx <- tx_summary_df1 %>% mutate(across(where(is.numeric), f)) %>% arrange(x, y, scenario) %>% as.data.frame()
  yy <- tx_summary_df2 %>% mutate(across(where(is.numeric), f)) %>%  arrange(x, y, scenario) %>% as.data.frame()
  row.names(yy) <- NULL
  identical(xx, yy)
  zz <- anti_join(xx, yy)
  qq <- full_join(xx, yy, by = c('x', 'y', 'scenario')) %>% inner_join(zz)
  
  system.time({
  tx_summary_df <- tx_tmp_dfs %>% 
    sample_decomp2()
})
  
  write_parquet(tx_summary_df, tx_sst_file)
  unlink(tx_tmp_fs)
  
}
```

