---
title: Map community change
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

For each pixel, identify for each scenario the changes in community composition by taxon relative to the current composition.

* Count: persistence (current & scenario), colonization (!current & scenario), extirpation (current & !scenario)
* Turnover: SSI $QS = \frac{2a}{2a + b + c}$ where:
    * $a$ is species shared by both sites (scenarios) - persistence
    * $b$ is species unique to first site (current) - extirpation
    * $c$ is species unique to second site (scenario) - colonization

## Data

* AquaX species distribution models from Reygondeau et al. (2026). AquaX: An enhanced and revised AquaMaps framework to model marine species distributions and biodiversity. PLOS ONE, 21(2), e0335823. https://doi.org/10.1371/journal.pone.0335823

## Methods

* Read in SDMs (by slice) for all species in taxon as long data frame
* For each scenario, by row (spp and pixel) classify from presence/absence to persistence/colonization/extirpation
* Group by pixel and classification, and summarize
* Group by pixel and summarize SSI
* Save out as raster or parquet.

Once complete for each taxon, then combine and calculate across all taxa

```{r attach packages}
#| message: false
#| 
library(tidyverse)
library(here)
library(duckdb) 
library(arrow)
library(collapse)
library(terra)

source(here('common_fxns.R'))
```

### Gather species, taxa, and vulnerability scores

```{r gather data}
available_spp <- list.files(here_aquax('SDM')) %>% 
  str_extract('[0-9]+') %>% 
  as.numeric()
aquax_meta <- get_aquax_meta() %>%
  filter(aphia_id %in% available_spp)
spp_tx_df <- get_spp_vuln() %>%
  ### spp vuln includes the taxa we want
  select(aphia_id, taxon) %>%
  distinct() %>%
  filter(aphia_id %in% aquax_meta$aphia_id)

```

### Iterate over taxa

* process_slice to grab sdms for a taxon, join vuln, and process down to summary stats

```{r define helper functions}

process_slice <- function(sdm_df, digits = NULL) {
  ### if no observations in this slice, move along
  if(nrow(sdm_df) == 0) return(data.frame())
  
  ### classify each spp, then join all scenarios
  sdm_tx_vuln_df <- lapply(X = scenario, 
                           FUN = pull_scenario_stats, 
                           d_array = sdm_tx_vuln_array) %>%
    ### purrr::reduce to apply full join programmatically across all scenarios
    purrr::reduce(.f = full_join, by = c('x', 'y', 'stat'))
  if(!is.null(digits)) {
    sdm_tx_vuln_df <- sdm_tx_vuln_df %>%
      ### collapse::fmutate doesn't understand this syntax alas
      collapse::fmutate(across(.cols = all_of(scenario), .fns = round, digits = digits))
  }
  return(sdm_tx_vuln_df)
}

classify_scenario <- function(sc, d_array) {
  df <- d_array[ , , sc] %>%
    as.data.frame() %>%
    janitor::clean_names() %>%
    ### the qsu seems to fill in zero where no observations...
    ### drop those obs here, let the purrr::reduce fill in values as NA
    collapse::fsubset(n > 0) %>%
    ### drop min and max values... keep mean, sd, skew, kurtosis, n
    select(-min, -max) %>%
    ### qsu creates a group column mashing x and y with a '.' separator
    separate_wider_regex(variable, patterns = c(x = '^-?[0-9]{1,3}\\.[0-9]{3}',
                                                '\\.', ### drop the dot
                                                y = '-?[0-9]{1,3}\\.[0-9]{3}$')) %>%
    pivot_longer(cols = -c('x', 'y'), names_to = 'stat', values_to = sc) %>%
    collapse::fmutate(x = as.numeric(x), y = as.numeric(y))

  return(df)
}
```


### Iterate over taxa

For each taxon, create a composition map.  "Map" is a parquet file containing columns:

* x, y (numeric): x/y in degrees for a pixel
* stat: the statistic described in this observation: n, persistence, colonization, extirpation, SSI
* rcp45_2050: projected value of this stat for this pixel for this scenario
* rcp85_2050: projected value of this stat for this pixel for this scenario

```{r}
### set the number of vertical slices by degrees
x_px     <- 360 / 0.05       ### number of pixels in x axis
xslice_w <- 5                ### slice width in degrees
xslice_n <- 360 / xslice_w   ### number of vertical slices
if(x_px %% xslice_n != 0) stop('Choose a slice width that divides evenly into 360!')

```


```{r iterate over taxa}
tx_vec <- spp_tx_df$taxon %>% unique() %>% sort()

scenario <- c('current', 'rcp45_2050', 'rcp85_2050')

for(tx in tx_vec) {
  ### tx <- tx_vec[1]

  out_fstem <- here::here('_output/spp_turnover_maps/turnover_by_taxon/tx_turnover_%s.parquet') 
   
  out_f <- sprintf(out_fstem, tx)
  if(file.exists(out_f)) next()

  message(paste0('Creating ', basename(out_f), '\n'))

  tx_df <- spp_tx_df %>%
      filter(taxon == tx)
  
  #########################################
  ### iterate slices for taxon/stressor ###
  #########################################

  tmp_out_fstem <- here::here('tmp/tx_turnover/tx_turnover_%s_%s.parquet')
  if(!dir.exists(dirname(tmp_out_fstem))) dir.create(dirname(tmp_out_fstem))
  
  ### set the number of vertical slices by degrees
  x_px     <- 360 / 0.05                  ### number of pixels in x axis
  xslice_w <- ifelse(tx == 'fish', 3, 8)  ### slice width in degrees (fish is huge taxon, smaller slices!)
  xslice_n <- 360 / xslice_w              ### number of vertical slices
  if(x_px %% xslice_n != 0) stop('Choose a slice width that divides evenly into 360!')

  slice_vec <- 1:xslice_n

  for(i in slice_vec) {
    ### i <- 1
    ### generate xrange for this iteration.  NOTE: this works
    ### because actual x values fall on pixel centers, not edges
    xrange <- -180 + c((i-1), i) * xslice_w
    
    tmp_out_f <- sprintf(tmp_out_fstem, tx, paste0(xrange, collapse = '_'))
    if(file.exists(tmp_out_f)) next()
    
    #########################################
    ### get sdms for this slice and taxon ###
    #########################################
    sdm_df <- get_sdm(aphia_id = tx_df$aphia_id %>% unique(), 
                      scenario = scenario,
                      xrange = xrange,
                      apply_thresh = TRUE)

    ### FOR EACH SCENARIO:
    future_scens <- scenario[scenario != 'current']

    ### Reclassify each spp/pixel as persistent, extirpated, or colonized
    sdm_reclass <- sdm_df %>%
      ### present = 1, absent = 0
      fmutate(across(all_of(scenario), \(z) !is.na(z))) %>%
      ### extirp = 1 + 0 = 1; coloniz = 0 + 2 = 2; persist = 1 + 2 = 3
      fmutate(across(all_of(future_scens), \(z) .$current + z*2))
    
    ### Summarize across species for each pixel
    reclass_vec <- c('extirpation', 'colonization', 'persistence')
    sdm_summary <- parallel::mclapply(mc.cores = length(reclass_vec),
        X = seq_along(reclass_vec),
        FUN = function(i) {
          sdm_reclass %>%
            fgroup_by(x, y) %>%
            fsummarize(across(all_of(future_scens), \(z) sum(z == i)))
        }) %>% 
      setNames(reclass_vec) %>%
      collapse::rowbind(idcol = 'stat') %>%
      fmutate(stat = as.character(stat))
    
    ### Calculate SSI from persistence/extirp/colonization
    sdm_ssi_df <- parallel::mclapply(mc.cores = length(future_scens),
        X = future_scens,
        FUN = function(sc) {
          ### sc <- future_scens[1]
          ssi_df <- sdm_summary %>%
            select(x, y, stat, scenario = all_of(sc)) %>%
            pivot_wider(names_from = 'stat', values_from = 'scenario') %>%
            fmutate(ssi = 2 * persistence / (2 * persistence + extirpation + colonization)) %>%
            select(x, y, !!sc := ssi)
        }) %>%
      purrr::reduce(.f = full_join, by = c('x', 'y')) %>%
      fmutate(stat = 'ssi')
      
    ### write out to tmp file
    arrow::write_parquet(rowbind(sdm_summary, sdm_ssi_df), tmp_out_f)
    
  } ### END ITERATION OVER SLICES

  ##########################################
  ### assemble slices for taxon turnover ###
  ##########################################

  tmp_fs <- list.files(dirname(tmp_out_fstem), pattern = tx, full.names = TRUE)
  
  complete_tx_turnover_df <- parallel::mclapply(
    X = tmp_fs,
    FUN = arrow::read_parquet,
    mc.cores = 36
  ) %>%
    bind_rows()

  arrow::write_parquet(complete_tx_turnover_df, out_f)
  
  unlink(tmp_fs)  ### clean up temp directory

} ### END LOOP OVER TAXA
```

```{r}
map_fs <- list.files(dirname(out_fstem), full.names = TRUE)

tx_turnover_map <- arrow::read_parquet(map_fs[1]) 

tx_turnover_map %>%
  fsubset(stat == 'persistence') %>%
  select(-stat) %>%
  rast() %>%
  plot(main = 'persistence')

tx_turnover_map %>%
  fsubset(stat == 'ssi') %>%
  select(-stat) %>%
  rast() %>%
  plot(main = 'ssi')
```

