---
title: Map functional change
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

For each pixel, identify for each scenario the changes in functional entity composition relative to the current composition.

* Count: persistence (current & scenario), colonization (!current & scenario), extirpation (current & !scenario)
* Turnover: SSI $QS = \frac{2a}{2a + b + c}$ where:
    * $a$ is FEs shared by both sites (scenarios) - persistence
    * $b$ is FEs unique to first site (current) - extirpation
    * $c$ is FEs unique to second site (scenario) - colonization
* Novelty: which, if any, FE combos exist in future scenarios that don't currently exist?

## Data

* AquaX species distribution models from Reygondeau et al. (2026). AquaX: An enhanced and revised AquaMaps framework to model marine species distributions and biodiversity. PLOS ONE, 21(2), e0335823. https://doi.org/10.1371/journal.pone.0335823
* Function entities from O'Hara et al. (2024). Cumulative human impacts on global marine fauna highlight risk to biological and functional diversity. PLOS ONE, 19(9), e0309788. https://doi.org/10.1371/journal.pone.0309788


## Methods

* Read in SDMs (by slice) for all species in functional entity as long data frame
* For each scenario, by row (spp and pixel) classify from presence/absence to persistence/colonization/extirpation
* Group by pixel and classification, and summarize
* Group by pixel and summarize SSI
* Save out as raster or parquet.

Once complete for each taxon, then combine and calculate across all taxa.

```{r attach packages}
#| message: false
#| 
library(tidyverse)
library(here)
library(arrow)
library(collapse)
library(terra)
library(cowplot)

source(here('common_fxns.R'))
```

### Gather species, traits

```{r gather data}
available_spp <- list.files(here_aquax('SDM')) %>% 
  str_extract('[0-9]+') %>% 
  as.numeric()
aquax_meta <- get_aquax_meta() %>%
  filter(aphia_id %in% available_spp)

spp_fe_df <- read_csv(here('_data/ohara_2024/funct_entity_traits/fe_spp_traits.csv')) %>%
  rename(len = length_cm, wcol = wcol_pos) %>%
  mutate(len = case_when(len <= 1.5 ~ 'tiny',
                         len <= 7   ~ 'vsm',
                         len <= 15  ~ 'sm',
                         len <= 30  ~ 'med',
                         len <= 50  ~ 'ml',
                         len <= 80  ~ 'lg',
                         len <= 150 ~ 'vlg',
                         len >  150 ~ 'huge',
                         TRUE ~ NA_character_),
         len = factor(len, levels = c('tiny', 'vsm', 'sm', 'med', 'ml', 'lg', 'vlg', 'huge'))) %>%
  mutate(trp = case_when(troph <= 2 ~ 'pri',
                         troph <= 3 ~ 'sec',
                         troph <= 4 ~ 'ter',
                         troph >  4 ~ 'apex',
                         TRUE ~ NA_character_)) %>%
  mutate(mob = case_when(str_detect(adult_mob, 'migratory') ~ 'migr',
                         str_detect(adult_mob, 'resident') ~ 'res',
                         str_detect(adult_mob, 'sedentary') ~ 'sed',
                         str_detect(adult_mob, 'sessile') ~ 'ses',
                         TRUE ~ NA_character_),
         mob = factor(mob, levels = c('migr', 'res', 'sed', 'ses'))) %>%
  select(sciname, aphia_id,
         wcol, mob, len, 
         # fec, gen, 
         trp) %>%
  drop_na() %>%
  distinct() %>%
  filter(aphia_id %in% aquax_meta$aphia_id) %>%
  mutate(fe = paste(len, trp, mob, wcol, sep = '_')) %>%
  group_by(fe) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  arrange(desc(n), fe)

table(spp_fe_df$len)
table(spp_fe_df$trp)
table(spp_fe_df$mob)
table(spp_fe_df$wcol)

```


### Iterate over functional entities

For each FE, create a composition map.  "Map" is a parquet file containing columns:

* x, y (numeric): x/y in degrees for a pixel
* stat: the statistic described in this observation: n, persistence, colonization, extirpation, SSI
* rcp45_2050: projected value of this stat for this pixel for this scenario
* rcp85_2050: projected value of this stat for this pixel for this scenario

```{r}
### set the number of vertical slices by degrees
x_px     <- 360 / 0.05       ### number of pixels in x axis
xslice_w <- 5                ### slice width in degrees
xslice_n <- 360 / xslice_w   ### number of vertical slices
if(x_px %% xslice_n != 0) stop('Choose a slice width that divides evenly into 360!')

```


```{r iterate over FEs}
fe_vec <- spp_fe_df$fe %>% unique()

scenario <- c('current', 'rcp45_2050', 'rcp85_2050')

for(f_e in fe_vec) {
  ### f_e <- fe_vec[1]

  out_fstem <- here::here('_output/fe_turnover_maps/turnover_by_fe/fe_turnover_%s.parquet') 
   
  out_f <- sprintf(out_fstem, f_e)
  if(file.exists(out_f)) next()

  message(paste0('Creating ', basename(out_f), '\n'))

  fe_df <- spp_fe_df %>%
      filter(fe == f_e)
  
  #########################################
  ### iterate slices for taxon/stressor ###
  #########################################

  tmp_out_fstem <- here::here('tmp/fe_turnover/fe_turnover_%s_%s.parquet')
  if(!dir.exists(dirname(tmp_out_fstem))) dir.create(dirname(tmp_out_fstem))
  
  ### set the number of vertical slices by degrees
  x_px     <- 360 / 0.05                  ### number of pixels in x axis
  xslice_n <- 360 / xslice_w              ### number of vertical slices
  if(x_px %% xslice_n != 0) stop('Choose a slice width that divides evenly into 360!')

  slice_vec <- 1:xslice_n

  for(i in slice_vec) {
    ### i <- 1
    ### generate xrange for this iteration.  NOTE: this works
    ### because actual x values fall on pixel centers, not edges
    xrange <- -180 + c((i-1), i) * xslice_w
    
    tmp_out_f <- sprintf(tmp_out_fstem, f_e, paste0(xrange, collapse = '_'))
    if(file.exists(tmp_out_f)) next()
    
    ########################################
    ###  get sdms for this slice and FE  ###
    ########################################
    sdm_df <- get_sdm(aphia_id = fe_df$aphia_id %>% unique(), 
                      scenario = scenario,
                      xrange = xrange,
                      apply_thresh = TRUE)
    if(nrow(sdm_df) == 0) {
      ### no spp here, save placeholder and move on!
      write_parquet(data.frame(), tmp_out_f)
      next()
    }

    ### FOR EACH SCENARIO:
    future_scens <- scenario[scenario != 'current']

    ### Reclassify each spp/pixel as persistent, extirpated, or colonized
    sdm_reclass <- sdm_df %>%
      ### present = 1, absent = 0
      fmutate(across(all_of(scenario), \(z) !is.na(z))) %>%
      ### extirp = 1 + 0 = 1; coloniz = 0 + 2 = 2; persist = 1 + 2 = 3
      fmutate(across(all_of(future_scens), \(z) .$current + z*2))
    
    ### Summarize across species for each pixel
    reclass_vec <- c('extirpation', 'colonization', 'persistence')
    sdm_summary <- parallel::mclapply(mc.cores = length(reclass_vec),
        X = seq_along(reclass_vec),
        FUN = function(i) {
          sdm_reclass %>%
            fgroup_by(x, y) %>%
            fsummarize(across(all_of(future_scens), \(z) sum(z == i)))
        }) %>% 
      setNames(reclass_vec) %>%
      collapse::rowbind(idcol = 'stat') %>%
      fmutate(stat = as.character(stat))
    
    ### Calculate SSI from persistence/extirp/colonization
    sdm_ssi_df <- parallel::mclapply(mc.cores = length(future_scens),
        X = future_scens,
        FUN = function(sc) {
          ### sc <- future_scens[1]
          ssi_df <- sdm_summary %>%
            select(x, y, stat, scenario = all_of(sc)) %>%
            pivot_wider(names_from = 'stat', values_from = 'scenario') %>%
            fmutate(ssi = 2 * persistence / (2 * persistence + extirpation + colonization)) %>%
            select(x, y, !!sc := ssi)
        }) %>%
      purrr::reduce(.f = full_join, by = c('x', 'y')) %>%
      fmutate(stat = 'ssi')
      
    ### write out to tmp file
    arrow::write_parquet(rowbind(sdm_summary, sdm_ssi_df), tmp_out_f)
    
  } ### END ITERATION OVER SLICES

  #############################################
  ### assemble slices for FE level turnover ###
  #############################################

  tmp_fs <- list.files(dirname(tmp_out_fstem), pattern = f_e, full.names = TRUE)
  
  complete_fe_turnover_df <- parallel::mclapply(
    X = tmp_fs,
    FUN = arrow::read_parquet,
    mc.cores = 36
  ) %>%
    bind_rows()

  arrow::write_parquet(complete_fe_turnover_df, out_f)
  
  unlink(tmp_fs)  ### clean up temp directory

} ### END LOOP OVER FEs
```

### Combine FE maps

```{r}

all_fe_turnover_f <- here('_output/fe_turnover_maps/all_fe_community_change.parquet')

if(!file.exists(all_fe_turnover_f)) {
  fe_map_fs <- list.files(here('_output/fe_turnover_maps/turnover_by_fe'), 
                        pattern = 'fe_turnover.+.parquet', full.names = TRUE)
  
  fe_vec <- basename(fe_map_fs) %>% str_remove_all('fe_turnover_|.parquet')
  
  fe_maps <- parallel::mclapply(mc.cores = min(length(fe_map_fs), 64),
                                X = fe_map_fs, 
                                FUN = \(z) {
                                  read_parquet(z) %>%
                                    pivot_wider(names_from = stat, values_from = c(rcp45_2050, rcp85_2050))
                                  }) %>%
    setNames(fe_vec) %>%
    rowbind(idcol = 'fe')
  
  ### summarize total spp for FEs: 
  ### * current presence = persistence (already here) + extirpation (disappear in future)
  ### * future presence = persistence (still here) + colonization (new entrants)
  fe_nspp_map_f <- here('_output/fe_turnover_maps/fe_nspp_maps.parquet')
  if(!file.exists(fe_nspp_map_f)) {
    nspp_maps <- fe_maps %>%
      fgroup_by(x, y) %>%
      fsummarize(nspp_current = sum(rcp45_2050_extirpation + rcp45_2050_persistence),
                 nspp_rcp45_2050 = sum(rcp45_2050_persistence + rcp45_2050_colonization),
                 nspp_rcp85_2050 = sum(rcp85_2050_persistence + rcp85_2050_colonization))
    write_parquet(nspp_maps, fe_nspp_map_f)
  }
  
  ### summarize turnover for FEs.  First collapse FE presence to persistence,
  ### colonization, or extirpation.  Each FE is already summarized per
  ### cell, so:
  ### * FE persists if ANY spp in present and ANY spp in future:
  ###     * sum(extirpation + persistence) > 0
  ###     * AND (persistence OR colonization) > 0
  ### * FE colonizes if NO spp in present and ANY spp in future
  ###     * (extirpation AND persistence) == 0
  ###     * AND colonization > 0
  ### * FE is extirpated if ANY spp in present and NO spp in future
  ###     * extirpation > 0
  ###     * AND (persistence AND colonization) == 0
  fe_exists <- fe_maps %>%
    fmutate(exists_current    = (rcp45_2050_extirpation | rcp45_2050_persistence),
            exists_rcp45_2050 = (rcp45_2050_persistence | rcp45_2050_colonization),
            exists_rcp85_2050 = (rcp85_2050_persistence | rcp85_2050_colonization)) %>%
    select(x, y, fe, starts_with('exists'))
  fe_status <- fe_exists %>%
    fmutate(rcp45_2050_extirpation = exists_current & !exists_rcp45_2050,
            rcp45_2050_colonization = !exists_current & exists_rcp45_2050,
            rcp45_2050_persistence = exists_current & exists_rcp45_2050,
            rcp85_2050_extirpation = exists_current & !exists_rcp85_2050,
            rcp85_2050_colonization = !exists_current & exists_rcp85_2050,
            rcp85_2050_persistence = exists_current & exists_rcp85_2050)

  ### Then sum up the three stats across the scenarios acrpss all FEs
  fe_turnover <- fe_status %>%
    select(-starts_with('exists')) %>%
    fgroup_by(x, y) %>%
    fsummarize(across(-fe, .fns = \(x) sum(as.numeric(x))))
  
  ### Then calc SSI for each scenario
  fe_ssi_df <- fe_turnover %>%
    fmutate(rcp45_2050 = 2 * rcp45_2050_persistence / 
              (2 * rcp45_2050_persistence + rcp45_2050_extirpation + rcp45_2050_colonization),
            rcp85_2050 = 2 * rcp85_2050_persistence / 
              (2 * rcp85_2050_persistence + rcp85_2050_extirpation + rcp85_2050_colonization)) %>%
    select(x, y, rcp45_2050, rcp85_2050) %>%
    fmutate(stat = 'ssi')
  
  all_fe_turnover <- fe_turnover %>%
    select(-contains('ssi')) %>%
    setNames(str_replace(names(.), '2050_', '2050X')) %>%
    pivot_longer(cols = -c(x, y), names_to = c('.value', 'stat'), names_sep = 'X') %>%
    rowbind(fe_ssi_df)

  write_parquet(all_fe_turnover, all_fe_turnover_f)
  
  fe_nfe_map_f <- here('_output/fe_turnover_maps/fe_nfe_maps.parquet')
  if(!file.exists(fe_nfe_map_f)) {
    nfe_maps <- fe_exists %>%
      fgroup_by(x, y) %>%
      fsummarize(nfe_current = sum(exists_current),
                 nfe_rcp45_2050 = sum(exists_rcp45_2050),
                 nfe_rcp85_2050 = sum(exists_rcp85_2050))
    write_parquet(nfe_maps, fe_nfe_map_f)
  }


}
```

### Write out raster outputs

Reproject lat/long into Behrmann equal area projections for MS figure displays.  For turnover and functional richness, save out the results as two-layer raster (one for each future scenario).

```{r}
base_rast <- rast(ext(-180, 180, -90, 90),
                  resolution = 0.05,
                  crs = 'epsg:4326')
behr_rast <- base_rast %>%
  project(y = '+proj=cea +lat_ts=30 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs', 
          res = 5000)
```

::: panel-tabset

#### Turnover raster

```{r turnover rasters}
all_fe_turnover_df <- read_parquet(all_fe_turnover_f) 

turnover_out_f <- here('_output/fe_turnover_maps/all_fe_turnover.tif')

if(!file.exists(turnover_out_f)) {

  ssi_r <- all_fe_turnover_df %>%
    fsubset(stat == 'ssi') %>%
    select(-stat) %>%
    rast(crs = crs(base_rast), ext = ext(base_rast)) %>%
    project(behr_rast, res = 5000)
  turnover_r <- 1 - ssi_r
  writeRaster(turnover_r, turnover_out_f)
}

plot(rast(turnover_out_f), main = 'spp turnover')
```

#### FE Richness 

```{r FE richness rasters}
fe_rich_out_f <- here('_output/fe_turnover_maps/all_fe_richness.tif')

if(!file.exists(fe_rich_out_f)) {
  nfe_df <- all_fe_turnover_df %>%
    pivot_longer(cols = c('rcp45_2050', 'rcp85_2050'), names_to = 'scenario', values_to = 'value') %>%
    pivot_wider(names_from = stat, values_from = value) %>%
    fmutate(fe_r_current = persistence + extirpation,
            fe_r_future  = colonization + persistence) %>%
    select(x, y, scenario, fe_r_current, fe_r_future) %>%
    pivot_wider(names_from = scenario, values_from = fe_r_future)

  nfe_r <- nfe_df %>%
    rast(crs = crs(base_rast), ext = ext(base_rast)) %>%
    project(behr_rast, res = 5000)

  writeRaster(nfe_r, fe_rich_out_f)
}
plot(rast(fe_rich_out_f), main = 'fe richness')

```

