---
title: Map community change
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

For each pixel, identify for each scenario the changes in community composition by taxon relative to the current composition.

* Count: persistence (current & scenario), colonization (!current & scenario), extirpation (current & !scenario)
* Turnover: SSI $QS = \frac{2a}{2a + b + c}$ where:
    * $a$ is species shared by both sites (scenarios) - persistence
    * $b$ is species unique to first site (current) - extirpation
    * $c$ is species unique to second site (scenario) - colonization

## Data

* AquaX species distribution models from Reygondeau et al. (2026). AquaX: An enhanced and revised AquaMaps framework to model marine species distributions and biodiversity. PLOS ONE, 21(2), e0335823. https://doi.org/10.1371/journal.pone.0335823

## Methods

* Read in SDMs (by slice) for all species in taxon as long data frame
* For each scenario, by row (spp and pixel) classify from presence/absence to persistence/colonization/extirpation
* Group by pixel and classification, and summarize
* Group by pixel and summarize SSI
* Save out as raster or parquet.

Once complete for each taxon, then combine and calculate across all taxa.

```{r attach packages}
#| message: false
#| 
library(tidyverse)
library(here)
library(duckdb) 
library(arrow)
library(collapse)
library(terra)

source(here('common_fxns.R'))
```

### Gather species, taxa, and vulnerability scores

```{r gather data}
available_spp <- list.files(here_aquax('SDM')) %>% 
  str_extract('[0-9]+') %>% 
  as.numeric()
aquax_meta <- get_aquax_meta() %>%
  filter(aphia_id %in% available_spp)
spp_tx_df <- get_spp_vuln() %>%
  ### spp vuln includes the taxa we want
  select(aphia_id, taxon) %>%
  distinct() %>%
  filter(aphia_id %in% aquax_meta$aphia_id)

```

### Iterate over taxa

For each taxon, create a composition map.  "Map" is a parquet file containing columns:

* x, y (numeric): x/y in degrees for a pixel
* stat: the statistic described in this observation: n, persistence, colonization, extirpation, SSI
* rcp45_2050: projected value of this stat for this pixel for this scenario
* rcp85_2050: projected value of this stat for this pixel for this scenario

```{r}
### set the number of vertical slices by degrees
x_px     <- 360 / 0.05       ### number of pixels in x axis
xslice_w <- 5                ### slice width in degrees
xslice_n <- 360 / xslice_w   ### number of vertical slices
if(x_px %% xslice_n != 0) stop('Choose a slice width that divides evenly into 360!')

```


```{r iterate over taxa}
tx_vec <- spp_tx_df$taxon %>% unique() %>% sort()

scenario <- c('current', 'rcp45_2050', 'rcp85_2050')

for(tx in tx_vec) {
  ### tx <- tx_vec[1]

  out_fstem <- here::here('_output/spp_turnover_maps/turnover_by_taxon/tx_turnover_%s.parquet') 
   
  out_f <- sprintf(out_fstem, tx)
  if(file.exists(out_f)) next()

  message(paste0('Creating ', basename(out_f), '\n'))

  tx_df <- spp_tx_df %>%
      filter(taxon == tx)
  
  #########################################
  ### iterate slices for taxon/stressor ###
  #########################################

  tmp_out_fstem <- here::here('tmp/tx_turnover/tx_turnover_%s_%s.parquet')
  if(!dir.exists(dirname(tmp_out_fstem))) dir.create(dirname(tmp_out_fstem))
  
  ### set the number of vertical slices by degrees
  x_px     <- 360 / 0.05                  ### number of pixels in x axis
  xslice_w <- ifelse(tx == 'fish', 3, 8)  ### slice width in degrees (fish is huge taxon, smaller slices!)
  xslice_n <- 360 / xslice_w              ### number of vertical slices
  if(x_px %% xslice_n != 0) stop('Choose a slice width that divides evenly into 360!')

  slice_vec <- 1:xslice_n

  for(i in slice_vec) {
    ### i <- 1
    ### generate xrange for this iteration.  NOTE: this works
    ### because actual x values fall on pixel centers, not edges
    xrange <- -180 + c((i-1), i) * xslice_w
    
    tmp_out_f <- sprintf(tmp_out_fstem, tx, paste0(xrange, collapse = '_'))
    if(file.exists(tmp_out_f)) next()
    
    #########################################
    ### get sdms for this slice and taxon ###
    #########################################
    sdm_df <- get_sdm(aphia_id = tx_df$aphia_id %>% unique(), 
                      scenario = scenario,
                      xrange = xrange,
                      apply_thresh = TRUE)

    ### FOR EACH SCENARIO:
    future_scens <- scenario[scenario != 'current']

    ### Reclassify each spp/pixel as persistent, extirpated, or colonized
    sdm_reclass <- sdm_df %>%
      ### present = 1, absent = 0
      fmutate(across(all_of(scenario), \(z) !is.na(z))) %>%
      ### extirp = 1 + 0 = 1; coloniz = 0 + 2 = 2; persist = 1 + 2 = 3
      fmutate(across(all_of(future_scens), \(z) .$current + z*2))
    
    ### Summarize across species for each pixel
    reclass_vec <- c('extirpation', 'colonization', 'persistence')
    sdm_summary <- parallel::mclapply(mc.cores = length(reclass_vec),
        X = seq_along(reclass_vec),
        FUN = function(i) {
          sdm_reclass %>%
            fgroup_by(x, y) %>%
            fsummarize(across(all_of(future_scens), \(z) sum(z == i)))
        }) %>% 
      setNames(reclass_vec) %>%
      collapse::rowbind(idcol = 'stat') %>%
      fmutate(stat = as.character(stat))
    
    ### Calculate SSI from persistence/extirp/colonization
    sdm_ssi_df <- parallel::mclapply(mc.cores = length(future_scens),
        X = future_scens,
        FUN = function(sc) {
          ### sc <- future_scens[1]
          ssi_df <- sdm_summary %>%
            select(x, y, stat, scenario = all_of(sc)) %>%
            pivot_wider(names_from = 'stat', values_from = 'scenario') %>%
            fmutate(ssi = 2 * persistence / (2 * persistence + extirpation + colonization)) %>%
            select(x, y, !!sc := ssi)
        }) %>%
      purrr::reduce(.f = full_join, by = c('x', 'y')) %>%
      fmutate(stat = 'ssi')
      
    ### write out to tmp file
    arrow::write_parquet(rowbind(sdm_summary, sdm_ssi_df), tmp_out_f)
    
  } ### END ITERATION OVER SLICES

  ##########################################
  ### assemble slices for taxon turnover ###
  ##########################################

  tmp_fs <- list.files(dirname(tmp_out_fstem), pattern = tx, full.names = TRUE)
  
  complete_tx_turnover_df <- parallel::mclapply(
    X = tmp_fs,
    FUN = arrow::read_parquet,
    mc.cores = 36
  ) %>%
    bind_rows()

  arrow::write_parquet(complete_tx_turnover_df, out_f)
  
  unlink(tmp_fs)  ### clean up temp directory

} ### END LOOP OVER TAXA
```

### Combine taxa maps

```{r}
all_spp_turnover_f <- here('_output/spp_turnover_maps/all_spp_turnover.parquet')

if(!file.exists(all_spp_turnover_f)) {
  tx_map_fs <- list.files(here('_output/spp_turnover_maps/turnover_by_taxon'), 
                        pattern = 'tx_turnover.+.parquet', full.names = TRUE)
  
  tx_vec <- basename(tx_map_fs) %>% str_remove_all('tx_turnover_|.parquet')
  
  tx_maps <- parallel::mclapply(mc.cores = length(tx_map_fs),
                                X = tx_map_fs, 
                                FUN = \(z) {
                                  read_parquet(z) %>%
                                    pivot_wider(names_from = stat, values_from = c(rcp45_2050, rcp85_2050))
                                  }) %>%
    setNames(tx_vec) %>%
    rowbind(idcol = 'taxon')
  
  all_maps <- tx_maps %>%
    fgroup_by(x, y) %>%
    fsummarize(across(-c(x, y, taxon), sum))
  
  pers_colo_extirp <- all_maps %>%
    select(-contains('ssi')) %>%
    setNames(str_replace(names(.), '2050_', '2050X')) %>%
    pivot_longer(cols = -c(x, y), names_to = c('.value', 'stat'), names_sep = 'X')
  
  ssi_df <- all_maps %>%
    fmutate(rcp45_2050 = 2 * rcp45_2050_persistence / (2 * rcp45_2050_persistence + rcp45_2050_extirpation + rcp45_2050_colonization),
            rcp85_2050 = 2 * rcp85_2050_persistence / (2 * rcp85_2050_persistence + rcp85_2050_extirpation + rcp85_2050_colonization)) %>%
    select(x, y, rcp45_2050, rcp85_2050) %>%
    fmutate(stat = 'ssi')
  
  all_spp_turnover <- rowbind(pers_colo_extirp, ssi_df)
  
  write_parquet(all_spp_turnover, all_spp_turnover_f)
}
```

### Maps

:::panel-tabset
#### Persistence 

Number of spp in both current period and future

```{r persistence map}
#| fig-height: 3

all_spp_turnover <- read_parquet(all_spp_turnover_f) 

all_spp_turnover %>%
  fsubset(stat == 'persistence') %>%
  select(-stat) %>%
  rast() %>%
  plot()
```

#### Colonization 

Number of spp in future period but NOT in current

```{r colonization map}
#| fig-height: 3

all_spp_turnover %>%
  fsubset(stat == 'colonization') %>%
  select(-stat) %>%
  rast() %>%
  plot()
```

#### Extirpation 

Number of spp in current period but NOT future

```{r extirpation map}
#| fig-height: 3

all_spp_turnover %>%
  fsubset(stat == 'extirpation') %>%
  select(-stat) %>%
  rast() %>%
  plot()
```

#### Similarity 

Sorensen similarity index: 1 = identical community, 0 = complete turnover

```{r similarity map}
#| fig-height: 3

all_spp_turnover %>%
  fsubset(stat == 'ssi') %>%
  select(-stat) %>%
  rast() %>%
  plot()
```

