---
title: Map species and functional change against human impacts
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

Compare turnover metrics against areas of high human impact.

## Data

* AquaX species distribution models from Reygondeau et al. (2026). AquaX: An enhanced and revised AquaMaps framework to model marine species distributions and biodiversity. PLOS ONE, 21(2), e0335823. https://doi.org/10.1371/journal.pone.0335823
* Function entities from O'Hara et al. (2024). Cumulative human impacts on global marine fauna highlight risk to biological and functional diversity. PLOS ONE, 19(9), e0309788. https://doi.org/10.1371/journal.pone.0309788
* Impacts from Halpern, B. S., Frazier, M., O’Hara, C. C., Vargas-Fonseca, O. A., & Lombard, A. T. (2025). Cumulative impacts to global marine ecosystems projected to more than double by midcentury. Science, 0(0), eadv2906. https://doi.org/10.1126/science.adv2906

## Methods

* Reproject impacts to same resolution as spp and FE turnover maps
* Rasterize the SPP and FE turnover maps
* Ideas for potential zonal comparison:
    * CHI categories/quartiles?
    * Turnover quartiles?
* Other ideas for comparison
    * Regression?
    * PCA?
* NOTE: should the quartile breakdowns be implemented using an equal area projection, to avoid overemphasizing arctic and antarctic regions at the expense of equatorial regions?
    * YES! we will be using the behrmann projection, equal area cylindrical

```{r}
library(tidyverse)
library(here)
library(terra)
library(sf)
library(arrow)
library(collapse)
library(biscale)  ### for bivariate legend
library(cowplot)
```

### Reproject impacts

Let's reproject impacts into the Behrmann projection and save out as rasters (current, RCP4.5 2050, RCP8.5 2050)

```{r set up base and behrmann rasters}
base_rast <- rast(ext(-180, 180, -90, 90),
                  resolution = 0.05,
                  crs = 'epsg:4326')
behr_rast <- base_rast %>%
  project(y = '+proj=cea +lat_ts=30 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs', 
          res = 5000)
```
```{r reproject impacts}
impact_fs_moll <- data.frame(
    f = list.files(here('_data/halpern_2025/impacts_by_pressure'), 
                  pattern = '.tif', full.names = TRUE)) %>%
  mutate(stressor = sub('_ssp.+', '', basename(f)),
         scenario = gsub('[a-z-]+_|.tif', '', basename(f)),
         scenario = sub('medium-term', '2050', scenario)) %>%
  filter(scenario != 'ssp585_current')

out_fstem <- here('_data/halpern_2025/impacts_by_pressure_reprojected/%s_behr.tif')

for(str in unique(impact_fs_moll$stressor)) {
  ### str <- unique(impact_fs_moll$stressor)[1]
  f_out <- sprintf(out_fstem, str)
  if(file.exists(f_out)) next()

  str_df <- impact_fs_moll |> 
    filter(stressor == str) 
  file_vec <- str_df |> 
    pull(f)
  str_rasts <- rast(file_vec) |>
    setNames(str_df$scenario)
  str_rasts_behr <- str_rasts |>
    project(behr_rast)

  writeRaster(str_rasts_behr, f_out)
}
```

### Gather impacts

Here, focus on manageable impacts only, summarize per future scenarios

```{r}
impact_fs <- list.files(dirname(out_fstem), 
                        pattern = '.tif$', full.names = TRUE)
### focus on potentially manageable stressors?
impact_fs <- impact_fs[grepl('coast|fish|pollution', basename(impact_fs))]

impact_r <- rast(impact_fs)
impact_rcp45 <- impact_r[[str_detect(names(impact_r), 'ssp245_2050')]] |> sum(na.rm = TRUE)
impact_rcp85 <- impact_r[[str_detect(names(impact_r), 'ssp585_2050')]] |> sum(na.rm = TRUE)

impact_sum_r <- c(impact_rcp45, impact_rcp85) |>
  setNames(c('rcp45_2050', 'rcp85_2050'))
# plot(impact_sum_r)
```

### Read in and rasterize turnover metrics

THese rasters were created in scripts 3 (spp turnover) and 4 (fe turnover) in the Behrmann projection.

```{r read rasts}

spp_turn_r <- rast(here('_output/spp_turnover_maps/all_spp_turnover.tif'))
fe_turn_r  <- rast(here('_output/fe_turnover_maps/all_fe_turnover.tif'))

# plot(spp_turn_r); plot(fe_turn_r)
```

## Generate Manuscript Figures

Locate tropics lines and equator as well

```{r set up land and parallels}
land_sf <- read_sf(here('_data/eezs/esri_world_continents.gpkg')) %>%
  st_transform(crs(behr_rast))

# trop_df <- data.frame(x = 0, y = c(23.425, -23.425, 0)) |> xy_to_cell_id(drop = FALSE)
### cancer: 9586800, capricorn: 16333200
# cell_id_behr_df <- base_rast |> 
#   setValues(1:ncell(base_rast)) |> 
#   project(behr_rast) |> 
#   round() |>
#   as.data.frame(xy = TRUE) |>
#   filter(lyr.1 %in% trop_df$cell_id)
### cancer: 2907724.47007114; capricorn: -2907275.52992886

plot_map <- function(data, fill_var) {
  ggplot() +
    geom_raster(data = data, aes(x, y, fill = {{ fill_var }})) +
    ### continents:
    geom_sf(data = land_sf,
            fill = 'grey96', color = 'grey30', 
            size = .1) +
    ### tropics and equator:
    geom_hline(yintercept = c(2907724.470, -2907275.530), 
               linetype = 'dashed', linewidth = 0.25, color = 'grey30') +
    geom_hline(yintercept = 0, 
               linetype = 'solid', linewidth = 0.25, color = 'grey30') +
    ### standardized styling:
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_void()
}
```

### Map of SPP and FE richness

Current and change 
```{r spp and fe richness}
fig_f <- here('ms_figs/figX_spp_fe_richness.png')

if(!file.exists(fig_f)) {
  nspp_r <- rast(here('_output/spp_turnover_maps/all_spp_richness.tif'))
  nfe_r  <- rast(here('_output/fe_turnover_maps/all_fe_richness.tif'))

  ### Diverging palette from magenta to dark green, with yellow tone
  ### in middle - based on hcl.colors(pal = 'Red-Green')
  div_pal <- c("#841859", "#FFFFEE", "#005600")

  nspp_df <- as.data.frame(nspp_r, xy = TRUE) %>%
    fmutate(log_n = log10(spp_r_current),
            diff_rcp45_2050 = (rcp45_2050 - spp_r_current) / spp_r_current, 
            diff_rcp85_2050 = (rcp85_2050 - spp_r_current) / spp_r_current) %>%
    ### Treat anything over 1 as 1(+) (Inf: zero spp now 1), 
    ### Treat NaN as zero (zero spp still zero)
    fmutate(diff_rcp45_2050 = case_when((diff_rcp45_2050 > 1) ~ 1,
                                        is.nan(diff_rcp45_2050) ~ 0,
                                        TRUE ~ diff_rcp45_2050)) %>%
    fmutate(diff_rcp85_2050 = case_when((diff_rcp85_2050 > 1) ~ 1,
                                        is.nan(diff_rcp85_2050) ~ 0,
                                        TRUE ~ diff_rcp85_2050))

  nfe_df  <- as.data.frame(nfe_r, xy = TRUE) %>%
    fmutate(log_n = log10(fe_r_current),
            diff_rcp45_2050 = (rcp45_2050 - fe_r_current) / fe_r_current, 
            diff_rcp85_2050 = (rcp85_2050 - fe_r_current) / fe_r_current) %>% 
    fmutate(diff_rcp45_2050 = case_when((diff_rcp45_2050 > 1) ~ 1,
                                        is.nan(diff_rcp45_2050) ~ 0,
                                        TRUE ~ diff_rcp45_2050)) %>%
    fmutate(diff_rcp85_2050 = case_when((diff_rcp85_2050 > 1) ~ 1,
                                        is.nan(diff_rcp85_2050) ~ 0,
                                        TRUE ~ diff_rcp85_2050))
    
  nspp_max <- max(nspp_df$spp_r_current)
  nfe_max  <- max(nfe_df$fe_r_current)
  
  nspp_panel <- plot_map(data = nspp_df, 
                         fill_var = log_n) +
    scale_fill_viridis_c(breaks = c(0, 1, 2, 3, log10(nspp_max)), 
                         labels = c(1, 10, 100, 1000, nspp_max)) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(N[spp]))

  nfe_panel <- plot_map(data = nfe_df, 
                         fill_var = log_n) +
    scale_fill_viridis_c(breaks = c(0, 1, 2, log10(nfe_max)), 
                        labels = c(1, 10, 100, nfe_max)) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(N[FE]))

  dspp_panel <- plot_map(data = nspp_df, 
                         fill_var = diff_rcp45_2050) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression('∆' * N[spp]))

  dfe_panel <- plot_map(data = nfe_df, 
                        fill_var = diff_rcp45_2050) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression('∆' * N[FE]))

  p <- ggdraw() +
    draw_plot(nfe_panel, x = .02, y = 0, height = .5, width = .48) +
    draw_plot(dfe_panel, x = .51, y = 0, height = .5, width = .48) +
    draw_plot(nspp_panel, x = .02, y = .50, height = .5, width = .48) +
    draw_plot(dspp_panel, x = .51, y = .50, height = .5, width = .48) +
    draw_label('A', x = 0, y = 1, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('B', x = .51, y = 1, vjust = 1, hjust = 1, size = 14, fontface = 'bold') +
    draw_label('C', x = 0, y = .5, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('D', x = .51, y = .5, vjust = 1, hjust = 1, size = 14, fontface = 'bold')

  ggsave(fig_f, height = 3.8, width = 10, dpi = 300)
}
knitr::include_graphics(fig_f)
```

#### Focus on apex spp: richness and pct change
```{r apex spp richness}
fig_f <- here('ms_figs/figX_apex_spp_richness.png')

if(!file.exists(fig_f)) {
  nspp_r <- rast(here('_output/spp_turnover_maps/apex_spp_richness.tif'))

  ### Diverging palette from magenta to dark green, with yellow tone
  ### in middle - based on hcl.colors(pal = 'Red-Green')
  div_pal <- c("#841859", "#FFFFEE", "#005600")

  nspp_df <- as.data.frame(nspp_r, xy = TRUE) %>%
    fmutate(log_n = log10(spp_r_current),
            diff_rcp45_2050 = (rcp45_2050 - spp_r_current) / spp_r_current, 
            diff_rcp85_2050 = (rcp85_2050 - spp_r_current) / spp_r_current) %>%
    ### Treat anything over 1 as 1(+) (Inf: zero spp now 1), 
    ### Treat NaN as zero (zero spp still zero)
    fmutate(diff_rcp45_2050 = case_when((diff_rcp45_2050 > 1) ~ 1,
                                        is.nan(diff_rcp45_2050) ~ 0,
                                        TRUE ~ diff_rcp45_2050)) %>%
    fmutate(diff_rcp85_2050 = case_when((diff_rcp85_2050 > 1) ~ 1,
                                        is.nan(diff_rcp85_2050) ~ 0,
                                        TRUE ~ diff_rcp85_2050))
    
  nspp_max <- max(nspp_df$spp_r_current)
  nspp_brks <- seq(1, log10(nspp_max), 1)
  nspp_panel <- plot_map(data = nspp_df, fill_var = log_n) +
    scale_fill_viridis_c(breaks = c(0, 1, 2, 3, log10(nspp_max)), 
                        labels = c(1, 10, 100, 1000, nspp_max)) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(N[spp]))



  dspp_panel <- plot_map(data = nspp_df, fill_var = diff_rcp45_2050) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression('∆' * N[spp]))


  p <- ggdraw() +
    draw_plot(nspp_panel, x = .02, y = 0.5, height = .5, width = .98) +
    draw_plot(dspp_panel, x = .02, y = 0.0, height = .5, width = .98) +
    draw_label('A', x = 0, y = 1.0,  vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('B', x = 0, y = 0.5, vjust = 1, hjust = 0, size = 14, fontface = 'bold')
  ggsave(fig_f, height = 6, width = 8, dpi = 300)
}
knitr::include_graphics(fig_f)
```

#### Focus on apex spp: relative change vs other spp
```{r apex spp relative change}
fig_f <- here('ms_figs/figX_apex_spp_vs_other_spp.png')

if(!file.exists(fig_f)) {
  apex_r <- rast(here('_output/spp_turnover_maps/apex_spp_richness.tif'))
  all_r  <- rast(here('_output/spp_turnover_maps/all_spp_richness.tif'))

  ### Diverging palette orange-purple to indicate difference from pink-green
  div_pal <- c("#f1a340", "#f1f1f1", "#998ec3")

  apex_df <- as.data.frame(apex_r, xy = TRUE) %>%
    select(x, y, apex_current = spp_r_current, apex_2050 = rcp45_2050)
   
  all_df <- as.data.frame(all_r, xy = TRUE) %>%
    select(x, y, all_current = spp_r_current, all_2050 = rcp45_2050)

  compare_df <- apex_df %>%
    join(all_df, how = 'full', on = c('x', 'y')) %>%
    fmutate(other_current = all_current - apex_current,
            other_2050    = all_2050 - apex_2050,
            apex_diff     = (apex_2050 - apex_current) / apex_current,
            other_diff    = (other_2050 - other_current) / other_current) %>%
    fmutate(delta = apex_diff - other_diff,
            delta = ifelse(delta > 1, 1, delta),
            delta = ifelse(delta < -1, -1, delta))

  diff_panel <- plot_map(data = compare_df, fill_var = delta) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression('∆' * N[pred] - '∆' * N[other]))


  # p <- ggdraw() +
  #   draw_plot(nspp_panel, x = .02, y = 0.5, height = .5, width = .98) +
  #   draw_plot(dspp_panel, x = .02, y = 0.0, height = .5, width = .98) +
  #   draw_label('A', x = 0, y = 1.0,  vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
  #   draw_label('B', x = 0, y = 0.5, vjust = 1, hjust = 0, size = 14, fontface = 'bold')
  ggsave(fig_f, height = 3.1, width = 8, dpi = 300)
}
knitr::include_graphics(fig_f)
```

### Map of turnover

Compare species and FE turnover in a single plot, just examining RCP 4.5 (8.5 can go in SI?)
```{r spp and fe turnover}
fig_file <- here('ms_figs/figX_spp_fe_turnover.png')

if(!exists(fig_file)) {
  turnover_rasts <- c(spp_turn_r[['rcp45_2050']], fe_turn_r[['rcp45_2050']])
  turnover_df <- as.data.frame(turnover_rasts, xy = TRUE) %>%
    setNames(c('x', 'y', 'Species Turnover', 'Functional Entity Turnover')) %>%
    pivot_longer(cols = -c(x, y), names_to = 'var', values_to = 'val') %>%
    mutate(var = factor(var, levels = c('Species Turnover', 'Functional Entity Turnover'))) 

  spp_fe_turnover_plot <- plot_map(data = turnover_df, fill_var = val) +
    scale_fill_viridis_c() +
    theme(axis.text = element_blank(), 
          axis.title = element_blank(),
          strip.text.y.left = element_text(angle = 90)) +
    facet_wrap(~ var, ncol = 1, strip.position = 'left') +
    labs(fill = expression(beta[.Sør]))

  fig_file <- here('ms_figs/figX_spp_fe_turnover.png')
  ggsave(fig_file, height = 6, width = 8, dpi = 300)
}
knitr::include_graphics(fig_file)
```

### Map of turnover by component

Break turnover into extirpation and colonization components, plot as 4-panel.

```{r spp and fe turnover components}
fig_file <- here('ms_figs/figX_spp_fe_turnover_components.png')

if(!file.exists(fig_file)) {
  turnover_rasts <- c(spp_turn_r[['rcp45_2050']], fe_turn_r[['rcp45_2050']])

  spp_turn_ext_r <- rast(here('_output/spp_turnover_maps/all_spp_turnover_extirp.tif'))
  fe_turn_ext_r  <- rast(here('_output/fe_turnover_maps/all_fe_turnover_extirp.tif'))

  extirp_rasts <- c(spp_turn_ext_r[['rcp45_2050']], fe_turn_ext_r[['rcp45_2050']]) %>%
    setNames(c('SpeciesXExtirpation', 'FEXExtirpation'))
  coloniz_rasts <- (turnover_rasts - extirp_rasts) %>%
    setNames(c('SpeciesXColonization', 'FEXColonization'))

  turn_component_rasts <- c(extirp_rasts, coloniz_rasts)

  turn_comp_df <- as.data.frame(turn_component_rasts, xy = TRUE) %>%
    pivot_longer(cols = -c(x, y), names_to = 'var', values_to = 'val') %>%
    fmutate(unit = sub('X.+', '', var),
            comp = sub('.+X', '', var)) %>%
    fmutate(unit = factor(unit, levels = c('Species', 'FE')),
            comp = factor(comp, levels = c('Extirpation', 'Colonization')))

  spp_fe_turn_comp_plot <- plot_map(data = turn_comp_df, fill_var = val) +
    scale_fill_viridis_c() +
    theme(strip.text.y.left = element_text(angle = 90),
          axis.text = element_blank(), 
          axis.title = element_blank()) +
    facet_grid(unit ~ comp, switch = 'y') +
    labs(fill = expression(beta[.Sør]))

  ggsave(fig_file, height = 4, width = 10, dpi = 300)
}
knitr::include_graphics(fig_file)
```

### Map impacts vs turnover

Here let's do a bivariate plot to compare areas of high/low predicted impact to high/low projected turnover.  

```{r process to quartiles}
fe_turn_impact_f  <- here('_output/figure_data/fig_fe_turnover_impact.parquet')
spp_turn_impact_f <- here('_output/figure_data/fig_spp_turnover_impact.parquet')

if(any(!file.exists(c(fe_turn_impact_f, spp_turn_impact_f)))) {
  impact_qtile <- fe_turn_qtile <- spp_turn_qtile <- spp_turn_r ### initialize three new maps

  ### Set breaks for impacts
  impact_thresholds <- quantile(values(impact_sum_r[[2]]), 
                                probs = seq(0, 1, 0.25), 
                                na.rm = TRUE, names = FALSE)
  # [1] 0.000 0.081 0.096 0.144 1.550
  ### NOTE: between RCP4.5 and 8.5, the 25% and 50% remain the same, but 75% and 100% increase
  values(impact_qtile) <- data.table::fcase(values(impact_sum_r) < impact_thresholds[2], 1L,
                                            values(impact_sum_r) < impact_thresholds[3], 2L,
                                            values(impact_sum_r) < impact_thresholds[4], 3L,
                                            !is.na(values(impact_sum_r)), 4L)

  ### Set breaks for turnover
  ### Note: since turnover is a physical thing, just set as .25, .5, .75 and call it a day.
  ### Set zero turnover to be category 1
  values(fe_turn_qtile)  <- ceiling(values(fe_turn_r * 4))
  fe_turn_qtile[fe_turn_qtile == 0] <- 1
  values(spp_turn_qtile) <- ceiling(values(spp_turn_r * 4))
  spp_turn_qtile[spp_turn_qtile == 0] <- 1

  fe_turn_df <- as.data.frame(fe_turn_qtile, xy = TRUE) %>%
    setNames(c('x', 'y', 'fe_turn_rcp45', 'fe_turn_rcp85'))
  spp_turn_df <- as.data.frame(spp_turn_qtile, xy = TRUE) %>%
    setNames(c('x', 'y', 'spp_turn_rcp45', 'spp_turn_rcp85'))
  impact_df <- as.data.frame(impact_qtile, xy = TRUE) %>%
    setNames(c('x', 'y', 'impact_rcp45', 'impact_rcp85'))

  fe_turn_impact_df <- collapse::join(fe_turn_df, impact_df, 
    how = 'left', on = c('x', 'y')) %>%
    mutate(bi_class_rcp45 = paste(fe_turn_rcp45, impact_rcp45, sep = '-'),
          bi_class_rcp85 = paste(fe_turn_rcp85, impact_rcp85, sep = '-'))
  spp_turn_impact_df <- collapse::join(spp_turn_df, impact_df, 
    how = 'left', on = c('x', 'y')) %>%
    mutate(bi_class_rcp45 = paste(spp_turn_rcp45, impact_rcp45, sep = '-'),
           bi_class_rcp85 = paste(spp_turn_rcp85, impact_rcp85, sep = '-'))

  write_parquet(fe_turn_impact_df,  fe_turn_impact_f)
  write_parquet(spp_turn_impact_df, spp_turn_impact_f)
}

fe_turn_impact_df  <- read_parquet(fe_turn_impact_f)
spp_turn_impact_df <- read_parquet(spp_turn_impact_f)

```

```{r set up biplot}
calc_biplot_pcts <- function(df, scenario = 'rcp45') {
  df_sum <- df %>% 
    select(bi_class = all_of(paste0('bi_class_', scenario))) %>%
    fsubset(!grepl('NA', bi_class)) %>%
    fcount(bi_class) %>%
    mutate(pct = N / sum(N)) %>%
    separate_wider_delim(bi_class, names = c('turn', 'impact'), delim = '-') %>%
    mutate(turn = as.integer(turn), impact = as.integer(impact),
           lbl = paste0(round(pct, 3) * 100, '%'),
           dark = impact >= 3 & turn >= 3)
}
fe_turn_pcts_rcp45 <- fe_turn_impact_df %>%
  calc_biplot_pcts(scenario = 'rcp45')
fe_turn_pcts_rcp85 <- fe_turn_impact_df %>%
  calc_biplot_pcts(scenario = 'rcp85')
spp_turn_pcts_rcp45 <- spp_turn_impact_df %>%
  calc_biplot_pcts(scenario = 'rcp45')
spp_turn_pcts_rcp85 <- spp_turn_impact_df %>%
  calc_biplot_pcts(scenario = 'rcp85')
```
```{r build palette}

map_pal_raw <- bi_pal(pal = 'GrPink2', dim = 4, preview = FALSE)
  
map_pal_mtx      <- matrix(map_pal_raw, nrow = 4, ncol = 4)
map_pal_mtx[3, ] <- colorspace::lighten(map_pal_mtx[3, ], .1)
map_pal_mtx[2, ] <- colorspace::lighten(map_pal_mtx[2, ], .2)
map_pal_mtx[1, ] <- colorspace::lighten(map_pal_mtx[1, ], .3)
map_pal_mtx[ , 3] <- colorspace::lighten(map_pal_mtx[ , 3], .1)
map_pal_mtx[ , 2] <- colorspace::lighten(map_pal_mtx[ , 2], .2)
map_pal_mtx[ , 1] <- colorspace::lighten(map_pal_mtx[ , 1], .3)
map_pal_mtx[1, 1] <- '#ffffee'

map_pal <- as.vector(map_pal_mtx) %>% setNames(names(map_pal_raw))

panel_lgd_raw <- bi_legend(pal = map_pal, dim = 4,
                           x = 'Turnover',
                           y = 'Impact')
pnl_lgd_fe <- panel_lgd_raw +
  geom_text(data = fe_turn_pcts_rcp45, 
            aes(x = turn, y = impact, label = lbl, color = dark),
            size = 3, show.legend = FALSE) +
  scale_color_manual(values = c('grey20', 'white'))
pnl_lgd_spp <- panel_lgd_raw +
  geom_text(data = spp_turn_pcts_rcp45, 
            aes(x = turn, y = impact, label = lbl, color = dark),
            size = 3, show.legend = FALSE) +
  scale_color_manual(values = c('grey20', 'white'))
```

#### FE turnover vs impacts, RCP4.5
```{r fe_45}
panel_fe_45 <- plot_map(data = fe_turn_impact_df, fill_var = bi_class_rcp45) +
  bi_scale_fill(pal = map_pal, dim = 4) +
  theme(axis.text = element_blank(), 
        axis.title = element_blank(),
        legend.position = 'none')

# panel_fe_45
```

#### FE turnover vs impacts, RCP8.5
```{r fe_85}
panel_fe_85 <- plot_map(data = fe_turn_impact_df, fill_var = bi_class_rcp85) +
  bi_scale_fill(pal = map_pal, dim = 4) +
  theme(axis.text = element_blank(), 
        axis.title = element_blank(),
        legend.position = 'none')

# panel_fe_85
```

#### Spp turnover vs impacts, RCP4.5
```{r spp_45}
panel_spp_45 <- plot_map(data = spp_turn_impact_df, fill_var = bi_class_rcp45) +
  bi_scale_fill(pal = map_pal, dim = 4) +
  theme(axis.text = element_blank(), 
        axis.title = element_blank(),
        legend.position = 'none')

# panel_spp_45
```

#### Spp turnover vs impacts, RCP8.5
```{r spp_85}
panel_spp_85 <- plot_map(data = spp_turn_impact_df, fill_var = bi_class_rcp85) +
  bi_scale_fill(pal = map_pal, dim = 4) +
  theme(axis.text = element_blank(), 
        axis.title = element_blank(),
        legend.position = 'none')

# panel_spp_85
```

#### Assemble turnover X impact figure

Figure shows RCP4.5 spp and FE turnover vs. impacts.  Impacts are same, but FE turnover will be less than SPP turnover.

```{r pull panels}
# panel_a <- get_panel(panel_spp_45)
# panel_b <- get_panel(panel_fe_45)
panel_a <- panel_spp_45
panel_b <- panel_fe_45
bi_lgd_fe  <- pnl_lgd_fe
bi_lgd_spp <- pnl_lgd_spp
```

```{r assemble panels}
fig <- ggdraw() +
  draw_plot(pnl_lgd_spp,  x = .68, y = .55, width = .40, height = .45) +
  draw_plot(panel_spp_45, x = .012, y = 0.5, width = .74, height = .5) +
  draw_plot(pnl_lgd_fe,   x = .68, y = .05, width = .40, height = .45) +
  draw_plot(panel_fe_45,  x = .012, y = 0.0, width = .74, height = .5) +
  draw_label('A', x = 0,  y = 1.0, hjust = 0, vjust = 1, size = 16, fontface = 'bold') +
  draw_label('B', x = 0,  y = 0.5, hjust = 0, vjust = 1, size = 16, fontface = 'bold')
 
fig_file <- here('ms_figs/figX_spp_fe_impacts.png')
ggsave(fig_file, height = 5.75, width = 9.5, dpi = 300)
knitr::include_graphics(fig_file)
```