---
title: Test methods for large-scale data wrangling
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

## Summary

For each of the relevant uniform-exposure stressors, i.e., those that do not depend on species-specific interactions, for each pixel calculate a mean vulnerability across each taxonomic group and overall.  Calculate summary statistics as well - variance, skew, kurtosis.  This will be done for the current, RCP 4.5, and RCP 8.5 scenarios.

Because we will be working with so many species at such a fine scale, we will need to implement more efficient and fast methods.  Here we explore dplyr, data.table, and 

## Data

* Vulnerability scores from Butt et al. (2022). A trait-based framework for assessing the vulnerability of marine species to human impacts. Ecosphere, 13(2), e3919. https://doi.org/10.1002/ecs2.3919
* AquaX species distribution models from Reygondeau et al. (2026). AquaX: An enhanced and revised AquaMaps framework to model marine species distributions and biodiversity. PLOS ONE, 21(2), e0335823. https://doi.org/10.1371/journal.pone.0335823

## Methods

* Read in SDMs for multiple species as long data frame.
* Group by pixel and summarize stats.
* Save out as raster or parquet.

```{r}
library(tidyverse)
library(data.table)
library(here)
library(arrow)
library(e1071) ### for moments

source(here('common_fxns.R'))
```

### Identify test taxon

For prototyping, let's look at a subset of mammals, with ship strike as stressor.

```{r}
available_spp <- list.files(here_aquax('SDM')) %>% 
    str_extract('[0-9]+') %>% 
    as.numeric()
aquax_meta <- get_aquax_meta() %>%
    filter(aphia_id %in% available_spp)
spp_vuln_df <- get_spp_vuln() %>%
    filter(aphia_id %in% aquax_meta$aphia_id)
```

### Read in test spp

Load in the SDMs for this sample of species.  For nimble testing, let's take a small sample of cells that contain multiple species.

Let's define some versions of functions:

```{r}
get_sdm_duckdb <- function(aphia_id, 
                    scenario = c('Current', 
                                 'RCP26_2050', 'RCP26_2100',
                                 'RCP45_2050', 'RCP45_2100', 
                                 'RCP85_2050', 'RCP85_2100'),
                    xrange = NULL,
                    apply_thresh = TRUE) {

  ### define internal function:
  process_one_sdm <- function(id, scenario, apply_thresh) {
    filestem <- here_aquax('SDM/FINAL_EMSDM_EMMEAN_SP_%s.parquet')

    ### open connection, select/filter
    con <- duckdb::dbConnect(duckdb::duckdb())
    
    query_str <- paste0(
      "SELECT x, y, cutoff, ", paste(scenario, collapse = ', '),
      " FROM read_parquet('", sprintf(filestem, id), "')"
    )
    if(!is.null(xrange)) {
      ### tack on WHERE clause
      query_str <- paste0(
        query_str,
        " WHERE x <= ", max(xrange), " AND x >= ", min(xrange)
        )
    }
    df <- DBI::dbGetQuery(con, query_str) %>%
      janitor::clean_names() %>%
      mutate(aphia_id = id)
    
    ### close connection!
    duckdb::dbDisconnect(con, shutdown = TRUE)

    if(apply_thresh) {
      scen = tolower(scenario)
      df <- df %>%
        mutate(across(all_of(scen), ~ifelse(.x < cutoff, NA, 1))) %>%
        filter(!if_all(all_of(scen), is.na))
    }
    return(df)
  }
  ### if single ID, process directly; otherwise, parallel:
  if(length(aphia_id) == 1) {
    sdm_df <- process_one_sdm(id = aphia_id, 
                              scenario = scenario, 
                              apply_thresh = apply_thresh)
  } else {
    ### set # cores based on # ids, up to some max (20)
    n_cores <- min(length(aphia_id), 20)
    sdm_list <- parallel::mclapply(
      mc.cores = n_cores,
      X = aphia_id,
      FUN = process_one_sdm,
      scenario = scenario, apply_thresh = apply_thresh
    )
    sdm_df <- data.table::rbindlist(sdm_list)
  }
  return(sdm_df)
}

get_sdm_dplyr <- function(aphia_id, 
                          scenario = c('Current', 
                                       'RCP26_2050', 'RCP26_2100',
                                       'RCP45_2050', 'RCP45_2100', 
                                       'RCP85_2050', 'RCP85_2100'),
                          xrange = NULL,
                          apply_thresh = TRUE) {
  
  ### define internal function:
  process_one_sdm <- function(id, scenario, apply_thresh) {
    scen <- tolower(scenario)
    filestem <- here_aquax('SDM/FINAL_EMSDM_EMMEAN_SP_%s.parquet')
    df <- arrow::read_parquet(sprintf(filestem, id)) %>%
      janitor::clean_names() %>%
      select(x, y, all_of(scen), cutoff) %>%
      mutate(aphia_id = id)
    if(!is.null(xrange)) { 
      df <- df %>%
        filter(x <= max(xrange) & x >= min(xrange))
    }
    if(apply_thresh) {
      df <- df %>%
        mutate(across(all_of(scen), ~ifelse(.x < cutoff, NA, 1))) %>%
        filter(!if_all(all_of(scen), is.na))
    }
    return(df)
  }
  ### if single ID, process directly; otherwise, parallel:
  if(length(aphia_id) == 1) {
    sdm_df <- process_one_sdm(id = aphia_id, 
                              scen = scenario, 
                              apply_thresh = apply_thresh)
  } else {
    ### set # cores based on # ids, up to some max (20)
    n_cores <- min(length(aphia_id), 20)
    sdm_list <- parallel::mclapply(
      mc.cores = n_cores,
      X = aphia_id,
      FUN = process_one_sdm,
      scen = scenario, apply_thresh = apply_thresh
    )
    sdm_df <- data.table::rbindlist(sdm_list)
  }
  return(sdm_df)
}

```


```{r}
set.seed(42)
test_df <- spp_vuln_df %>%
    filter(taxon == 'marine_mammals' & stressor == 'wildlife_strike') %>%
    slice_sample(n = 10)

scenario <- c('current', 'rcp45_2050', 'rcp85_2050')

system.time({
    sdm_df1 <- get_sdm_dplyr(aphia_id = test_df$aphia_id, scenario = scenario)
})

system.time({
    sdm_df2 <- get_sdm_duckdb(aphia_id = test_df$aphia_id, 
                              scenario = scenario)
})

# system.time({
#   ### in data.table syntax since dplyr took forever
#   sdm_nspp_dt <- sdm_df1 %>%
#     as.data.table() %>%
#     .[ , .(n_spp = n_distinct(aphia_id)), by = .(x, y)]
# })
### Processed 14871578 groups out of 14871578. 100% done. Time elapsed: 221s. ETA: 0s.
###    user  system elapsed 
### 620.189  35.136 227.403 
# 
# sdm_sample_cells <- sdm_nspp_dt %>%
#   slice_max(order_by = n_spp, prop = 0.1)
# 
# write_csv(sdm_sample_cells, here('tmp/methods_test_sample_cells.csv'))
sdm_sample_cells <- read_csv(here('tmp/methods_test_sample_cells.csv'))

sdm_df <- sdm_df1 %>% inner_join(sdm_sample_cells, by = c('x', 'y'))

```

test with more species, including filtering by x columns:

```{r}
### all mammals:
test_df2 <- spp_vuln_df %>%
    filter(taxon == 'marine_mammals' & stressor == 'wildlife_strike')

set.seed(123)
xrange <- c(-180, -175)
system.time({
  for(i in 1:100) {
    sdm_df3 <- get_sdm_dplyr(aphia_id = test_df2$aphia_id, 
                             scenario = scenario,
                             xrange = xrange)
  }
})

system.time({
  for(i in 1:100) {
    sdm_df4 <- get_sdm_duckdb(aphia_id = test_df2$aphia_id, 
                              scenario = scenario,
                              xrange = xrange)
  }
})

```

After thresholding, presence = NA or 1; multiply presence in each scenario column by vulnerability for each species, then group_by/summarize per pixel.

### Test dplyr speed

```{r dplyr_version}
 system.time({
    spp_vuln_df <- sdm_df %>%
        left_join(test_df %>% select(aphia_id, vuln), by = 'aphia_id') %>%
        mutate(across(all_of(scenario), ~.x * vuln))
})

system.time({
    vuln_current_df <- spp_vuln_df %>%
        group_by(x, y) %>%
        summarize(nspp = n_distinct(aphia_id),
                  v_mean = mean(current, na.rm = TRUE),
                  v_sdev = sd(current, na.rm = TRUE),
                  v_skew = e1071::skewness(current, na.rm = TRUE),
                  v_kurt = e1071::kurtosis(current, na.rm = TRUE),
                  .groups = 'drop')
})
```

### Test data.table speed

The `data.table` methods are about twice as fast:

```{r data.table_version}
test_dt <- test_df %>% select(aphia_id, vuln) %>%
    as.data.table(key = 'aphia_id')
sdm_dt <- sdm_df %>%
    as.data.table(key = 'aphia_id') 

system.time({
    spp_vuln_dt <- test_dt[sdm_dt, on = "aphia_id"] %>%
        .[ , (scenario) := lapply(.SD, \(x) x * vuln), .SDcols = scenario]
}) 

### data.table results identical to dplyr results:
# asdf <- spp_vuln_df %>% 
#   as.data.frame() %>%
#   select(x, y, current, rcp45_2050, rcp85_2050, aphia_id) %>% 
#   arrange(aphia_id, x, y)
# qwer <- spp_vuln_dt %>% 
#   as.data.frame() %>%
#   select(x, y, current, rcp45_2050, rcp85_2050, aphia_id) %>% 
#   arrange(aphia_id, x, y)
# identical(asdf, qwer)
# [1] TRUE


system.time({ 
    vuln_current_dt <- spp_vuln_dt %>%
        .[ , .( nspp = n_distinct(aphia_id),
                v_mean = mean(current, na.rm = TRUE),
                v_sdev = sd(current, na.rm = TRUE),
                v_skew = e1071::skewness(current, na.rm = TRUE),
                v_kurt = e1071::kurtosis(current, na.rm = TRUE)
              ), by = .(x, y)]
})

### data.table results identical to dplyr results:
# asdf <- vuln_current_df %>%
#   as.data.frame() %>%
#   select(x, y, nspp, v_mean, v_sdev, v_skew, v_kurt) %>%
#   arrange(x, y)
# qwer <- vuln_current_dt %>%
#   as.data.frame() %>%
#   select(x, y, nspp, v_mean, v_sdev, v_skew, v_kurt) %>%
#   arrange(x, y)
# identical(asdf, qwer)
# [1] TRUE
```

### Test collapse speed

See https://fastverse.org/collapse/articles/collapse_for_tidyverse_users.html

This can calculate across all columns in 1/10th the time data.table takes to calculate one column.  Returns as a 3-d array:

1. the grouped variables (here, as x-y pairs like `"-179.975.-53.375"`)
2. the summary variables (`"N", "Mean", "SD", "Min", "Max", "Skew", "Kurt"`)
3. the summarized variables (`"current", "rcp45_2050", "rcp85_2050", "cutoff", "aphia_id", "n_spp", "vuln"`)

Does the `N` accurately count distinct aphia IDs?  There should be few occasions where `n_distinct(aphia_id)` is different than the number of observations.

```{r}
library(collapse)

system.time({
    spp_vuln_collapse <- sdm_df %>%
        collapse::join(test_df %>% select(aphia_id, vuln), how = 'left', on = 'aphia_id') %>%
        collapse::fmutate(across(all_of(scenario), .fns = \(x) x * .$vuln))
})
# identical(spp_vuln_collapse, spp_vuln_df)
# [1] TRUE
system.time({
    vuln_current_collapse <- spp_vuln_collapse %>%
        fgroup_by(x, y) %>%
        qsu(higher = TRUE)
})
### NOTE: returns as 3-d array: rows = groups, cols = stats, layers = ungrouped vars (including scenarios)
### NOTE ALSO: this does all the desired columns at once, rather than one at a time like previous attempts

### try a more dplyr format:
system.time({
    vuln_current_collapse <- spp_vuln_collapse %>%
        fgroup_by(x, y) %>%
        fsummarize(nspp   = fndistinct(aphia_id),
                   v_mean = fmean(current, na.rm = TRUE),
                   v_sdev = fsd(current, na.rm = TRUE),
                   # v_skew = qsu(current, na.rm = TRUE)$Skew,
                   # v_kurt = qsu(current, na.rm = TRUE)$Kurt,
                   keep.group_vars = FALSE)
})
### NOTE: fast for mean and sd, but skew and kurtosis not so much
```

### Reading in: piece by piece?

Can we read in each SDM in pieces, e.g., just observations where `x` is a specific value or range?  If we can do this quickly and efficiently, we can summarize more species in smaller chunks and save out.  It doesn't look like `arrow::read_parquet` can do this...  but perhaps read in and filter with `data.table`:  `dt[x == -179.975]`:

```{r}
test_id <- test_df$aphia_id[1]
scenario = c('Current', 'RCP26_2050', 'RCP26_2100',
             'RCP45_2050', 'RCP45_2100', 'RCP85_2050', 'RCP85_2100') %>%
  tolower()
filestem <- here_aquax('SDM/FINAL_EMSDM_EMMEAN_SP_%s.parquet')

### check timing of read/clean/select using dplyr
system.time({
  x <- lapply(1:100, FUN = function(x, id, scen, fstem) {
    arrow::read_parquet(sprintf(fstem, id)) %>%
      janitor::clean_names() %>%
      select(x, y, all_of(scen), cutoff)
  }, id = test_id, scen = scenario, fstem = filestem)
})

### check timing of read/clean/select using data.table
system.time({
  y <- lapply(1:100, FUN = function(x, id, scen, fstem) {
    arrow::read_parquet(sprintf(fstem, id)) %>%
      janitor::clean_names() %>%
      setDT() %>% ### copy in place 
      .[ , c('x', 'y', scen, 'cutoff'), with = FALSE]
  }, id = test_id, scen = scenario, fstem = filestem)
})
```


```{r}
### Using dplyr::select
system.time({
  x <- lapply(1:100, FUN = function(i, id, scen, fstem, x) {
    df <- arrow::read_parquet(sprintf(fstem, id)) %>%
      janitor::clean_names() %>%
      select(x, y, all_of(scen), cutoff)
    df <- df %>%
      mutate(across(all_of(scen), ~ifelse(.x < cutoff, NA, 1))) %>%
      filter(!if_all(all_of(scen), is.na)) %>%
      mutate(aphia_id = id)
    return(df)
  }, id = test_id, scen = scenario, fstem = filestem)
})

### Using data.table methods but dplyr::select
system.time({
  y <- lapply(1:100, FUN = function(i, id, scen, fstem) {
    dt <- arrow::read_parquet(sprintf(fstem, id)) %>%
      janitor::clean_names() %>%
      select(x, y, all_of(scen), cutoff) %>%
      setDT()
    dt[, (scen) := lapply(.SD, \(x) fifelse(x < cutoff, NA_real_, 1)), .SDcols = scen]
    ### filter to keep only rows where at least one scenario is non-NA
    keep <- rowSums(is.na(dt[, ..scen])) < length(scen)
    dt <- dt[keep]
    
    dt <- dt[, aphia_id := id]
    
    return(dt)
  }, id = test_id, scen = scenario, fstem = filestem)
})

identical(
  x[[1]] %>% as.data.frame(),
  y[[1]] %>% as.data.frame()
)

# head(x[[1]]); head(y[[1]])
```

```{r}
### Using dplyr::select; add in x filter
xvec <- c(-179.875, -179.825)
system.time({
  x <- lapply(1:100, FUN = function(i, id, scen, fstem, xvec) {
    df <- arrow::read_parquet(sprintf(fstem, id)) %>%
      janitor::clean_names() %>%
      select(x, y, all_of(scen), cutoff)
    df <- df %>%
      filter(x %in% xvec) %>%
      mutate(across(all_of(scen), ~ifelse(.x < cutoff, NA, 1))) %>%
      filter(!if_all(all_of(scen), is.na)) %>%
      mutate(aphia_id = id)
    return(df)
  }, id = test_id, scen = scenario, fstem = filestem, xvec = xvec)
})

### Using data.table methods but dplyr::select
system.time({
  y <- lapply(1:100, FUN = function(i, id, scen, fstem, xvec) {
    dt <- arrow::read_parquet(sprintf(fstem, id)) %>%
      janitor::clean_names() %>%
      select(x, y, all_of(scen), cutoff) %>%
      setDT()
    dt <- dt[x %in% xvec]
    dt[ , (scen) := lapply(.SD, \(x) fifelse(x < cutoff, NA_real_, 1)), .SDcols = scen]
    ### filter to keep only rows where at least one scenario is non-NA:
    keep <- rowSums(is.na(dt[, ..scen])) < length(scen)
    dt <- dt[keep]
    dt <- dt[, aphia_id := id]
    return(dt)
  }, id = test_id, scen = scenario, fstem = filestem, xvec = xvec)
})

identical(
  x[[1]] %>% as.data.frame(),
  y[[1]] %>% as.data.frame()
)
# head(x[[1]]); head(y[[1]])
```


## Takeaways

For reading in and filtering files, it looks like dplyr does pretty well, especially if we consider that data.table parallelization may not provide as much of a boost when included in mclapply...

For processing large datasets quickly, collapse::qsu looks amazing, but will require some finesse to get it back into data frame formats (and drop the unnecessary values).