---
title: Test tuna atlas dataset against AquaX
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

Juliette discovered that the AquaX dataset predicts no Bigeye Tuna (_thunnus obesus_) in a location where a high density of Bigeye Tuna is harvested according to the Tuna Atlas.  Let's see if this is a systematic problem!

## Data

* FAO. 2026. FIRMS Global Tuna Atlas (GTA). In: Fisheries and Aquaculture [online]. [Cited Tuesday, September 15th 2026].
http://www.fao.org/fishery/collection/firms-tuna-atlas/en
    * Data: https://zenodo.org/records/21114897 Tuna catch dataset at 5 degree resolution
        * https://zenodo.org/records/21114897/files/global_catch_5deg_1m_firms_level0_harmonized.parquet?download=1
    * Metadata: https://www.fao.org/fishery/collection/asfis/en
        * https://www.fao.org/fishery/static/ASFIS/ASFIS_sp.zip
* AquaX

## Methods

```{r}
library(here)
library(tidyverse)
library(collapse)
library(arrow)
library(duckdb)
library(terra)

source(here('common_fxns.R'))
```

### Download tuna atlas data

```{r}
src <- 'https://zenodo.org/records/21114897/files/global_catch_5deg_1m_firms_level0_harmonized.parquet?download=1'
dst <- here('_data/fao_2026/tuna/tuna_atlas_5deg.parquet')
if(!file.exists(dst)) {
    if(!dir.exists(dst)) dir.create(dirname(dst))
    download.file(src, dst, mode = 'wb')
}

tuna_df <- read_parquet(dst) |> 
    janitor::clean_names()
```

### Download tuna atlas metadata

```{r}
src <- 'https://www.fao.org/fishery/static/ASFIS/ASFIS_sp.zip'
dst <- here('_data/fao_2026/tuna/ASFIS_sp_2026.1.csv')
if(!file.exists(dst)) {
    if(!dir.exists(dst)) dir.create(dirname(dst))
    zip_tmp <- here('tmp/tuna_meta.zip')
    download.file(src, zip_tmp, mode = 'wb')
    unzip(zipfile = zip_tmp, exdir = dirname(dst))
    unlink(zip_tmp)
}

gta_meta_df <- read_csv(dst) |> 
    janitor::clean_names()
```

### Convert geographic identifier to lat-long

The `geographic_identifier` field: The Tuna Atlas (and most tRFMO tuna datasets) use the CWP (Coordinating Working Party on Fishery Statistics) grid code with **S**ize, **Q**uadrant, **LA**titude, and **LON**gitude: **S|Q|LA|LON** (# of letters = # of digits)

* Size digit = 6 indicates 5°×5°
* Quadrant — which hemisphere combo the cell's corner sits in:
    *  1:	NE (lat ≥0, lon ≥0)
    *  2:	SE (lat <0, lon ≥0)
    *  3:	SW (lat <0, lon <0)
    *  4:	NW (lat ≥0, lon <0)
* Latitude: Lat digits (next 2) — absolute latitude of the cell's corner nearest 0°N/0°E (i.e., the corner touching the Equator/Greenwich reference point), in whole degrees.
* Lon digits (last 3) — same idea for longitude.

UGHHH but we want cell CENTERS not CORNERS.  For positive, add 2.5, for negative, subtract 2.5.

Also join with metadata to get scientific names.

```{r}
tuna_xy_df <- tuna_df |>
    fsubset(time_start >= 2020) |>
    separate_wider_position(geographic_identifier, 
        widths = c(s = 1, q = 1, y = 2, x = 3)) |> 
    fmutate(s = as.integer(s), q = as.integer(q), y = as.integer(y), x = as.integer(x)) |> 
    fmutate(x = ifelse(q >= 3, -x - 2.5, x + 2.5),
           y = ifelse(q %in% 2:3, -y - 2.5, y + 2.5)) |> 
    join(gta_meta_df, on = c('species' = 'alpha3_code'), how = 'left')

# bigeye_r <- tuna_xy_df |> 
#     fsubset(scientific_name == 'Thunnus obesus') |> 
#     fgroup_by(x, y) |> 
#     fsummarize(total = sum(measurement_value)) |> 
#     rast()
# plot(bigeye_r)

# tuna_xy_df$scientific_name %>% n_distinct()
```

### Check species against AquaX HSI

```{r}
spp_vec <- tuna_xy_df$scientific_name |> unique() |> sort()
aquax_gta_meta <- get_aquax_meta() |> 
    inner_join(gta_meta_df, by = c('scientific_name')) |> 
    filter(scientific_name %in% spp_vec)
### 52 of 67 total spp in the tuna dataset
missing <- spp_vec[!spp_vec %in% aquax_gta_meta$scientific_name]
### not at spp level - disregard!

spp_to_run <- spp_vec[!spp_vec %in% missing]
```

* Filter tuna atlas catch map and summarize to each cell
* Load AquaX distribution and aggregate to 5 degree blocks, averaging HSI
* Linear model?  look for poor correlation between tuna catch and HSI
    * HSI should correlate well with OBIS/GBIF observations
    * If tuna catch and HSI poorly correlated, indicates maybe tuna catch is poorly correlated with OBIS/GBIF observations?
* Or: Look at areas with low values of HSI and see if there are high values for fishing
    * Quadrants - 


```{r set up function}
tuna_vs_hsi <- function(spp) {
    ### spp <- 'thunnus obesus'
    spp <- tolower(spp)

    ### generate sdm at 5 deg resolution, averaging HSI values
    spp_id_aquax <- aquax_gta_meta |> 
        filter(tolower(scientific_name) == spp) |> 
        pull(aphia_id)
    spp_sdm_aquax_df <- get_sdm(spp_id_aquax, scenario = 'current', 
            apply_thresh = FALSE)
    cutoff <- spp_sdm_aquax_df$cutoff |> unique()
        
    spp_sdm_aquax_r <- spp_sdm_aquax_df |> 
        cell_id_to_xy()  |> 
        select(x, y, current) |> 
        rast(crs = 'epsg:4326') |> 
        extend(ext(-180, 180, -90, 90))

    sdm_agg_r <- aggregate(spp_sdm_aquax_r, fact = 100, fun = 'mean', na.rm = TRUE)
    n_sdm  <- sum(!is.na(values(sdm_agg_r)))

    ### Generate catch map summarizing across 2020-present
    spp_gta_df <- tuna_xy_df |> 
        ### put tolower() in mutate instead of filter
        mutate(scientific_name = tolower(scientific_name)) |> 
        filter(scientific_name == spp) |> 
        select(x, y, measurement_value) |> 
        fgroup_by(x, y) |> 
        fsummarize(total = sum(measurement_value)) |> 
        ungroup()
        
    ### use sdm raster as template
    spp_gta_r <- rasterize(vect(spp_gta_df, geom = c('x', 'y')),
            y = sdm_agg_r, field = 'total') |> 
        setNames('total')

    ### combine rasters and go from there
    df <- c(spp_gta_r, sdm_agg_r) |> 
        as.data.frame(xy = TRUE) |> 
        mutate(current = ifelse(is.na(current), 0, current),
               spp = spp,
               cutoff = cutoff,
               n_aquax = n_sdm)
    # ggplot(df, aes(x = current, y = log(total))) + geom_point()

    return(df)
}
```

For each spp, summarize total catch, and look at HSI values in cells with catch

```{r}
test_outf <- here('tmp/tuna_test.csv')
if(!file.exists(test_outf)) {
    # tuna_test_list <- lapply(X = spp_to_run, FUN = tuna_vs_hsi)
    tuna_test_list <- vector('list', length = length(spp_to_run))
    for(i in seq_along(spp_to_run)) {
        spp <- spp_to_run[i]
        tuna_test_list[[i]] <- tuna_vs_hsi(spp)
    }
    tuna_test_df <- rowbind(tuna_test_list)
    write_csv(tuna_test_df, test_outf)
}
tuna_test_df <- read_csv(test_outf) |> 
    mutate(presence = current >= cutoff)

high_catch <- mean(log(tuna_test_df$total), na.rm = TRUE)

ggplot(tuna_test_df, aes(x = current, y = log(total), color = presence)) + 
    geom_point() +
    geom_hline(yintercept = high_catch)
```

### Identify problematic species

Problematic spp might be those with a high proportion of high-catch cells in non-suitable locations.

```{r}
problems_df <- tuna_test_df |> 
    filter(!is.na(total)) |> 
    group_by(spp) |> 
    mutate(high = log(total) > mean(log(total))) |> 
    ungroup()

problem_summary_df <- problems_df |> 
    group_by(spp, n_aquax) |> 
    summarize(n_catch_cells = n(),
              n_problem = sum(high & !presence),
              pct_prob = n_problem / n_catch_cells,
              pct_fished = n_catch_cells / first(n_aquax)) |> 
    ungroup() |> 
    mutate(add_label = pct_fished > .5 & n_catch_cells > 600 |
                       pct_prob > .35)

ggplot(problem_summary_df, aes(x = pct_prob, y = pct_fished)) +
    geom_point(aes(size = n_catch_cells)) +
    ggrepel::geom_text_repel(data = problem_summary_df %>% filter(add_label),
              aes(label = spp), hjust = 0, nudge_x = .02)
```

### Plot AquaX HSI vs fishing for select spp

For large-ranged species with fishing making up a substantial portion of their range, plot the AquaX distribution and the GTA distribution.  Use the 5 degree aggregated data.

```{r}
#| fig-width: 6
#| fig-height: 8

high_catch_spp <- tuna_test_df |> 
    group_by(spp) |> 
    summarize(total = sum(total, na.rm = TRUE)) |> 
    filter(total > 1e6)
compare_spp <- problem_summary_df |> 
    filter(add_label | str_detect(spp, 'thunnus') | spp %in% high_catch_spp$spp) |> 
    pull(spp)

for(s in compare_spp) {
    # s <- compare_spp[1]
    spp_map <- tuna_test_df |> 
        filter(spp == s) |> 
        mutate(current = ifelse(current == 0, NA, current),
               high_catch = log(total) > mean(log(total), na.rm = TRUE),
               ### let's lower the "presence" threshold for the large cell
               presence = current >= cutoff * .7)
    aquax_plot <- ggplot(spp_map, aes(x, y, fill = current)) +
        geom_raster(show.legend = FALSE) +
        scale_fill_viridis_c() +
        coord_fixed() +
        theme_void() +
        labs(title = paste('AquaX (aggr to 5 deg):', s))
    gta_plot <- ggplot(spp_map, aes(x, y, fill = log(total))) +
        geom_raster(show.legend = FALSE) +
        scale_fill_viridis_c() +
        coord_fixed() +
        theme_void() +
        labs(title = paste('GTA (5 deg):', s))
    problem_plot <- ggplot() +
        geom_raster(data = spp_map, aes(x, y), fill = 'grey80') +
        geom_raster(data = spp_map %>% filter(high_catch & !presence), 
                    aes(x, y), fill = 'red') +
        coord_fixed() +
        theme_void() +
        labs(title = paste('Problem areas:', s))
    tri_plot <- cowplot::plot_grid(aquax_plot, gta_plot, problem_plot, ncol = 1)

    print(tri_plot)
}
```
