---
title: Map impacts - fisheries
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

Vulnerability and Stressor Intensity calculated in previous scripts.  **Impact**:  Once we have stressor intensity layers for each subgroup, we then apply the vulnerability scores for each of those layers to calc impact for each size/location group
  
  * We could keep each layer as a separate stressor, or take a biomass density-weighted mean for the overall impact

## Data

* Vulnerability maps generated in `map_vuln_fisheries` script
    * files in `_output/vuln_maps/biomass_removal'
    * filenames: `biomass_removal_vuln_<class>_gapfilled.tif`
* Stressor intensity maps generated in `map_stressors_fisheries` script
    * files in '_output/stressor_maps/biomass_removal'
    * `biomass_removal_biomass_by_class.tif`
    * `biomass_removal_stressor_intensity.tif`
* All these rasters are in Behrmann projection at 5 km resolution


## Methods

```{r attach packages}
#| message: false
#| 
library(tidyverse)
library(here)
library(arrow)
library(collapse)
library(here)
library(terra)

source(here('common_fxns.R'))
```

### Load vulnerability layers by size and position

The vulnerability maps have been generated and gapfilled in a previous script.

```{r}
vuln_r_fs <- list.files(here('_output/vuln_maps/biomass_removal'), 
                        pattern = '.tif$', full.names = TRUE)
vuln_names_df <- data.frame(f = vuln_r_fs) |>
  mutate(name = str_remove_all(basename(f), '.+_vuln_|_[a-z]+_gapfilled.tif'),
         name = str_replace(name, "lg", "90cm"),
         name = str_replace(name, "md", "30to90cm"),
         name = str_replace(name, "sm", "30cm"),
         name = ifelse(str_detect(f, 'pelagic'), paste0('bp', name), paste0('bd', name)))
vuln_r <- rast(vuln_r_fs) %>%
  setNames(vuln_names_df$name)
plot(vuln_r)
```

### Load stressor layers and density layers

Stressor layers are rescaled stressor intensity; density layers are used to calculate a weighted average of (str * vuln) for total impacts.

```{r}
str_f <- list.files(here('_output/stressor_maps/biomass_removal'), 
                    pattern = 'stressor_intensity.tif', full.names = TRUE)
str_r <- rast(str_f)
names(str_r) <- str_remove(names(str_r), 'str_intensity_')
plot(str_r)
```

```{r}
biomass_f <- list.files(here('_output/stressor_maps/biomass_removal'), 
                        pattern = 'biomass_by_class.tif', full.names = TRUE)
biomass_r <- rast(biomass_f)
plot(biomass_r)
```

Do the names align for vuln, biomass and stressor?

```{r}
all(names(biomass_r) == names(str_r))
all(names(vuln_r) == names(str_r))
```

### Calculate impact

Impact is just the product of vulnerability and intensity

```{r}
impact_r <- vuln_r * str_r
dem_vec <- str_detect(names(impact_r), 'bd'); pel_vec <- !dem_vec

biomass_dem <- biomass_r[[dem_vec]]
impact_dem_all <- impact_r[[dem_vec]]
impact_dem_sum <- sum(impact_dem_all * biomass_dem, na.rm = TRUE) / sum(biomass_dem, na.rm = TRUE)

biomass_pel <- biomass_r[[pel_vec]]
impact_pel_all <- impact_r[[pel_vec]]
impact_pel_sum <- sum(impact_pel_all * biomass_pel, na.rm = TRUE) / sum(biomass_pel, na.rm = TRUE)

impact_all_sum <- sum(impact_r * biomass_r, na.rm = TRUE) / sum(biomass_r, na.rm = TRUE)

impact_r <- c(impact_dem_sum, impact_pel_sum, impact_all_sum) %>%
  setNames(c('impact_demersal', 'impact_pelagic', 'impact_total'))
plot(impact_r)

writeRaster(impact_r, here('_output/biomass_removal_impact.tif'), overwrite = TRUE)
```