---
title: Map vulnerability - fisheries
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary


Using FAO info, identify currently-fished species - and then use the distributions and "biomass removal" vulnerability scores for these species to find average vulnerability in the different size classes and water column positions (we have trait values for these for most spp)

* i.e., we'll have fisheries vulnerability maps for 6 groups: sm/med/lg pelagics and sm/med/lg demersals.
* Alternately, instead of using FAO info to identify currently-fished species, just throw all species into the mix - i.e., all small pelagics drive the vuln scores, instead of just the currently-fished small pelagics.


## Data

FAO capture data: https://www.fao.org/fishery/collection/capture/en and https://www.fao.org/fishery/collection/aquaculture/en.  Here we are only interested in which species exist in the

* Taxonomic information stored in `_data/fao_2026/capture/CL_FI_SPECIES_GROUPS.csv`
* Catch information stored in `_data/fao_2026/capture/Capture_Quantity.csv`
* See `data_access_fisheries` script for download.

## Methods

* Read in SDMs for multiple species as long data frame
* Group by pixel and summarize stats.
* Save out as parquet.

```{r attach packages}
#| message: false
#| 
library(tidyverse)
library(here)
library(arrow)
library(collapse)
library(here)
library(terra)

source(here('common_fxns.R'))
```

### Identify fisheries classes

ISIMIP models break fishery species into size classes (<30 cm, 30-90 cm, >90 cm) and pelagic/demersal.  

* Identify all fished species from FAO, and for each, classify into these different classes.
* For each class, generate a biomass removal vulnerability map.  Should this be weighted by FAO species catch?

```{r define FAO helper functions}
get_clean_fao_spp <- function() {
    fao_spp_gps <- read_csv(here('_data/fao_2026/capture/CL_FI_SPECIES_GROUPS.csv'),
                          show_col_types = FALSE) |>
      janitor::clean_names() |>
      rename(sciname = scientific_name, comname = name_en) |>
      mutate(sciname2 = str_split(sciname, ', ?')) |>
      unnest(sciname2) |>
      mutate(sciname2 = str_replace(sciname2, '^[A-Z]\\. ?', 
                                    paste0(str_extract(sciname, '^[A-z]+'), ' ') )) |>
      mutate(nei = str_detect(sciname, ' spp$') | 
              !is.na(comname) & str_detect(comname, 'NEI') |
              !str_detect(sciname, ' ')) |> 
      select(sciname = sciname2, comname, species_alpha_3_code = x3a_code, nei) |>
      clean_fao_aphia_id()
                        
    ### otherwise: return it all
    return(fao_spp_gps)
}

clean_fao_aphia_id <- function(df) {
  fao_spp_aphia_f <- here('_data/fao_2026/fao_spp_aphiaid.csv')
  if(!file.exists(fao_spp_aphia_f)) {
    library(taxize)
    df_chunks <- df |>
      filter(!nei) |>
      mutate(chunk = rep(1:100, length.out = n()))
    tmp_fstem <- here('tmp/tmp_fao_aphia_%03d.csv')
    for(i in 1:100) {
      ### i <- 1
      message(i)
      tmp_f <- sprintf(tmp_fstem, i)
      if(file.exists(tmp_f)) next()
      tmp_df <- df_chunks |>
        filter(chunk == i) |>
        mutate(aphia_id = get_wormsid(sciname, marine_only = FALSE, 
                                      accepted = TRUE, ask = FALSE)) |>
        select(sciname, aphia_id) |>
        distinct()
      write_csv(tmp_df, tmp_f)
    }
    tmp_fs <- list.files('tmp', pattern =  'tmp_fao_aphia_.+.csv', full.names = TRUE)
    fao_spp_aphia_df <- lapply(
      X = tmp_fs,
      FUN = read_csv, show_col_types = FALSE) |>
    bind_rows()
    write_csv(fao_spp_aphia_df |> mutate(aphia_id = as.integer(aphia_id)), fao_spp_aphia_f)
    unlink(tmp_fs)
  }
  df <- df |>
    left_join(read_csv(fao_spp_aphia_f, show_col_types = FALSE),
              by = 'sciname', relationship = 'many-to-many')
}

clean_fao_value <- function(df) {

  df <- df |>
    janitor::clean_names() |>
    group_by(species_alpha_3_code, period) |>
    ### summarize by species ID first
    summarize(total = sum(value, na.rm = TRUE), .groups = 'drop') |>
    left_join(get_clean_fao_spp(), by = 'species_alpha_3_code',
              relationship = 'many-to-many') |>
    ### split multi-spp IDs and reallocate catch
    group_by(species_alpha_3_code, period) |>
    mutate(total = total / n()) |>
    ### then group by spp again (since some spp are both solo and in multi spp)
    group_by(sciname, aphia_id, period, nei) |>
    summarize(total = sum(total), .groups = 'drop')

}
```

Let's use genus-level species catches to supplement the species-level catches (adds about 2600 spp: 3160 --> 5777).  This will include a few more species within the vulnerability scores.  Using family-level adds more than 7100 species, more doubling the species count - mostly small and medium demersal.  But there is probably a lot of uncertainty as to which of these spp are actually fished.

```{r load fao data}

fao_capture_raw <- read_csv(here('_data/fao_2026/capture/Capture_Quantity.csv'),
                            show_col_types = FALSE) |>
  clean_fao_value()

aquax_taxa <- get_aquax_meta() |>
  select(aphia_id, sciname = scientific_name, kingdom, phylum, class, order, family, genus)

fao_species <- fao_capture_raw |> filter(!nei) |> 
  mutate(level = 'species')

fao_nei <- fao_capture_raw |> filter(nei) |>
  filter(!str_detect(sciname, ' spp$') & !str_detect(sciname, 'idae')) |>
  mutate(level = 'nei')

fao_genus <- fao_capture_raw |> filter(nei) |>
  filter(str_detect(sciname, ' spp$')) |>
  mutate(genus = str_remove(sciname, ' .+'), nei = FALSE) |>
  select(-aphia_id, -sciname) |>
  left_join(aquax_taxa, by = 'genus', relationship = 'many-to-many') |>
  filter(!is.na(aphia_id)) |>
  ### divide genus-level total evenly across all species
  group_by(genus, period) |>
  mutate(total = total / n()) |>
  ungroup() |>
  select(-genus) |>
  mutate(level = 'genus')

fao_family <- fao_capture_raw |> filter(nei) |>
  filter(str_detect(sciname, 'idae$')) |>
  rename(family = sciname) |>
  select(-aphia_id) |>
  left_join(aquax_taxa, by = 'family', relationship = 'many-to-many') |>
  filter(!is.na(aphia_id)) |>
  ### divide family-level total evenly across all species
  group_by(family, period) |>
  mutate(total = total / n()) |>
  ungroup() |>
  select(-family) |>
  mutate(level = 'family')

fao_capture <- bind_rows(fao_species, fao_genus, fao_family, fao_nei) |>
  select(aphia_id, sciname, level, period, total) |>
  distinct()

write_csv(fao_capture, here('_data/fao_2026/fao_capture_spp_genus_fam.csv'))

### What proportion are not differentiated at spp level (e.g., NEI)?
fao_cap_sum <- fao_capture |>
  group_by(period, level) |>
  summarize(total = sum(total), .groups = 'drop') |>
  mutate(level = factor(level, levels = c('species', 'genus', 'family', 'nei')))

ggplot(fao_cap_sum, aes(x = period, y = total)) +
  geom_area(aes(fill = level))

fao_capture_spp <- fao_capture |>
  filter(period > 2000 & !is.na(aphia_id)) |>
  filter(!level %in% c('nei', 'family')) |>
  group_by(sciname, aphia_id, period) |>
  summarize(total = sum(total), .groups = 'drop') |>
  group_by(sciname, aphia_id) |>
  summarize(mean_catch = mean(total, na.rm = TRUE), .groups = 'drop')
```

Which of these species are mapped in AquaX?  Match to Aphia ID.  Then bind traits for body size and water column position, and group species into the size/position classes.

```{r}
aquax_meta <- get_aquax_meta()

size_df  <- read_csv(here('_data/ohara_2024/funct_entity_traits/fe_spp_traits.csv'),
                     show_col_types = FALSE) |>
  filter(!is.na(length_cm)) |>
  mutate(size = case_when(length_cm < 30 ~ 'sm',
                          length_cm < 90 ~ 'md',
                          !is.na(length_cm) ~ 'lg',
                          TRUE ~ 'oops')) |> 
  select(aphia_id, size) |>
  distinct()

fao_aquax <- fao_capture_spp |> 
  inner_join(aquax_meta, by = c('aphia_id', 'sciname' = 'scientific_name')) |>
  mutate(position = ifelse(pelagic, 'pelagic', 'demersal')) |>
  select(aphia_id, sciname, mean_catch, position) |>
  inner_join(size_df, by = 'aphia_id') |>
  mutate(class = paste(size, position, sep = '_'))

### Some moderately-fished spp without size info - update FishBase data?
# check_nas <- fao_aquax |> 
#   filter(is.na(size)) |>
#   filter(mean_catch > 100)

biomass_removal_vuln_df <- get_spp_vuln() |>
  filter(stressor == 'biomass_removal') |>
  select(aphia_id, vuln)

fao_aquax_vuln_df <- fao_aquax |> 
  filter(!is.na(size)) |>
  left_join(biomass_removal_vuln_df, by = 'aphia_id')

# fao_aquax_vuln_df$class |> table()
# # lg_demersal  lg_pelagic md_demersal  md_pelagic sm_demersal  sm_pelagic 
# #         477         168        1630         230        1882         287 
```

### Iterate over classes to develop vulnerability maps

```{r define vulnerability helper functions}

process_slice <- function(sdm_df, vuln_df, scenario, digits = NULL) {
  ### if no observations in this slice, move along
  if(nrow(sdm_df) == 0) return(data.frame())
  
  ### join with vuln scores - use %>% to support .$ call
  sdm_spp_vuln <- sdm_df %>%
    collapse::join(vuln_df |> select(aphia_id, vuln), 
                   how = 'left', on = 'aphia_id', verbose = FALSE) %>%
    collapse::fmutate(across(.cols = all_of(scenario), .fns = \(x) x * .$vuln))
  
  ### calc summary stats including higher moments
  sdm_tx_vuln_array <- sdm_spp_vuln |>
    fgroup_by(cell_id) |>
    qsu(higher = FALSE) ### drop skew and kurtosis

  ### convert qsu output into df for each scenario, then join all scenarios
  sdm_tx_vuln_df <- lapply(X = scenario, 
                           FUN = pull_scenario_stats, 
                           d_array = sdm_tx_vuln_array) |>
    ### purrr::reduce to apply full join programmatically across all scenarios
    purrr::reduce(.f = full_join, by = c('cell_id', 'stat'))
  if(!is.null(digits)) {
    sdm_tx_vuln_df <- sdm_tx_vuln_df |>
      ### collapse::fmutate doesn't understand this syntax alas
      collapse::fmutate(across(.cols = all_of(scenario), .fns = round, digits = digits))
  }
  return(sdm_tx_vuln_df)
}

pull_scenario_stats <- function(sc, d_array) {
  df <- d_array[ , , sc] |>
    as.data.frame() |>
    janitor::clean_names() |>
    ### the qsu seems to fill in zero where no observations...
    ### drop those obs here, let the purrr::reduce fill in values as NA
    collapse::fsubset(n > 0) |>
    ### drop min and max values... keep mean, sd, skew, kurtosis, n
    select(-min, -max) |>
    ### qsu creates a group column representing cell_id
    rename(cell_id = variable) |>
    pivot_longer(cols = -cell_id, names_to = 'stat', values_to = sc) |>
    collapse::fmutate(cell_id = as.integer(as.character(cell_id)))

  return(df)
}
```

```{r iterate over size classes}
size_classes <- fao_aquax_vuln_df$class |> unique()

for(size_class in size_classes) {
  ### size_class <- 'lg_demersal'

  out_fstem <- here('_output/vuln_maps/biomass_removal/biomass_removal_vuln_%s.parquet') 
  if(!dir.exists(dirname(out_fstem))) dir.create(dirname(out_fstem))
   
  out_f <- sprintf(out_fstem, size_class)
  
  if(file.exists(out_f)) next()
    
  message(paste0('Creating ', basename(out_f), '\n'))

  class_vuln_df <- fao_aquax_vuln_df |>
      fsubset(class == size_class)
  
  #####################################
  ### iterate slices for size class ###
  #####################################
  
  tmp_out_fstem <- here('tmp/tx_vuln/tmp_biomass_removal_vuln_%s_%s.parquet')
  if(!dir.exists(dirname(tmp_out_fstem))) dir.create(dirname(tmp_out_fstem))
  
  ### set the number of row slices by cell_id: because cell_id scans rowwise,
  ### row 1 is 1-7200, 2 is 7201-14400, etc.
  yslice_w <- 5                ### slice width in degrees
  yslice_n <- 180 / yslice_w   ### number of vertical slices

  slice_vec <- 1:yslice_n

  for(i in slice_vec) {
    ### i <- 2
    ### generate cellrange for this slice of y rows
    cellrange <- calc_cellrange(i, yslice_w)
    
    ### Check for existing temp files - if all exist, skip to next slice
    slice_tmp_out_f <- sprintf(tmp_out_fstem, size_class, paste0(cellrange, collapse = '_'))
    if(file.exists(slice_tmp_out_f)) next()

    #########################################
    ### get sdms for this slice and class ###
    #########################################
    scenario = c('current', 'rcp45_2050', 'rcp85_2050')
    sdm_df <- get_sdm(aphia_id = class_vuln_df$aphia_id |> unique(), 
                      scenario = scenario,
                      cellrange = cellrange,
                      apply_thresh = TRUE)
    
    # message('Calculating ', basename(tmp_out_f))
    sdm_vuln_df <- process_slice(sdm_df = sdm_df,
                                 vuln_df = class_vuln_df,
                                 scenario = scenario,
                                 digits = 4)
    ### write out to tmp file
    arrow::write_parquet(sdm_vuln_df, slice_tmp_out_f)
    
  } ### END ITERATION OVER SLICES

  ##########################################
  ###   assemble slices for size class   ###
  ##########################################

  tmp_fs <- list.files(dirname(tmp_out_fstem), 
                       pattern = sprintf('biomass_removal_vuln_%s', size_class), 
                       full.names = TRUE)
  ### drop any tiny files (499 B is an empty file)
  tmp_fs_clean <- tmp_fs[file.size(tmp_fs) > 500]
  
  ###########################################
  ### gather all maps for this size class ###
  ###########################################
  complete_sdm_vuln_df <- arrow::open_dataset(tmp_fs_clean, format = 'parquet') |>
    collect() 

  ####################################################
  ### save nspp separately ONCE if not yet created ###
  ####################################################
  class_spp_rich_f <- sprintf(here('_output/vuln_maps/biomass_removal/spp_richness_by_size_class',
                                   'spp_richness_%s.parquet'), size_class)
  if(!dir.exists(dirname(class_spp_rich_f))) dir.create(dirname(class_spp_rich_f))

  if(!file.exists(class_spp_rich_f)) {
    spp_richness <- complete_sdm_vuln_df |>
      collapse::fsubset(stat == 'n') |>
      select(-stat) |>
      fmutate(across(-cell_id, as.integer))
    arrow::write_parquet(spp_richness, class_spp_rich_f)
  }
  
  #################################################
  ### drop nspp stat and save out for str/taxon ###
  #################################################
  complete_drop_n <- complete_sdm_vuln_df |>
    collapse::fsubset(stat != 'n')
  
  arrow::write_parquet(complete_drop_n, out_f)
  
  unlink(tmp_fs)  ### clean up temp directory
  
} ### END LOOP over size classes
```

### Gapfill and project

The vulnerability layers, based on currently-fished species, have large gaps in some cases.  If these align with zero-value stressors, great, otherwise some gapfilling will need to occur.

* Read in parquet files and convert to rasters
* Gapfill at the 5 km resolution

```{r define gapfill function}
rast_gapfill <- function(r, w = 3, iter = 5) {
  if(iter == 1) {
    r_focal <- focal(r, w = 3, fun = "mean", na.policy = "only", na.rm = TRUE)

    r_filled <- cover(r, r_focal)
    ### cover() returns r where r is non-NA, and r_focal elsewhere
    return(r_filled)
  }

  r_filled <- r
  if(iter >= 5) ocean_r <- get_ocean_rast()
  for (i in seq_len(iter)) {
    na_count_before <- global(is.na(r_filled), "sum")[[1]]
    r_focal  <- focal(r_filled, w = 3, fun = "mean", na.policy = "only", na.rm = TRUE)
    r_filled <- cover(r_filled, r_focal)
    na_count_after <- global(is.na(r_filled), "sum")[[1]]
    message(sprintf("Iter %d: %d NAs -> %d NAs", i, na_count_before, na_count_after))
    if (all(na_count_after == 0 | na_count_after == na_count_before)) break
    if(i %% 5 == 0) {
      ### check against ocean_rast - if no remaining ocean cells, exit
      check_r <- mask(ocean_r, r_filled, inverse = TRUE)
      n_ocean_remaining <- sum(!is.na(values(check_r)))
      message(sprintf("* Iter %d: %d ocean cells remaining", i, n_ocean_remaining))
      if(n_ocean_remaining == 0) break
    }
  }
  return(r_filled)
}
```

```{r gapfill vuln}
ocean_r <- get_ocean_rast() ### for masking and projecting

vuln_parq_fs <- list.files(here('_output/vuln_maps/biomass_removal'), 
                           pattern = 'demersal.parquet$|pelagic.parquet$', 
                           full.names = TRUE)
vuln_rast_fs <- sub('.parquet', '_gapfilled.tif', vuln_parq_fs)

for(i in seq_along(vuln_parq_fs)) {
  ### i <- 1
  pre_r <- parquet_to_rast(vuln_parq_fs[i], param = 'mean', scenario = 'rcp45_2050')

  post_f <- vuln_rast_fs[i]
  if(!file.exists(post_f)) {
    vuln_r_filled <- rast_gapfill(pre_r, iter = 100) |>
      mask(ocean_r)
    writeRaster(vuln_r_filled, post_f)
  }
  post_r <- rast(post_f)
  lbl <- gsub('_gapfilled.tif|^.+_vuln_', '', basename(post_f))
  plot(c(pre_r, post_r), main = lbl)
}

# if(any(!file.exists(vuln_rast_fs))) {
#   vuln_r_filled <- rast_gapfill(vuln_r, iter = 30)
# }
```