---
title: Map stressor intensity - fisheries
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

Try prototyping with EcoOcean model (and maybe total biomass from BOATS, instead of/alongside biomass density).  Process rasters into analysis grid (Behrmann, 5 km resolution).  Gapfill as needed.

* **Stressor reference points**: I think the ratio of fished biomass to unfished biomass is the starting place for a reference point
    * **Max stressor**: If we set max stressor (stressor = 1) to say 99.9th percentile of B_{fished}/B_{unfished}, that seems like a typical starting point
        * But, should there be consideration for total biomass instead of just the ratio?  A few low-biomass pixels with high fishing might (or might not) drive it if we ignore total biomass, or emphasize high fishing on high-biomass/high-productivity areas if we include it
        * My thought is to include total biomass if possible, but soften it with a log transform, so something like: (_biomass density (fished)_ / _biomass density (unfished)_) * log(_unfished biomass_) (and then do like 99th percentile)
    * ** Min stressor**: 
        * If we use a B/BMSY type of reference, say BMSY = 35-40% of unfished biomass, then that says that fishing at BMSY (or lower intensity) is a stressor of 0
        * If we use unfished biomass as a reference, then fishing at B/BMSY would still be a pretty substantial stressor
        * My inclination is use unfished biomass - taking out half the fish, even if sustainable over time, imposes an impact
* **Stressor intensity**: once reference points are established, we calculate the fished/unfished ratios for the various size/location groups as six separate layers


## Data

* fisheries models from ISIMIP: .nc files from https://data.isimip.org/search/tree/ISIMIP3b
* See `data_access_fisheries` script for download
* NetCDF files stored in `_data/isimip_2026/rasts/` with `_data/isimip_2026/filelist.txt` for filenames
* file names key: `ecoocean_gfdl-esm4_nobasd_ssp585_2015soc-from-histsoc_default_bd30cm_global_monthly_2015_2100.nc`
    * Model (`ecoocean`)
    * Earth system model (`gfdl-esm4`)
    * bias correction (`nobasd`)
    * climate scenario (SSP5-8.5)
    * fishery scenario (`2015soc-from-histso`: carry forward from 2015; `nat`: no fishing)
    * parameters (default)
    * variable (`b`iomass density, `d`emersal/`p`elagic, `30cm`/`30to90cm`/`90cm`)
    * scope (global)
    * period (monthly)
    * date range (2015-2100)

## Methods

```{r attach packages}
#| message: false
#| 
library(tidyverse)
library(here)
library(arrow)
library(collapse)
library(here)
library(terra)

source(here('common_fxns.R'))
```


### Process Fisheries Stressors

```{r}
fish_rast_df <- data.frame(file = list.files(here('_data/isimip_2026/rasts'), full.names = TRUE)) |> 
  mutate(base_f = basename(file)) |> 
  separate_wider_delim(
    base_f, 
    names = c('mdl', 'sys_mdl', NA, 'clim_scen', 'fish_scen', NA, 'var', NA, NA, 'yr_start', 'yr_end'), 
    delim = '_')
```

Try processing the demersal layers.  Monthly from 2015-01-01 to 2100-12-01.  Units are in g m^-2^.

```{r raster processing function}
grab_rast_layers <- function(c = 'ssp585', f, v, y = c(2046, 2055)) {
  rast_f <- fish_rast_df |> 
    filter(clim_scen == c & fish_scen == f & var == v) |> 
    pull(file)
  r <- rast(rast_f)

  r_time <- time(r)
  yrs    <- as.numeric(format(r_time, "%Y"))

  idx <- which(yrs >= min(y) & yrs <= max(y))
  r_sub <- r[[idx]]
  r_mean <- mean(r_sub)
  names(r_mean) <- v

  return(r_mean)
}

calc_stressor <- function(c = 'ssp585', v, y = c(2046, 2055)) {
  ### unfished and fished biomass density
  bd_unfished_r <- grab_rast_layers(c, f = 'nat', v, y)
  bd_fished_r <- grab_rast_layers(c, f = '2015soc-from-histsoc', v, y)
  ratio <- (bd_unfished_r - bd_fished_r) / (bd_unfished_r)

  ratio[ratio < 0] <- 0 ### some anomalous negative values in there

  ### Normalize to 1 at the 99.9th percentile observed  
  ref <- quantile(values(ratio), .999, na.rm = TRUE)
  str_r <- ratio / ref
  str_r[str_r > 1] <- 1

  names(str_r) <- paste0('str_intensity_', v)
  return(str_r)
}

rast_gapfill <- function(r, w = 3, iter = 5) {
  if(iter == 1) {
    r_focal <- focal(r, w = 3, fun = "mean", na.policy = "only", na.rm = TRUE)

    r_filled <- cover(r, r_focal)
    ### cover() returns r where r is non-NA, and r_focal elsewhere
    return(r_filled)
  }

  r_filled <- r

  for (i in seq_len(iter)) {
    na_count_before <- global(is.na(r_filled), "sum")[[1]]
    r_focal  <- focal(r_filled, w = 3, fun = "mean", na.policy = "only", na.rm = TRUE)
    r_filled <- cover(r_filled, r_focal)
    na_count_after <- global(is.na(r_filled), "sum")[[1]]
    message(sprintf("Iter %d: %d NAs -> %d NAs\n", i, na_count_before, na_count_after))
    if (all(na_count_after == 0 | na_count_after == na_count_before)) break
  }
  return(r_filled)
}
```

Fisheries stressor per size/position class $c$ is based on the proportion of biomass removed.

$$BD_{c}^{removed} = \frac{BD_c^{unfished} - BD_c^{fished}}{BD_c^{unfished}}$$

This ratio is normalized by the 99.9th percentile value as reference point.  Later we may weight the percentile using unfished biomass.  The `collapse::fquantile` function has a `w`eights argument.  

NOTE: We don't want to combine layers at the stressor intensity level - first apply the vulnerability scores, then combine the impact layers.

```{r calc stressor intensities}
str_intensity_f <- here('_output/stressor_maps/biomass_removal', 
                        'biomass_removal_stressor_intensity.tif')

if(!file.exists(str_intensity_f)) {
  str_dsm <- calc_stressor(v = 'bd30cm')
  str_dmd <- calc_stressor(v = 'bd30to90cm')
  str_dlg <- calc_stressor(v = 'bd90cm')

  str_psm <- calc_stressor(v = 'bp30cm')
  str_pmd <- calc_stressor(v = 'bp30to90cm')
  str_plg <- calc_stressor(v = 'bp90cm')

  ocean_r <- get_ocean_rast()

  ### consolidate stressor rasts and project  to Behrmann
  str_r <- c(str_dlg, str_plg, str_dmd, str_pmd, str_dsm, str_psm) |>
    ### need to gapfill coastal cell gaps
    rast_gapfill(iter = 5) |>
    project(ocean_r)

  ### Mask to just ocean cells
  str_r <- mask(str_r, ocean_r)

  ### any unfilled ocean cells?  inverse mask result against ocean rast
  # check_result <- mask(ocean_r, str_r, inverse = TRUE)
  # sum(!is.na(values(check_result)))
  # # [1] 0
  writeRaster(str_r, str_intensity_f)
}
str_r <- rast(str_intensity_f)
plot(str_r)
```

### Create a size class weighting layer

This raster will contain the biomass density for the various layers (sm/md/lg at demersal/pelagic).  This can be used to weight the impacts across layers for combining later.

```{r}
layer_biomass_f <- here('_output/stressor_maps/biomass_removal', 
                        'biomass_removal_biomass_by_class.tif')

if(!file.exists(layer_biomass_f)) {
  dsm <- grab_rast_layers(f = 'nat', v = 'bd30cm')
  dmd <- grab_rast_layers(f = 'nat', v = 'bd30to90cm')
  dlg <- grab_rast_layers(f = 'nat', v = 'bd90cm')
  psm <- grab_rast_layers(f = 'nat', v = 'bp30cm') 
  pmd <- grab_rast_layers(f = 'nat', v = 'bp30to90cm')
  plg <- grab_rast_layers(f = 'nat', v = 'bp90cm')

  biomass_r <- c(dlg, plg, dmd, pmd, dsm, psm) |>
    ### need to gapfill coastal cell gaps
    rast_gapfill(iter = 5) |>
    project(ocean_r) |>
    mask(ocean_r)

  writeRaster(biomass_r, layer_biomass_f)
}
biomass_r <- rast(layer_biomass_f)
plot(biomass_r)
```

