---
title: MS Figs for species and functional change
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

Generate MS figures for species and functional turnover:

* Species and functional richness, current and projected change
* Species and richness projected change in richness for large predators
    * Species and richness projected change in richness for fishery species (too similar?)
* Spp and FE change due to extirpation and colonization
* Overlap of SPP and FE turnover with fisheries biomass removal impacts


## Data

* AquaX species distribution models from Reygondeau et al. (2026). AquaX: An enhanced and revised AquaMaps framework to model marine species distributions and biodiversity. PLOS ONE, 21(2), e0335823. https://doi.org/10.1371/journal.pone.0335823
* Function entities from O'Hara et al. (2024). Cumulative human impacts on global marine fauna highlight risk to biological and functional diversity. PLOS ONE, 19(9), e0309788. https://doi.org/10.1371/journal.pone.0309788

## Methods

* Reproject impacts to same resolution as spp and FE turnover maps
* Rasterize the SPP and FE turnover maps
* Ideas for potential zonal comparison:
    * Fisheries quartiles?
    * Turnover quartiles?
* Other ideas for comparison
    * Regression?
    * PCA?
* NOTE: should the quartile breakdowns be implemented using an equal area projection, to avoid overemphasizing arctic and antarctic regions at the expense of equatorial regions?
    * YES! we will be using the behrmann projection, equal area cylindrical

```{r}
library(tidyverse)
library(here)
library(terra)
library(sf)
library(arrow)
library(collapse)
library(biscale)  ### for bivariate legend
library(cowplot)
```

### Read in and rasterize turnover metrics

THese rasters were created in scripts 3 (spp turnover) and 4 (fe turnover) in the Behrmann projection.

```{r read rasts}

spp_turn_r <- rast(here('_output/spp_turnover_maps/all_spp_turnover.tif'))
fe_turn_r  <- rast(here('_output/fe_turnover_maps/all_fe_turnover.tif'))

# plot(spp_turn_r); plot(fe_turn_r)
```

## Generate Manuscript Figures

Locate tropics lines and equator as well

```{r set up base and behrmann rasters}
base_rast <- rast(ext(-180, 180, -90, 90),
                  resolution = 0.05,
                  crs = 'epsg:4326')
behr_rast <- base_rast %>%
  project(y = '+proj=cea +lat_ts=30 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs', 
          res = 5000)
```

```{r set up land and parallels}
land_sf <- read_sf(here('_data/eezs/esri_world_continents.gpkg')) %>%
  st_transform(crs(behr_rast))

# trop_df <- data.frame(x = 0, y = c(23.425, -23.425, 0)) |> xy_to_cell_id(drop = FALSE)
### cancer: 9586800, capricorn: 16333200
# cell_id_behr_df <- base_rast |> 
#   setValues(1:ncell(base_rast)) |> 
#   project(behr_rast) |> 
#   round() |>
#   as.data.frame(xy = TRUE) |>
#   filter(lyr.1 %in% trop_df$cell_id)
### cancer: 2907724.47007114; capricorn: -2907275.52992886

plot_map <- function(data, fill_var) {
  data <- data |>
    filter(!is.na({{ fill_var }}))
  ggplot() +
    geom_raster(data = data, aes(x, y, fill = {{ fill_var }})) +
    ### continents:
    geom_sf(data = land_sf,
            fill = 'grey96', color = 'grey30', 
            size = .1) +
    ### tropics and equator:
    geom_hline(yintercept = c(2907724.470, -2907275.530), 
               linetype = 'dashed', linewidth = 0.25, color = 'grey30') +
    geom_hline(yintercept = 0, 
               linetype = 'solid', linewidth = 0.25, color = 'grey30') +
    ### standardized styling:
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_void()
}
```

### Fig. 1: Map of SPP and FE richness

Current and change 
```{r spp and fe richness}
fig1_f <- here('ms_figs/fig1_spp_fe_richness.png')

if(!file.exists(fig1_f)) {
  nspp_r <- rast(here('_output/spp_turnover_maps/all_spp_richness.tif'))
  nfe_r  <- rast(here('_output/fe_turnover_maps/all_fe_richness.tif'))

  ### Diverging palette from magenta to dark green, with yellow tone
  ### in middle - based on hcl.colors(pal = 'Red-Green')
  div_pal <- c("#841859", "#FFFFEE", "#005600")

  nspp_df <- as.data.frame(nspp_r, xy = TRUE) %>%
    fmutate(log_n = log10(spp_r_current),
            diff_rcp45_2050 = (rcp45_2050 - spp_r_current) / spp_r_current, 
            diff_rcp85_2050 = (rcp85_2050 - spp_r_current) / spp_r_current) %>%
    ### Treat anything over 1 as 1(+) (Inf: zero spp now 1), 
    ### Treat NaN as zero (zero spp still zero)
    fmutate(log_n = ifelse(is.infinite(log_n), NA, log_n)) |>
    fmutate(diff_rcp45_2050 = case_when((diff_rcp45_2050 > 1) ~ 1,
                                        is.nan(diff_rcp45_2050) ~ 0,
                                        TRUE ~ diff_rcp45_2050)) %>%
    fmutate(diff_rcp85_2050 = case_when((diff_rcp85_2050 > 1) ~ 1,
                                        is.nan(diff_rcp85_2050) ~ 0,
                                        TRUE ~ diff_rcp85_2050))

  nfe_df  <- as.data.frame(nfe_r, xy = TRUE) %>%
    fmutate(log_n = log10(fe_r_current),
            diff_rcp45_2050 = (rcp45_2050 - fe_r_current) / fe_r_current, 
            diff_rcp85_2050 = (rcp85_2050 - fe_r_current) / fe_r_current) %>% 
    fmutate(log_n = ifelse(is.infinite(log_n), NA, log_n)) |>
    fmutate(diff_rcp45_2050 = case_when((diff_rcp45_2050 > 1) ~ 1,
                                        is.nan(diff_rcp45_2050) ~ 0,
                                        TRUE ~ diff_rcp45_2050)) %>%
    fmutate(diff_rcp85_2050 = case_when((diff_rcp85_2050 > 1) ~ 1,
                                        is.nan(diff_rcp85_2050) ~ 0,
                                        TRUE ~ diff_rcp85_2050))
    
  nspp_max <- max(nspp_df$spp_r_current)
  nfe_max  <- max(nfe_df$fe_r_current)
  
  nspp_panel <- plot_map(data = nspp_df |> filter(!is.infinite(log_n)), 
                         fill_var = log_n) +
    scale_fill_viridis_c(breaks = c(0, 1, 2, 3, log10(nspp_max)), 
                         labels = c(1, 10, 100, 1000, nspp_max)) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(N[spp]))

  nfe_panel <- plot_map(data = nfe_df, 
                         fill_var = log_n) +
    scale_fill_viridis_c(breaks = c(0, 1, 2, log10(nfe_max)), 
                        labels = c(1, 10, 100, nfe_max)) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(N[FE]))

  dspp_panel <- plot_map(data = nspp_df, 
                         fill_var = diff_rcp45_2050) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression('∆' * N[spp]))

  dfe_panel <- plot_map(data = nfe_df, 
                        fill_var = diff_rcp45_2050) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression('∆' * N[FE]))

  p <- ggdraw() +
    draw_plot(nfe_panel, x = .02, y = 0, height = .5, width = .48) +
    draw_plot(dfe_panel, x = .51, y = 0, height = .5, width = .48) +
    draw_plot(nspp_panel, x = .02, y = .50, height = .5, width = .48) +
    draw_plot(dspp_panel, x = .51, y = .50, height = .5, width = .48) +
    draw_label('A', x = 0, y = 1, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('B', x = .51, y = 1, vjust = 1, hjust = 1, size = 14, fontface = 'bold') +
    draw_label('C', x = 0, y = .5, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('D', x = .51, y = .5, vjust = 1, hjust = 1, size = 14, fontface = 'bold')

  ggsave(fig1_f, height = 3.8, width = 10, dpi = 300)
}
knitr::include_graphics(fig1_f)
```

### Fig. 2: Focus on apex spp: richness and pct change

Generate panels A and B
```{r apex spp richness A and B}
fig2_f <- here('ms_figs/fig2_apex_spp_richness.png')

if(!file.exists(fig2_f)) {
  apex_nspp_r <- rast(here('_output/spp_turnover_maps/apex_spp_richness.tif'))
  all_nspp_r  <- rast(here('_output/spp_turnover_maps/all_spp_richness.tif'))

  ### Diverging palette from magenta to dark green, with yellow tone
  ### in middle - based on hcl.colors(pal = 'Red-Green')
  div_pal <- c("#841859", "#FFFFEE", "#005600")

  nspp_df <- as.data.frame(apex_nspp_r, xy = TRUE) %>%
    fmutate(log_n = log10(spp_r_current),
            diff_rcp45_2050 = (rcp45_2050 - spp_r_current) / spp_r_current) %>%
    ### Treat anything over 1 as 1(+) (Inf: zero spp now 1), 
    ### Treat NaN as zero (zero spp still zero)
    fmutate(log_n = ifelse(is.infinite(log_n), NA, log_n)) |>
    fmutate(diff_rcp45_2050 = case_when((diff_rcp45_2050 > 1) ~ 1,
                                        is.nan(diff_rcp45_2050) ~ 0,
                                        TRUE ~ diff_rcp45_2050))
    
  nspp_max <- max(nspp_df$spp_r_current)
  nspp_brks <- seq(1, log10(nspp_max), 1)
  nspp_plot <- plot_map(data = nspp_df, fill_var = log_n) +
    scale_fill_viridis_c(breaks = c(0, 1, 2, 3, log10(nspp_max)), 
                        labels = c(1, 10, 100, 1000, nspp_max)) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(N[pred]))

  dspp_plot <- plot_map(data = nspp_df, fill_var = diff_rcp45_2050) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression('∆' * N[pred]))
}
```

Generate panel C
```{r apex spp richness C}

if(!file.exists(fig2_f)) {
  ### Diverging palette orange-purple to indicate difference from pink-green
  div_pal <- c("#f1a340", "#f1f1f1", "#998ec3")

  apex_df <- as.data.frame(apex_nspp_r, xy = TRUE) %>%
    select(x, y, apex_current = spp_r_current, apex_2050 = rcp45_2050)
   
  all_df <- as.data.frame(all_nspp_r, xy = TRUE) %>%
    select(x, y, all_current = spp_r_current, all_2050 = rcp45_2050)

  compare_df <- apex_df %>%
    join(all_df, how = 'full', on = c('x', 'y')) %>%
    fmutate(other_current = all_current - apex_current,
            other_2050    = all_2050 - apex_2050,
            apex_diff     = (apex_2050 - apex_current) / apex_current,
            other_diff    = (other_2050 - other_current) / other_current) %>%
    fmutate(delta = apex_diff - other_diff,
            delta = ifelse(delta > 1, 1, delta),
            delta = ifelse(delta < -1, -1, delta))

  diff_plot <- plot_map(data = compare_df, fill_var = delta) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(atop('∆' * N[pred], -'∆' * N[other])))
}
```

Assemble three panel figure

```{r fig 2 pull panels}
if(!file.exists(fig2_f)) {
  nspp_panel <- get_panel(nspp_plot)
  dspp_panel <- get_panel(dspp_plot)
  diff_panel <- get_panel(diff_plot)
  nspp_lgd <- get_legend(nspp_plot)
  dspp_lgd <- get_legend(dspp_plot)
  diff_lgd <- get_legend(diff_plot)
}
```

```{r apex spp richness assemble figure}
if(!file.exists(fig2_f)) {
  x1 <- 0.02; w1 <- 0.88; h1 <- 0.33
  x2 <- 0.90; w2 <- 0.10; h2 <- 0.33
  ya <- 0.667; yb <- 0.333; yc <- 0.000

  p <- ggdraw() +
    ### plot panels
    draw_plot(nspp_panel, x = x1, y = ya, height = h1, width = w1) +
    draw_plot(dspp_panel, x = x1, y = yb, height = h1, width = w1) +
    draw_plot(diff_panel, x = x1, y = yc, height = h1, width = w1) +
    ### plot legends
    draw_plot(nspp_lgd, x = x2-.012, y = ya, height = h2, width = w2) +
    draw_plot(dspp_lgd, x = x2, y = yb, height = h2, width = w2) +
    draw_plot(diff_lgd, x = x2, y = yc, height = h2, width = w2) +
    ### draw labels
    draw_label('A', x = 0, y = ya+h1, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('B', x = 0, y = yb+h1, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('C', x = 0, y = yc+h1, vjust = 1, hjust = 0, size = 14, fontface = 'bold')
  ggsave(fig2_f, height = 9, width = 8, dpi = 300)
}
knitr::include_graphics(fig2_f)
```

### Focus on fisheries spp: richness and pct change
Generate panels A and B
```{r fish spp richness A and B}
figX_f <- here('ms_figs/figX_fisheries_spp_richness.png')

if(!file.exists(figX_f)) {
  fish_nspp_r <- rast(here('_output/spp_turnover_maps/fisheries_spp_richness.tif'))
  all_nspp_r  <- rast(here('_output/spp_turnover_maps/all_spp_richness.tif'))

  ### Diverging palette from magenta to dark green, with yellow tone
  ### in middle - based on hcl.colors(pal = 'Red-Green')
  div_pal <- c("#841859", "#FFFFEE", "#005600")

  nspp_df <- as.data.frame(fish_nspp_r, xy = TRUE) %>%
    fmutate(log_n = log10(spp_r_current),
            diff_rcp45_2050 = (rcp45_2050 - spp_r_current) / spp_r_current) %>%
    ### Treat anything over 1 as 1(+) (Inf: zero spp now 1), 
    ### Treat NaN as zero (zero spp still zero)
    fmutate(log_n = ifelse(is.infinite(log_n), NA, log_n)) |>
    fmutate(diff_rcp45_2050 = case_when((diff_rcp45_2050 > 1) ~ 1,
                                        is.nan(diff_rcp45_2050) ~ 0,
                                        TRUE ~ diff_rcp45_2050))
    
  nspp_max <- max(nspp_df$spp_r_current)
  nspp_brks <- seq(1, log10(nspp_max), 1)
  nspp_plot <- plot_map(data = nspp_df, fill_var = log_n) +
    scale_fill_viridis_c(breaks = c(0, 1, 2, 3, log10(nspp_max)), 
                        labels = c(1, 10, 100, 1000, nspp_max)) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(N[fishery]))

  dspp_plot <- plot_map(data = nspp_df, fill_var = diff_rcp45_2050) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression('∆' * N[fishery]))
}
```

Generate panel C
```{r fishery spp richness C}

if(!file.exists(figX_f)) {
  ### Diverging palette orange-purple to indicate difference from pink-green
  div_pal <- c("#f1a340", "#f1f1f1", "#998ec3")

  fish_df <- as.data.frame(fish_nspp_r, xy = TRUE) %>%
    select(x, y, fish_current = spp_r_current, fish_2050 = rcp45_2050)
   
  all_df <- as.data.frame(all_nspp_r, xy = TRUE) %>%
    select(x, y, all_current = spp_r_current, all_2050 = rcp45_2050)

  compare_df <- fish_df %>%
    join(all_df, how = 'full', on = c('x', 'y')) %>%
    fmutate(other_current = all_current - fish_current,
            other_2050    = all_2050 - fish_2050,
            fish_diff     = (fish_2050 - fish_current) / fish_current,
            other_diff    = (other_2050 - other_current) / other_current) %>%
    fmutate(delta = fish_diff - other_diff,
            delta = ifelse(delta > 1, 1, delta),
            delta = ifelse(delta < -1, -1, delta))

  diff_plot <- plot_map(data = compare_df, fill_var = delta) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(atop('∆' * N[fishery], -'∆' * N[other])))
}
```

Assemble three panel figure

```{r fig fisheries pull panels}
if(!file.exists(figX_f)) {
  nspp_panel <- get_panel(nspp_plot)
  dspp_panel <- get_panel(dspp_plot)
  diff_panel <- get_panel(diff_plot)
  nspp_lgd <- get_legend(nspp_plot)
  dspp_lgd <- get_legend(dspp_plot)
  diff_lgd <- get_legend(diff_plot)
}
```

```{r fisheries spp richness assemble figure}
if(!file.exists(figX_f)) {
  x1 <- 0.02; w1 <- 0.88; h1 <- 0.33
  x2 <- 0.90; w2 <- 0.10; h2 <- 0.33
  ya <- 0.667; yb <- 0.333; yc <- 0.000

  p <- ggdraw() +
    ### plot panels
    draw_plot(nspp_panel, x = x1, y = ya, height = h1, width = w1) +
    draw_plot(dspp_panel, x = x1, y = yb, height = h1, width = w1) +
    draw_plot(diff_panel, x = x1, y = yc, height = h1, width = w1) +
    ### plot legends
    draw_plot(nspp_lgd, x = x2-.012, y = ya, height = h2, width = w2) +
    draw_plot(dspp_lgd, x = x2, y = yb, height = h2, width = w2) +
    draw_plot(diff_lgd, x = x2, y = yc, height = h2, width = w2) +
    ### draw labels
    draw_label('A', x = 0, y = ya+h1, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('B', x = 0, y = yb+h1, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('C', x = 0, y = yc+h1, vjust = 1, hjust = 0, size = 14, fontface = 'bold')
  ggsave(figX_f, height = 9, width = 8, dpi = 300)
}
knitr::include_graphics(figX_f)
```

### Focus on aquaculture spp: richness and pct change
Generate panels A and B
```{r aqua spp richness A and B}
figX_f <- here('ms_figs/figX_aquaculture_spp_richness.png')

if(!file.exists(figX_f)) {
  aqua_nspp_r <- rast(here('_output/spp_turnover_maps/aquaculture_spp_richness.tif'))
  all_nspp_r  <- rast(here('_output/spp_turnover_maps/all_spp_richness.tif'))

  ### Diverging palette from magenta to dark green, with yellow tone
  ### in middle - based on hcl.colors(pal = 'Red-Green')
  div_pal <- c("#841859", "#FFFFEE", "#005600")

  nspp_df <- as.data.frame(aqua_nspp_r, xy = TRUE) %>%
    fmutate(log_n = log10(spp_r_current),
            diff_rcp45_2050 = (rcp45_2050 - spp_r_current) / spp_r_current) %>%
    ### Treat anything over 1 as 1(+) (Inf: zero spp now 1), 
    ### Treat NaN as zero (zero spp still zero)
    fmutate(log_n = ifelse(is.infinite(log_n), NA, log_n)) |>
    fmutate(diff_rcp45_2050 = case_when((diff_rcp45_2050 > 1) ~ 1,
                                        is.nan(diff_rcp45_2050) ~ 0,
                                        TRUE ~ diff_rcp45_2050))
    
  nspp_max <- max(nspp_df$spp_r_current)
  nspp_brks <- seq(1, log10(nspp_max), 1)
  nspp_plot <- plot_map(data = nspp_df, fill_var = log_n) +
    scale_fill_viridis_c(breaks = c(0, 1, 2, 3, log10(nspp_max)), 
                        labels = c(1, 10, 100, 1000, nspp_max)) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(N[aqua]))

  dspp_plot <- plot_map(data = nspp_df, fill_var = diff_rcp45_2050) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression('∆' * N[aqua]))
}
```

Generate panel C
```{r aquaculture spp richness C}

if(!file.exists(figX_f)) {
  ### Diverging palette orange-purple to indicate difference from pink-green
  div_pal <- c("#f1a340", "#f1f1f1", "#998ec3")

  aqua_df <- as.data.frame(aqua_nspp_r, xy = TRUE) %>%
    select(x, y, aqua_current = spp_r_current, aqua_2050 = rcp45_2050)
   
  all_df <- as.data.frame(all_nspp_r, xy = TRUE) %>%
    select(x, y, all_current = spp_r_current, all_2050 = rcp45_2050)

  compare_df <- aqua_df %>%
    join(all_df, how = 'full', on = c('x', 'y')) %>%
    fmutate(other_current = all_current - aqua_current,
            other_2050    = all_2050 - aqua_2050,
            aqua_diff     = (aqua_2050 - aqua_current) / aqua_current,
            other_diff    = (other_2050 - other_current) / other_current) %>%
    fmutate(delta = aqua_diff - other_diff,
            delta = ifelse(delta > 1, 1, delta),
            delta = ifelse(delta < -1, -1, delta))

  diff_plot <- plot_map(data = compare_df, fill_var = delta) +
    scale_fill_gradient2(low = div_pal[1], mid = div_pal[2], 
                        high = div_pal[3], midpoint = 0,
                        breaks = c(-1, -.5, 0, .5, 1),
                        labels = c('-100%', '-50%', '0%', '50%', '≥100%')) +
    theme(text = element_text(size = 9)) +
    labs(fill = expression(atop('∆' * N[aqua], -'∆' * N[other])))
}
```

Assemble three panel figure

```{r fig aquaculture pull panels}
if(!file.exists(figX_f)) {
  nspp_panel <- get_panel(nspp_plot)
  dspp_panel <- get_panel(dspp_plot)
  diff_panel <- get_panel(diff_plot)
  nspp_lgd <- get_legend(nspp_plot)
  dspp_lgd <- get_legend(dspp_plot)
  diff_lgd <- get_legend(diff_plot)
}
```

```{r aquaculture spp richness assemble figure}
if(!file.exists(figX_f)) {
  x1 <- 0.02; w1 <- 0.88; h1 <- 0.33
  x2 <- 0.90; w2 <- 0.10; h2 <- 0.33
  ya <- 0.667; yb <- 0.333; yc <- 0.000

  p <- ggdraw() +
    ### plot panels
    draw_plot(nspp_panel, x = x1, y = ya, height = h1, width = w1) +
    draw_plot(dspp_panel, x = x1, y = yb, height = h1, width = w1) +
    draw_plot(diff_panel, x = x1, y = yc, height = h1, width = w1) +
    ### plot legends
    draw_plot(nspp_lgd, x = x2-.012, y = ya, height = h2, width = w2) +
    draw_plot(dspp_lgd, x = x2, y = yb, height = h2, width = w2) +
    draw_plot(diff_lgd, x = x2, y = yc, height = h2, width = w2) +
    ### draw labels
    draw_label('A', x = 0, y = ya+h1, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('B', x = 0, y = yb+h1, vjust = 1, hjust = 0, size = 14, fontface = 'bold') +
    draw_label('C', x = 0, y = yc+h1, vjust = 1, hjust = 0, size = 14, fontface = 'bold')
  ggsave(figX_f, height = 9, width = 8, dpi = 300)
}
knitr::include_graphics(figX_f)
```


### Map of turnover

Compare species and FE turnover in a single plot, just examining RCP 4.5 (8.5 can go in SI?)
```{r spp and fe turnover}
fig_file <- here('ms_figs/figX_spp_fe_turnover.png')

if(!exists(fig_file)) {
  turnover_rasts <- c(spp_turn_r[['rcp45_2050']], fe_turn_r[['rcp45_2050']])
  turnover_df <- as.data.frame(turnover_rasts, xy = TRUE) %>%
    setNames(c('x', 'y', 'Species Turnover', 'Functional Entity Turnover')) %>%
    pivot_longer(cols = -c(x, y), names_to = 'var', values_to = 'val') %>%
    mutate(var = factor(var, levels = c('Species Turnover', 'Functional Entity Turnover'))) 

  spp_fe_turnover_plot <- plot_map(data = turnover_df, fill_var = val) +
    scale_fill_viridis_c() +
    theme(axis.text = element_blank(), 
          axis.title = element_blank(),
          strip.text.y.left = element_text(angle = 90)) +
    facet_wrap(~ var, ncol = 1, strip.position = 'left') +
    labs(fill = expression(beta[.Sør]))

  fig_file <- here('ms_figs/figX_spp_fe_turnover.png')
  ggsave(fig_file, height = 6, width = 8, dpi = 300)
}
knitr::include_graphics(fig_file)
```

### Map of turnover by component

Break turnover into extirpation and colonization components, plot as 4-panel.

```{r spp and fe turnover components}
fig_file <- here('ms_figs/figX_spp_fe_turnover_components.png')

if(!file.exists(fig_file)) {
  turnover_rasts <- c(spp_turn_r[['rcp45_2050']], fe_turn_r[['rcp45_2050']])

  spp_turn_ext_r <- rast(here('_output/spp_turnover_maps/all_spp_turnover_extirp.tif'))
  fe_turn_ext_r  <- rast(here('_output/fe_turnover_maps/all_fe_turnover_extirp.tif'))

  extirp_rasts <- c(spp_turn_ext_r[['rcp45_2050']], fe_turn_ext_r[['rcp45_2050']]) %>%
    setNames(c('SpeciesXExtirpation', 'FEXExtirpation'))
  coloniz_rasts <- (turnover_rasts - extirp_rasts) %>%
    setNames(c('SpeciesXColonization', 'FEXColonization'))

  turn_component_rasts <- c(extirp_rasts, coloniz_rasts)

  turn_comp_df <- as.data.frame(turn_component_rasts, xy = TRUE) %>%
    pivot_longer(cols = -c(x, y), names_to = 'var', values_to = 'val') %>%
    fmutate(unit = sub('X.+', '', var),
            comp = sub('.+X', '', var)) %>%
    fmutate(unit = factor(unit, levels = c('Species', 'FE')),
            comp = factor(comp, levels = c('Extirpation', 'Colonization')))

  spp_fe_turn_comp_plot <- plot_map(data = turn_comp_df, fill_var = val) +
    scale_fill_viridis_c() +
    theme(strip.text.y.left = element_text(angle = 90),
          axis.text = element_blank(), 
          axis.title = element_blank()) +
    facet_grid(unit ~ comp, switch = 'y') +
    labs(fill = expression(beta[.Sør]))

  ggsave(fig_file, height = 4, width = 10, dpi = 300)
}
knitr::include_graphics(fig_file)
```

