---
title: Map species and functional change against fisheries impacts
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

Compare turnover metrics against fisheries impacts.

## Data

* AquaX species distribution models from Reygondeau et al. (2026). AquaX: An enhanced and revised AquaMaps framework to model marine species distributions and biodiversity. PLOS ONE, 21(2), e0335823. https://doi.org/10.1371/journal.pone.0335823
* Function entities from O'Hara et al. (2024). Cumulative human impacts on global marine fauna highlight risk to biological and functional diversity. PLOS ONE, 19(9), e0309788. https://doi.org/10.1371/journal.pone.0309788
* Fisheries projections from ISIMIP 3b BOATSv1

## Methods

* Reproject impacts to same resolution as spp and FE turnover maps
* Rasterize the SPP and FE turnover maps
* Ideas for potential zonal comparison:
    * Fisheries quartiles?
    * Turnover quartiles?
* Other ideas for comparison
    * Regression?
    * PCA?
* NOTE: should the quartile breakdowns be implemented using an equal area projection, to avoid overemphasizing arctic and antarctic regions at the expense of equatorial regions?
    * YES! we will be using the behrmann projection, equal area cylindrical

```{r}
library(tidyverse)
library(here)
library(terra)
library(sf)
library(arrow)
library(collapse)
library(biscale)  ### for bivariate legend
library(cowplot)
```

### Load fisheries impacts

Biomass removal impacts generated in scripts `map_vuln_fisheries.qmd`, `map_stressors_fisheries.qmd`, and `map_impact_fisheries.qmd`.  These are already in raster format and Behrmann projection at 5 km resolution.

Layers include `impact_demersal`, `impact_pelagic`, and `impact_total` (biomass-weighted mean of demersal and pelagic).

```{r fisheries impacts}
fish_impact_r <- rast(here('_output/impact_maps/biomass_removal_impact.tif'))

```


### Read in and rasterize turnover metrics

THese rasters were created in scripts 3 (spp turnover) and 4 (fe turnover) in the Behrmann projection.

```{r read rasts}

spp_turn_r <- rast(here('_output/spp_turnover_maps/all_spp_turnover.tif'))
fe_turn_r  <- rast(here('_output/fe_turnover_maps/all_fe_turnover.tif'))

# plot(spp_turn_r); plot(fe_turn_r)
```

## Generate Manuscript Figures

Locate tropics lines and equator as well

```{r set up base and behrmann rasters}
base_rast <- rast(ext(-180, 180, -90, 90),
                  resolution = 0.05,
                  crs = 'epsg:4326')
behr_rast <- base_rast %>%
  project(y = '+proj=cea +lat_ts=30 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs', 
          res = 5000)
```

```{r set up land and parallels}
land_sf <- read_sf(here('_data/eezs/esri_world_continents.gpkg')) %>%
  st_transform(crs(behr_rast))

# trop_df <- data.frame(x = 0, y = c(23.425, -23.425, 0)) |> xy_to_cell_id(drop = FALSE)
### cancer: 9586800, capricorn: 16333200
# cell_id_behr_df <- base_rast |> 
#   setValues(1:ncell(base_rast)) |> 
#   project(behr_rast) |> 
#   round() |>
#   as.data.frame(xy = TRUE) |>
#   filter(lyr.1 %in% trop_df$cell_id)
### cancer: 2907724.47007114; capricorn: -2907275.52992886

plot_map <- function(data, fill_var) {
  data <- data |>
    filter(!is.na({{ fill_var }}))
  ggplot() +
    geom_raster(data = data, aes(x, y, fill = {{ fill_var }})) +
    ### continents:
    geom_sf(data = land_sf,
            fill = 'grey96', color = 'grey30', 
            size = .1) +
    ### tropics and equator:
    geom_hline(yintercept = c(2907724.470, -2907275.530), 
               linetype = 'dashed', linewidth = 0.25, color = 'grey30') +
    geom_hline(yintercept = 0, 
               linetype = 'solid', linewidth = 0.25, color = 'grey30') +
    ### standardized styling:
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_void()
}
```


### Map fisheries impacts vs turnover

Here let's do a bivariate plot to compare areas of high/low predicted fisheries impact to high/low projected turnover.  

```{r process to quartiles}
fe_turn_impact_f  <- here('_output/figure_data/fig_fe_turnover_fisheries.parquet')
spp_turn_impact_f <- here('_output/figure_data/fig_spp_turnover_fisheries.parquet')

if(any(!file.exists(c(fe_turn_impact_f, spp_turn_impact_f)))) {
  impact_qtile <- fe_turn_qtile <- spp_turn_qtile <- spp_turn_r ### initialize three new maps

  impact_qtile <- impact_qtile[[1]]
  ### Set breaks for impacts
  fish_r <- fish_impact_r[[3]] ### total fisheries impact

  fish_thresholds <- quantile(values(fish_r), 
                                probs = seq(0, 1, 0.25), 
                                na.rm = TRUE, names = FALSE)
  # [1] 0.000 0.081 0.096 0.144 1.550
  ### NOTE: between RCP4.5 and 8.5, the 25% and 50% remain the same, but 75% and 100% increase
  values(impact_qtile) <- data.table::fcase(values(fish_r) < fish_thresholds[2], 1L,
                                            values(fish_r) < fish_thresholds[3], 2L,
                                            values(fish_r) < fish_thresholds[4], 3L,
                                            !is.na(values(fish_r)), 4L)

  ### Set breaks for turnover
  ### Note: since turnover is a physical thing, just set as .25, .5, .75 and call it a day.
  ### Set zero turnover to be category 1
  values(fe_turn_qtile)  <- ceiling(values(fe_turn_r * 4))
  fe_turn_qtile[fe_turn_qtile == 0] <- 1
  values(spp_turn_qtile) <- ceiling(values(spp_turn_r * 4))
  spp_turn_qtile[spp_turn_qtile == 0] <- 1

  fe_turn_df <- as.data.frame(fe_turn_qtile, xy = TRUE) %>%
    setNames(c('x', 'y', 'fe_turn_rcp45', 'fe_turn_rcp85'))
  spp_turn_df <- as.data.frame(spp_turn_qtile, xy = TRUE) %>%
    setNames(c('x', 'y', 'spp_turn_rcp45', 'spp_turn_rcp85'))
  impact_df <- as.data.frame(impact_qtile, xy = TRUE) %>%
    setNames(c('x', 'y', 'impact_rcp85'))

  fe_turn_impact_df <- collapse::join(fe_turn_df, impact_df, 
    how = 'left', on = c('x', 'y')) %>%
    mutate(bi_class_rcp45 = paste(fe_turn_rcp45, impact_rcp85, sep = '-'),
           bi_class_rcp85 = paste(fe_turn_rcp85, impact_rcp85, sep = '-'))
  spp_turn_impact_df <- collapse::join(spp_turn_df, impact_df, 
    how = 'left', on = c('x', 'y')) %>%
    mutate(bi_class_rcp45 = paste(spp_turn_rcp45, impact_rcp85, sep = '-'),
           bi_class_rcp85 = paste(spp_turn_rcp85, impact_rcp85, sep = '-'))

  write_parquet(fe_turn_impact_df,  fe_turn_impact_f)
  write_parquet(spp_turn_impact_df, spp_turn_impact_f)
}

fe_turn_impact_df  <- read_parquet(fe_turn_impact_f)
spp_turn_impact_df <- read_parquet(spp_turn_impact_f)

```

```{r set up biplot}
calc_biplot_pcts <- function(df, scenario = 'rcp45') {
  df_sum <- df %>% 
    select(bi_class = all_of(paste0('bi_class_', scenario))) %>%
    fsubset(!grepl('NA', bi_class)) %>%
    fcount(bi_class) %>%
    mutate(pct = N / sum(N)) %>%
    separate_wider_delim(bi_class, names = c('turn', 'impact'), delim = '-') %>%
    mutate(turn = as.integer(turn), impact = as.integer(impact),
           lbl = paste0(round(pct, 3) * 100, '%'),
           dark = impact >= 3 & turn >= 3)
}
fe_turn_pcts_rcp45 <- fe_turn_impact_df %>%
  calc_biplot_pcts(scenario = 'rcp45')
spp_turn_pcts_rcp45 <- spp_turn_impact_df %>%
  calc_biplot_pcts(scenario = 'rcp45')
```
```{r build palette}

map_pal_raw <- bi_pal(pal = 'GrPink2', dim = 4, preview = FALSE)
  
map_pal_mtx      <- matrix(map_pal_raw, nrow = 4, ncol = 4)
map_pal_mtx[3, ] <- colorspace::lighten(map_pal_mtx[3, ], .1)
map_pal_mtx[2, ] <- colorspace::lighten(map_pal_mtx[2, ], .2)
map_pal_mtx[1, ] <- colorspace::lighten(map_pal_mtx[1, ], .3)
map_pal_mtx[ , 3] <- colorspace::lighten(map_pal_mtx[ , 3], .1)
map_pal_mtx[ , 2] <- colorspace::lighten(map_pal_mtx[ , 2], .2)
map_pal_mtx[ , 1] <- colorspace::lighten(map_pal_mtx[ , 1], .3)
map_pal_mtx[1, 1] <- '#ffffee'

map_pal <- as.vector(map_pal_mtx) %>% setNames(names(map_pal_raw))

pnl_lgd_fe <- bi_legend(pal = map_pal, dim = 4,
                        x = 'Functional turnover',
                        y = 'Biomass removal') +
  geom_text(data = fe_turn_pcts_rcp45, 
            aes(x = turn, y = impact, label = lbl, color = dark),
            size = 3, show.legend = FALSE) +
  scale_color_manual(values = c('grey20', 'white'))
pnl_lgd_spp <- bi_legend(pal = map_pal, dim = 4,
                        x = 'Species turnover',
                        y = 'Biomass removal') +
  geom_text(data = spp_turn_pcts_rcp45, 
            aes(x = turn, y = impact, label = lbl, color = dark),
            size = 3, show.legend = FALSE) +
  scale_color_manual(values = c('grey20', 'white'))
```

#### FE turnover vs impacts, RCP4.5
```{r fe_45}
panel_fe_45 <- plot_map(data = fe_turn_impact_df, fill_var = bi_class_rcp45) +
  bi_scale_fill(pal = map_pal, dim = 4) +
  theme(axis.text = element_blank(), 
        axis.title = element_blank(),
        legend.position = 'none')

# panel_fe_45
```


#### Spp turnover vs impacts, RCP4.5
```{r spp_45}
panel_spp_45 <- plot_map(data = spp_turn_impact_df, fill_var = bi_class_rcp45) +
  bi_scale_fill(pal = map_pal, dim = 4) +
  theme(axis.text = element_blank(), 
        axis.title = element_blank(),
        legend.position = 'none')

# panel_spp_45
```


#### Assemble turnover X impact figure

Figure shows RCP4.5 spp and FE turnover vs. impacts.  Impacts are same, but FE turnover will be less than SPP turnover.

```{r pull panels}
# panel_a <- get_panel(panel_spp_45)
# panel_b <- get_panel(panel_fe_45)
panel_a <- panel_spp_45
panel_b <- panel_fe_45
bi_lgd_fe  <- pnl_lgd_fe
bi_lgd_spp <- pnl_lgd_spp
```

```{r assemble panels}
fig <- ggdraw() +
  draw_plot(pnl_lgd_spp,  x = .68, y = .55, width = .40, height = .45) +
  draw_plot(panel_spp_45, x = .012, y = 0.5, width = .74, height = .5) +
  draw_plot(pnl_lgd_fe,   x = .68, y = .05, width = .40, height = .45) +
  draw_plot(panel_fe_45,  x = .012, y = 0.0, width = .74, height = .5) +
  draw_label('A', x = 0,  y = 1.0, hjust = 0, vjust = 1, size = 16, fontface = 'bold') +
  draw_label('B', x = 0,  y = 0.5, hjust = 0, vjust = 1, size = 16, fontface = 'bold')
 
fig_file <- here('ms_figs/fig4_spp_fe_impacts_fisheries.png')
ggsave(fig_file, height = 5.75, width = 9.5, dpi = 300)
knitr::include_graphics(fig_file)
```