---
title: MS Figs coastal vs oceanic
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
execute:
    warning: false
---

## Summary

Generate MS figures for species and functional turnover:

* Species and functional richness, current and projected change
* Species and richness projected change in richness for large predators
    * Species and richness projected change in richness for fishery species (too similar?)
* Spp and FE change due to extirpation and colonization
* Overlap of SPP and FE turnover with fisheries biomass removal impacts


## Data

* GEBCO gridded elevation - GEBCO_2026 Grid (sub-ice topo/bathy): https://www.gebco.net/data-products/gridded-bathymetry-data#global
    * download link: https://dap.ceda.ac.uk/bodc/gebco/global/gebco_2026/sub_ice_topography_bathymetry/netcdf/GEBCO_2026_sub_ice.zip?download=1
    * Stored in `_data/gebco_2026/GEBCO_2026_sub_ice_0.05.nc`
        * This is an aggregated NetCDF from 15 arc-second to 0.05 degree; the original 7 GB NetCDF is aggregated then deleted.

## Methods

Load turnover maps, partition out by neritic/coastal (<= 200m) and oceanic (>200m) using GEBCO bathymetry.

```{r}
library(tidyverse)
library(here)
library(terra)
library(sf)
library(arrow)
library(collapse)
library(cowplot)
```

### Read in and rasterize turnover metrics

THese rasters were created in scripts 3 (spp turnover) and 4 (fe turnover) in the Behrmann projection.

```{r read rasts}

spp_turn_r <- rast(here('_output/spp_turnover_maps/all_spp_turnover.tif'))
fe_turn_r  <- rast(here('_output/fe_turnover_maps/all_fe_turnover.tif'))

# plot(spp_turn_r); plot(fe_turn_r)
```

### Read in and transform GEBCO bathymetry

```{r}
bathy_r_4326 <- rast(here('_data/gebco_2026/GEBCO_2026_sub_ice_0.05.nc'))
bathy_r_4326[bathy_r_4326 >= -200] <- 1 ### 1 is coastal
bathy_r_4326[bathy_r_4326 < -200]  <- 2 ### 2 is oceanic

bathy_r <- project(bathy_r_4326, spp_turn_r) |> 
  round() |> 
  mask(spp_turn_r[[1]])
```

### Compare oceanic and coastal turnover


```{r}
combined_r <- c(spp_turn_r[[1]], fe_turn_r[[1]], bathy_r)
combined_df <- as.data.frame(combined_r, xy = TRUE) |> 
  setNames(c('x', 'y', 'spp', 'fe', 'depth'))

summary_df <- combined_df |> 
  slice_sample(n = 1e6, by = depth) |> 
  pivot_longer(
    cols = c(spp, fe),
    names_to = "variable",
    values_to = "turnover"
  ) |> 
  group_by(variable) |> 
  summarise(
    test = list(t.test(turnover ~ depth)), 
    .groups = "drop"
  ) |> 
  mutate(tidy_result = map(test, broom::tidy)) %>%
  unnest(tidy_result)

summary_df |> select(-test, -parameter, -method)
```

Do we need to go any further, e.g., a visual?  seems pretty clear cut